---
title: "Tuberculosis Surveillance"
subtitle: "Epidemiological Description of TB Notifications · NMC Surveillance, South Africa"
date: today
sidebar: special-reports
execute:
  echo: false
  warning: false
  message: false
  error: false
  cache: false
---

::: {.callout-important appearance="minimal"}
**Priority condition** — Tuberculosis remains a leading cause of morbidity and mortality in South Africa. This report describes the epidemiology of notifiable TB forms based on NMC surveillance data, 2020–present, including drug-resistant TB (MDR-TB and XDR-TB).
:::

```{r setup}
#| label: setup
#| include: false
#| cache: false

pacman::p_load(
  ggplot2, gtsummary, tidyverse, magrittr,
  conflicted, scales, data.table, lubridate,
  flextable, gt, ggrepel, zoo,
  janitor
)

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::lag)
conflicted::conflicts_prefer(tidyr::extract)
conflicted::conflicts_prefer(dplyr::first)
conflicted::conflicts_prefer(lubridate::year)
conflicted::conflicts_prefer(lubridate::month)
conflicted::conflicts_prefer(lubridate::quarter)

library(NMCleaner)

# ── Load master dataset ───────────────────────────────────────────────────────
load("~/Desktop/SAFETP/CLA/NMC_database/master/new_master.rda")

# ── Define TB conditions ──────────────────────────────────────────────────────
TB_CONDITIONS <- c(
  "Tuberculosis:pulmonary",
  "Tuberculosis:extra-pulmonary",
  "Tuberculosis: multidrug- resistant (MDR -TB)",
  "Tuberculosis: extensively drug -resistant (XDR -TB)"
)

# Shorter labels for display
TB_LABELS <- c(
  "Tuberculosis:pulmonary"                              = "Pulmonary TB",
  "Tuberculosis:extra-pulmonary"                        = "Extra-pulmonary TB",
  "Tuberculosis: multidrug- resistant (MDR -TB)"        = "MDR-TB",
  "Tuberculosis: extensively drug -resistant (XDR -TB)" = "XDR-TB"
)

# ── Filter to TB conditions only ──────────────────────────────────────────────
new_master %>%
  filter(
    condition %in% TB_CONDITIONS,
    as.integer(Year_notification) %in% 2020:year(Sys.Date()),
    !is.na(prov_)
  ) %>%
  mutate(
    Year_notification = as.integer(Year_notification),
    patient_vital_status = if_else(
      patient_vital_status %like% "Unknown", "Alive", patient_vital_status
    ),
    Age_years = as.numeric(Age_years),
    age_group = cut(
      Age_years,
      breaks = c(-Inf, 1, 5, 15, 25, 35, 45, 55, 65, Inf),
      labels = c("<1", "1–4", "5–14", "15–24", "25–34",
                 "35–44", "45–54", "55–64", "65+"),
      right  = FALSE
    ),
    sex = case_when(
      patient_gender %in% c("Male", "M")   ~ "Male",
      patient_gender %in% c("Female", "F") ~ "Female",
      TRUE                                  ~ NA_character_
    ),
    deceased = patient_vital_status %in% "Deceased",
    tb_type = recode(condition, !!!TB_LABELS),
    # Drug resistance category
    dr_category = case_when(
      condition == "Tuberculosis: extensively drug -resistant (XDR -TB)" ~ "XDR-TB",
      condition == "Tuberculosis: multidrug- resistant (MDR -TB)"        ~ "MDR-TB",
      TRUE                                                                ~ "Drug-susceptible/Unknown"
    ),
    # Site category
    site_category = case_when(
      condition == "Tuberculosis:extra-pulmonary" ~ "Extra-pulmonary",
      TRUE                                        ~ "Pulmonary"
    )
  ) -> tb

# ── Shared theme ──────────────────────────────────────────────────────────────
theme_nmc <- function(...) {
  theme_classic() +
    theme(
      text              = element_text(family = "Century Gothic", size = 10, color = "#333333"),
      panel.background  = element_rect(fill = "#F8F8F8"),
      plot.background   = element_rect(fill = "#F8F8F8"),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      axis.ticks        = element_line(linewidth = 0.15, color = "grey40"),
      legend.background = element_rect(fill = "#F8F8F8", linewidth = 0.25,
                                       linetype = "solid", colour = "grey50"),
      legend.key        = element_blank(),
      strip.background  = element_rect(color = NA, fill = NA),
      ...
    )
}

# TB-specific colour palette
tb_cols <- c(
  "Pulmonary TB"       = "#1565C0",
  "Extra-pulmonary TB" = "#7B1FA2",
  "MDR-TB"             = "#E65100",
  "XDR-TB"             = "#B71C1C"
)
```

::: {.callout-note appearance="minimal" collapse="true"}
## Report metadata
| | |
|---|---|
| **Conditions** | `r paste(TB_CONDITIONS, collapse = "; ")` |
| **Years** | 2020 – `r lubridate::year(Sys.Date())` |
| **Data source** | NMC Notifiable Medical Conditions surveillance database |
| **Generated** | `r format(Sys.Date(), "%d %B %Y")` |
| **Note** | This report covers **notifiable TB forms** only (pulmonary, extra-pulmonary, MDR-TB, XDR-TB). Routine TB case data is managed separately through the ETR.Net system. |
:::

---

# Situation Update for Policy Makers {#sec-executive}

```{r exec-setup}
#| label: exec-setup
#| include: false

# ── Current year and comparison calculations ──────────────────────────────────
current_year <- year(Sys.Date())
prev_year    <- current_year - 1
current_month <- month(Sys.Date())

# Year-to-date (Jan 1 to current date) for current and previous year
ytd_current <- tb %>%
  filter(Year_notification == current_year,
         month(as_date(notification_date)) <= current_month)
ytd_prev <- tb %>%
  filter(Year_notification == prev_year,
         month(as_date(notification_date)) <= current_month)

n_ytd_current   <- nrow(ytd_current)
n_ytd_prev      <- nrow(ytd_prev)
deaths_ytd_curr <- sum(ytd_current$deceased, na.rm = TRUE)
deaths_ytd_prev <- sum(ytd_prev$deceased, na.rm = TRUE)
cfr_ytd_curr    <- round(deaths_ytd_curr / max(n_ytd_current, 1) * 100, 1)
cfr_ytd_prev    <- round(deaths_ytd_prev / max(n_ytd_prev, 1) * 100, 1)

pct_change_n      <- round((n_ytd_current - n_ytd_prev) / max(n_ytd_prev, 1) * 100)
pct_change_deaths <- round((deaths_ytd_curr - deaths_ytd_prev) / max(deaths_ytd_prev, 1) * 100)

# Direction arrows and colours
arrow_n      <- if (pct_change_n > 0) "↑" else if (pct_change_n < 0) "↓" else "→"
arrow_deaths <- if (pct_change_deaths > 0) "↑" else if (pct_change_deaths < 0) "↓" else "→"
col_n        <- if (pct_change_n > 10) "#C62828" else if (pct_change_n < -10) "#2E7D32" else "#F57C00"
col_deaths   <- if (pct_change_deaths > 10) "#C62828" else if (pct_change_deaths < -10) "#2E7D32" else "#F57C00"

# Drug-resistant TB counts
mdr_ytd_curr <- sum(ytd_current$dr_category == "MDR-TB")
mdr_ytd_prev <- sum(ytd_prev$dr_category == "MDR-TB")
xdr_ytd_curr <- sum(ytd_current$dr_category == "XDR-TB")
xdr_ytd_prev <- sum(ytd_prev$dr_category == "XDR-TB")
pct_change_mdr <- round((mdr_ytd_curr - mdr_ytd_prev) / max(mdr_ytd_prev, 1) * 100)
pct_change_xdr <- round((xdr_ytd_curr - xdr_ytd_prev) / max(xdr_ytd_prev, 1) * 100)
arrow_mdr <- if (pct_change_mdr > 0) "↑" else if (pct_change_mdr < 0) "↓" else "→"
arrow_xdr <- if (pct_change_xdr > 0) "↑" else if (pct_change_xdr < 0) "↓" else "→"

# Extra-pulmonary proportion
eptb_ytd_curr <- sum(ytd_current$site_category == "Extra-pulmonary")
eptb_pct_curr <- round(eptb_ytd_curr / max(n_ytd_current, 1) * 100, 1)

# Top 3 provinces this year
top3_prov <- ytd_current %>%
  count(prov_, sort = TRUE) %>%
  slice(1:3) %>%
  pull(prov_) %>%
  paste(collapse = ", ")
```

::: {.panel-tabset}

## Key Metrics

```{r exec-metrics}
#| label: tbl-exec-metrics
#| tbl-cap: !expr 'paste0("Year-to-date comparison: Jan–", format(Sys.Date(), "%b %Y"), " vs same period ", prev_year)'

tibble(
  Indicator = c(
    "All TB notifications (YTD)",
    "Deaths (YTD)",
    "Case Fatality Ratio",
    "MDR-TB notifications",
    "XDR-TB notifications",
    "Extra-pulmonary TB (%)",
    "Top 3 provinces (by case count)"
  ),
  `Current Year` = c(
    scales::comma(n_ytd_current),
    scales::comma(deaths_ytd_curr),
    paste0(cfr_ytd_curr, "%"),
    scales::comma(mdr_ytd_curr),
    scales::comma(xdr_ytd_curr),
    paste0(eptb_pct_curr, "%"),
    top3_prov
  ),
  `Same Period Last Year` = c(
    scales::comma(n_ytd_prev),
    scales::comma(deaths_ytd_prev),
    paste0(cfr_ytd_prev, "%"),
    scales::comma(mdr_ytd_prev),
    scales::comma(xdr_ytd_prev),
    "—",
    "—"
  ),
  `Change` = c(
    paste0(arrow_n, " ", ifelse(pct_change_n >= 0, "+", ""), pct_change_n, "%"),
    paste0(arrow_deaths, " ", ifelse(pct_change_deaths >= 0, "+", ""), pct_change_deaths, "%"),
    paste0(ifelse(cfr_ytd_curr >= cfr_ytd_prev, "↑", "↓"), " ",
           ifelse(cfr_ytd_curr - cfr_ytd_prev >= 0, "+", ""),
           round(cfr_ytd_curr - cfr_ytd_prev, 1), " pp"),
    paste0(arrow_mdr, " ", ifelse(pct_change_mdr >= 0, "+", ""), pct_change_mdr, "%"),
    paste0(arrow_xdr, " ", ifelse(pct_change_xdr >= 0, "+", ""), pct_change_xdr, "%"),
    "—",
    "—"
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = 1:3, j = 1) %>%
  color(i = 1, j = 4, color = col_n) %>%
  color(i = 2, j = 4, color = col_deaths) %>%
  bg(i = 4:5, bg = "#FFF3E0") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 11, part = "all") %>%
  align(j = 2:4, align = "center", part = "all") %>%
  add_footer_lines(paste0("Data extracted: ", format(Sys.Date(), "%d %B %Y"),
                          ". YTD = Year-to-date (Jan 1 – ", format(Sys.Date(), "%d %b"), "). pp = percentage points."))
```

## 12-Month Trend by TB Type

```{r exec-trend}
#| label: fig-exec-trend
#| fig-cap: "Monthly TB notification trend by type — last 12 months. Drug-resistant TB forms highlighted."
#| fig-height: 5

# Last 12 complete months
end_date   <- floor_date(Sys.Date(), "month") - days(1)
start_date <- floor_date(end_date - months(11), "month")

tb %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  filter(month_date >= start_date, month_date <= end_date) %>%
  count(month_date, tb_type) %>%
  complete(month_date, tb_type, fill = list(n = 0)) -> trend12_df

ggplot(trend12_df, aes(x = month_date, y = n, fill = tb_type)) +
  geom_col(width = 22) +
  geom_text(aes(label = ifelse(n > 0, n, "")), 
            position = position_stack(vjust = 0.5),
            size = 2.5, colour = "white", fontface = "bold") +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "1 month", expand = expansion(add = c(10, 10))) +
  scale_y_continuous(name = "Notifications", expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = tb_cols, name = "TB Type") +
  labs(x = NULL) +
  theme_nmc() +
  theme(axis.text.x = element_text(size = 8), legend.position = "bottom")
```

## Provincial Burden

```{r exec-prov-heatmap}
#| label: fig-exec-prov
#| fig-cap: "Provincial burden heatmap — quarterly TB notifications by type, last 2 years."
#| fig-height: 6

# Get provincial population for incidence calculation
prov_pop <- NMCleaner::pop %>%
  filter(as.integer(Year) == year(Sys.Date())) %>%
  group_by(prov) %>%
  summarise(population = sum(Population), .groups = "drop")

# Last 8 quarters
tb %>%
  filter(!is.na(notification_date)) %>%
  mutate(
    nd  = as_date(notification_date),
    qtr = paste0(year(nd), " Q", quarter(nd))
  ) %>%
  filter(nd >= (Sys.Date() - years(2))) %>%
  group_by(prov_, qtr) %>%
  summarise(
    n      = n(),
    deaths = sum(deceased, na.rm = TRUE),
    n_mdr  = sum(dr_category == "MDR-TB"),
    n_xdr  = sum(dr_category == "XDR-TB"),
    .groups = "drop"
  ) %>%
  complete(prov_, qtr, fill = list(n = 0, deaths = 0, n_mdr = 0, n_xdr = 0)) %>%
  left_join(prov_pop, by = c("prov_" = "prov")) %>%
  mutate(
    inc       = round(n / population * 1e5 * 4, 1),  # annualized
    dr_pct    = round((n_mdr + n_xdr) / pmax(n, 1) * 100, 0)
  ) -> prov_qtr_df

# Order quarters chronologically
qtr_order <- prov_qtr_df %>% distinct(qtr) %>% arrange(qtr) %>% pull(qtr)
prov_qtr_df$qtr <- factor(prov_qtr_df$qtr, levels = qtr_order)

ggplot(prov_qtr_df, aes(x = qtr, y = prov_, fill = inc)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(n, ifelse(deaths > 0, paste0(" (", deaths, "†)"), ""),
                               "\n", dr_pct, "% DR")),
            size = 2.3, colour = "black", lineheight = 0.85) +
  scale_fill_gradient(low = "#E3F2FD", high = "#0D47A1", name = "Incidence\n(per 100k/yr)") +
  labs(x = "Quarter", y = NULL,
       caption = "Top = notifications (†deaths); bottom = % drug-resistant (MDR + XDR).") +
  theme_nmc() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "right",
    panel.grid = element_blank()
  )
```

:::

---

# Summary Statistics {#sec-summary}

```{r summary-stats}
#| label: tbl-summary
#| tbl-cap: "Summary of TB notifications by type, South Africa, 2020–present."

total_n      <- nrow(tb)
total_deaths <- sum(tb$deceased, na.rm = TRUE)
overall_cfr  <- round(total_deaths / total_n * 100, 1)

tb %>%
  group_by(tb_type) %>%
  summarise(
    n      = n(),
    deaths = sum(deceased, na.rm = TRUE),
    cfr    = round(deaths / n * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n)) -> type_summary

tibble(
  Metric = c(
    "Total TB notifications",
    "Total deaths reported",
    "Overall case fatality ratio",
    "",
    type_summary$tb_type,
    "",
    "Provinces affected",
    "Districts affected",
    "Unique reporting facilities",
    "Date range"
  ),
  Value = c(
    scales::comma(total_n),
    scales::comma(total_deaths),
    paste0(overall_cfr, "%"),
    "--- By TB type ---",
    paste0(scales::comma(type_summary$n), " (CFR ", type_summary$cfr, "%)"),
    "",
    as.character(tb$prov_ %>% na.omit() %>% n_distinct()),
    as.character(tb$district %>% na.omit() %>% n_distinct()),
    as.character(tb$facility %>% na.omit() %>% n_distinct()),
    paste(
      format(as_date(min(tb$notification_date, na.rm = TRUE)), "%d %b %Y"),
      "to",
      format(as_date(max(tb$notification_date, na.rm = TRUE)), "%d %b %Y")
    )
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = c(1, 4, 5:8), j = 1) %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 10, part = "all")
```

---

# Temporal Trends {#sec-trends}

## Annual notifications by TB type

```{r annual-trend}
#| label: fig-annual-trend
#| fig-cap: "Annual TB notifications by type and case fatality ratio, South Africa, 2020–present."
#| fig-height: 5

tb %>%
  group_by(year = Year_notification, tb_type) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") -> annual_type

tb %>%
  group_by(year = Year_notification) %>%
  summarise(total = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / total * 100, .groups = "drop") -> annual_cfr

sf_ax <- max(annual_type %>% group_by(year) %>% summarise(n = sum(n)) %>% pull(n)) / 
         max(annual_cfr$cfr, na.rm = TRUE) * 1.3

ggplot(annual_type, aes(x = year, y = n, fill = tb_type)) +
  geom_col(width = 0.65) +
  geom_line(data = annual_cfr, aes(x = year, y = cfr * sf_ax, fill = NULL),
            colour = "#C62828", linewidth = 1.2, group = 1) +
  geom_point(data = annual_cfr, aes(x = year, y = cfr * sf_ax, fill = NULL),
             colour = "#C62828", size = 3) +
  geom_text(data = annual_cfr, aes(x = year, y = cfr * sf_ax + max(cfr * sf_ax) * 0.05,
                                   label = paste0(round(cfr, 1), "%"), fill = NULL),
            size = 3, colour = "#C62828") +
  scale_y_continuous(
    name     = "Notifications (n)",
    sec.axis = sec_axis(~ . / sf_ax, name = "CFR (%)", breaks = pretty_breaks(4))
  ) +
  scale_fill_manual(values = tb_cols, name = "TB Type") +
  scale_x_continuous(breaks = unique(annual_type$year)) +
  labs(x = "Year", caption = "Red line = overall case fatality ratio") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Monthly epicurve

```{r epicurve-monthly}
#| label: fig-epicurve-monthly
#| fig-cap: "Monthly TB notification counts by type, South Africa, 2020–present."
#| fig-height: 5

tb %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  count(month_date, tb_type) %>%
  complete(month_date, tb_type, fill = list(n = 0)) -> epicurve_df

ggplot(epicurve_df, aes(x = month_date, y = n, fill = tb_type)) +
  geom_col(width = 25) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months",
               date_minor_breaks = "3 months", expand = expansion(add = c(0, 0))) +
  scale_y_continuous(name = "Notifications (n)", breaks = pretty_breaks(5)) +
  scale_fill_manual(values = tb_cols, name = "TB Type") +
  labs(x = "Date") +
  theme_nmc() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
        legend.position = "bottom")
```

## Drug-resistant TB proportion over time

```{r dr-trend}
#| label: fig-dr-trend
#| fig-cap: "Monthly proportion of TB notifications that are drug-resistant (MDR + XDR), South Africa."
#| fig-height: 4

tb %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  group_by(month_date) %>%
  summarise(
    n_total = n(),
    n_dr    = sum(dr_category %in% c("MDR-TB", "XDR-TB")),
    pct_dr  = n_dr / n_total * 100,
    .groups = "drop"
  ) %>%
  mutate(rolling3 = zoo::rollmean(pct_dr, k = 3, fill = NA, align = "right")) -> dr_trend

ggplot(dr_trend, aes(x = month_date)) +
  geom_col(aes(y = pct_dr), fill = "#E65100", alpha = 0.4, width = 25) +
  geom_line(aes(y = rolling3), colour = "#B71C1C", linewidth = 1.2, na.rm = TRUE) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
  scale_y_continuous(name = "% Drug-resistant", limits = c(0, NA)) +
  labs(x = NULL, caption = "Red line = 3-month rolling average") +
  theme_nmc()
```

---

# Age & Sex Demographics {#sec-demographics}

## Population pyramid

```{r age-sex-pyramid}
#| label: fig-age-sex-pyramid
#| fig-cap: "Age–sex population pyramid of TB notifications, South Africa, 2020–present."
#| fig-height: 5

tb %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  count(age_group, sex) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyramid_df

max_n <- max(abs(pyramid_df$n))
step  <- ifelse(max_n > 500, 200, ifelse(max_n > 200, 50, 25))
br    <- seq(0, ceiling(max_n / step) * step, step)

ggplot(pyramid_df, aes(x = age_group, y = n_plot, fill = sex)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  geom_hline(yintercept = 0, colour = "white", linewidth = 0.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y = -max_n * 0.5, label = "Male",   color = "#1565C0", fontface = "bold", size = 3.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y =  max_n * 0.5, label = "Female", color = "#C62828", fontface = "bold", size = 3.5) +
  labs(x = "Age group (years)", y = "Number of notifications") +
  theme_nmc() +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(color = "#DDDDDD", linewidth = 0.3))
```

## Age distribution by TB type

```{r age-by-type}
#| label: fig-age-by-type
#| fig-cap: "Age distribution of TB notifications by TB type, South Africa, 2020–present."
#| fig-height: 5

tb %>%
  filter(!is.na(age_group)) %>%
  count(age_group, tb_type) %>%
  group_by(tb_type) %>%
  mutate(pct = n / sum(n) * 100) -> age_type_df

ggplot(age_type_df, aes(x = age_group, y = pct, fill = tb_type)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = tb_cols, name = "TB Type") +
  labs(x = "Age group (years)", y = "% within TB type") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Case fatality ratio by age group and TB type

```{r cfr-age-type}
#| label: fig-cfr-age-type
#| fig-cap: "Case fatality ratio (%) by age group and TB type, South Africa, 2020–present."
#| fig-height: 5

tb %>%
  filter(!is.na(age_group)) %>%
  group_by(age_group, tb_type) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") -> cfr_df

ggplot(cfr_df, aes(x = age_group, y = cfr, fill = tb_type)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = ifelse(cfr > 0, paste0(round(cfr, 1), "%"), "")),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 2.5, color = "#333333") +
  scale_fill_manual(values = tb_cols, name = "TB Type") +
  scale_y_continuous(limits = c(0, max(cfr_df$cfr, na.rm = TRUE) * 1.3)) +
  labs(x = "Age group (years)", y = "Case fatality ratio (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Summary demographics table

```{r demo-table}
#| label: tbl-demographics
#| tbl-cap: "Summary demographics of TB notifications by TB type and vital status, South Africa, 2020–present."

tb %>%
  select(tb_type, sex, age_group, Age_years, patient_vital_status, prov_) %>%
  tbl_strata(
    strata   = tb_type,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by        = patient_vital_status,
        missing   = "no",
        label     = list(sex ~ "Sex", age_group ~ "Age group",
                         Age_years ~ "Age (years)", prov_ ~ "Province"),
        type      = list(Age_years ~ "continuous"),
        statistic = list(Age_years ~ "{median} ({p25}–{p75})")
      ) %>%
      bold_labels()
  ) %>%
  as_flex_table()
```

---

# Geographical Distribution {#sec-geo}

```{r geo-setup}
#| include: false

# ── Provincial incidence ──────────────────────────────────────────────────────
NMCleaner::pop %>%
  mutate(Year = as.integer(Year)) %>%
  group_by(prov, Year) %>%
  summarise(population = sum(Population), .groups = "drop") -> pop_prov

tb %>%
  group_by(prov_, year = Year_notification) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(pop_prov, by = c("prov_" = "prov", "year" = "Year")) %>%
  mutate(incidence = n / population * 1e5) -> incidence_prov_df

# ── District incidence ────────────────────────────────────────────────────────
NMCleaner::pop %>%
  filter(as.integer(Year) %in% 2020:year(Sys.Date())) %>%
  group_by(district_standard) %>%
  summarise(pop_total = sum(Population), .groups = "drop") -> pop_dist

tb %>%
  filter(!district %in% c("Unknown", "Isazi Import", "Not Applicable"),
         !is.na(district)) %>%
  mutate_district_name("district") %>%
  group_by(Province = prov_, district_standard) %>%
  summarise(
    Notifications = n(),
    Deaths        = sum(deceased, na.rm = TRUE),
    CFR           = round(Deaths / Notifications * 100, 1),
    n_mdr         = sum(dr_category == "MDR-TB"),
    n_xdr         = sum(dr_category == "XDR-TB"),
    .groups = "drop"
  ) %>%
  left_join(pop_dist, by = "district_standard") %>%
  mutate(
    Incidence = round(Notifications / pop_total * 1e5, 1),
    DR_pct    = round((n_mdr + n_xdr) / Notifications * 100, 1),
    pop_total = round(pop_total / 1e3, 1)
  ) -> incidence_dist_df
```

## Provincial incidence over time

```{r prov-incidence-line}
#| label: fig-prov-incidence-line
#| fig-cap: "TB notification incidence per 100,000 population by province and year, South Africa."
#| fig-height: 5.5

incidence_prov_df %>%
  group_by(prov_) %>%
  filter(incidence == max(incidence, na.rm = TRUE)) %>%
  ungroup() -> peak_prov_df

ggplot(incidence_prov_df, aes(x = year, y = incidence, colour = prov_, group = prov_)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  ggrepel::geom_label_repel(data = peak_prov_df, aes(label = prov_),
    size = 2.8, show.legend = FALSE, label.size = 0.15, max.overlaps = 20) +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  scale_x_continuous(breaks = unique(incidence_prov_df$year)) +
  labs(x = "Year", y = "Incidence per 100,000 population") +
  theme_nmc() +
  theme(legend.position = "none")
```

## Drug-resistant TB by province

```{r dr-by-prov}
#| label: fig-dr-by-prov
#| fig-cap: "Proportion of TB notifications that are drug-resistant (MDR + XDR) by province, South Africa, 2020–present."
#| fig-height: 4.5

tb %>%
  group_by(prov_) %>%
  summarise(
    n_total = n(),
    n_dr    = sum(dr_category %in% c("MDR-TB", "XDR-TB")),
    pct_dr  = n_dr / n_total * 100,
    .groups = "drop"
  ) %>%
  mutate(prov_ = fct_reorder(prov_, pct_dr)) -> dr_prov

ggplot(dr_prov, aes(x = prov_, y = pct_dr)) +
  geom_col(fill = "#E65100", width = 0.7) +
  geom_text(aes(label = paste0(round(pct_dr, 1), "%")), hjust = -0.2, size = 3.2) +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(dr_prov$pct_dr) * 1.2)) +
  labs(x = NULL, y = "% Drug-resistant TB (MDR + XDR)") +
  theme_nmc()
```

## Top 20 districts by notification count

```{r top-districts}
#| label: tbl-top-districts
#| tbl-cap: "Top 20 districts by total TB notification count, South Africa, 2020–present."

incidence_dist_df %>%
  arrange(desc(Notifications)) %>%
  slice(1:20) %>%
  rename(District = district_standard, `Pop (k)` = pop_total,
         `DR %` = DR_pct) %>%
  select(Province, District, Notifications, Deaths, `CFR (%)` = CFR,
         `Incidence per 100k` = Incidence, `DR %`, `Pop (k)`) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  bg(i = ~ `DR %` > 10, j = "DR %", bg = "#FFF3E0") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 9, part = "all") %>%
  add_footer_lines("Pink = CFR > 5%. Orange = DR > 10%.")
```

---

# Mortality Profile {#sec-mortality}

## Deaths by TB type and year

```{r deaths-type}
#| label: fig-deaths-type
#| fig-cap: "Annual TB deaths by TB type, South Africa."
#| fig-height: 5

tb %>%
  filter(deceased) %>%
  count(year = Year_notification, tb_type) %>%
  ggplot(aes(x = factor(year), y = n, fill = tb_type)) +
  geom_col(position = "stack", width = 0.75) +
  scale_fill_manual(values = tb_cols, name = "TB Type") +
  labs(x = "Year", y = "Deaths (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## CFR by TB type over time

```{r cfr-type-time}
#| label: fig-cfr-type-time
#| fig-cap: "Case fatality ratio (%) by TB type and year, South Africa."
#| fig-height: 5

tb %>%
  group_by(year = Year_notification, tb_type) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") %>%
  ggplot(aes(x = factor(year), y = cfr, fill = tb_type)) +
  geom_col(position = "dodge", width = 0.85) +
  geom_text(aes(label = paste0(round(cfr, 1), "%")),
            position = position_dodge(width = 0.85),
            vjust = -0.3, size = 2.5) +
  scale_fill_manual(values = tb_cols, name = "TB Type") +
  labs(x = "Year", y = "CFR (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

---

# Reporting & Facility Analysis {#sec-facilities}

## Unique reporting facilities per year

```{r facilities-trend}
#| label: fig-facilities
#| fig-cap: "Number of unique health facilities reporting at least one TB notification per year, South Africa."
#| fig-height: 4

tb %>%
  filter(!is.na(facility)) %>%
  group_by(Year_notification) %>%
  summarise(n_facilities = n_distinct(facility), .groups = "drop") -> fac_df

ggplot(fac_df, aes(x = factor(Year_notification))) +
  geom_col(aes(y = n_facilities, fill = "Unique facilities"), width = 0.6, alpha = 0.85) +
  geom_text(aes(y = n_facilities, label = n_facilities), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Unique facilities" = "#4CAF50"), name = NULL) +
  labs(x = "Year", y = "Unique reporting facilities (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

---

::: {.callout-note}
## Interpretation Notes

**Drug-resistant TB (DR-TB):**

- **MDR-TB** = Multidrug-resistant TB, resistant to at least isoniazid and rifampicin
- **XDR-TB** = Extensively drug-resistant TB, MDR-TB plus resistance to fluoroquinolones and at least one injectable second-line drug

**Data limitations:**

- This report covers notifiable TB forms reported through NMC only
- Routine TB case data is managed through the ETR.Net electronic TB register system
- Case fatality reflects deaths among notified cases; actual mortality may differ

See [Data Quality](../../quality.html) for reporting completeness and [Methods](../../methods.html) for full definitions.
:::

---

*Report generated: `r format(Sys.Date(), "%d %B %Y")` | NMC Surveillance Unit*
