---
title: "National Overview"
subtitle: "High-level surveillance snapshot across all notifiable conditions"
format:
  html:
    page-layout: full
    toc: true
---

::: {.callout-note collapse="true"}
## Development Roadmap — National Overview

| Phase | Feature | Status |
|-------|---------|--------|
| **v0.1** | Summary value boxes (total cases, conditions, provinces) | ✅ Done |
| **v0.1** | Top-10 conditions bar chart (current month) | ✅ Done |
| **v0.1** | National weekly epicurve (all conditions combined) | ✅ Done |
| **v0.2** | Provincial choropleth map (total cases) | 🔲 Planned |
| **v0.2** | Month-on-month change indicators (↑/↓) | 🔲 Planned |
| **v0.3** | Age/sex pyramid (when demographic data available) | 🔲 Planned |
| **v0.3** | Seasonal overlay / historical comparison bands | 🔲 Planned |
| **v0.4** | EpiNow2 national Rt estimate cards | 🔲 Planned |

*Data pipeline:* `R/prepare_dashboard_data.r` → `data/processed/agg_national.rds`, `agg_province.rds`
:::

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(data.table)
  library(plotly)
  library(DT)
  library(lubridate)
})

agg_nat  <- readRDS("data/processed/agg_national.rds")
agg_prov <- readRDS("data/processed/agg_province.rds")
setDT(agg_nat); setDT(agg_prov)

if (file.exists("data/processed/agg_source.rds")) {
  agg_source <- readRDS("data/processed/agg_source.rds")
  setDT(agg_source)
} else {
  agg_source <- NULL
}

if (file.exists("data/processed/agg_timeliness.rds")) {
  agg_time <- readRDS("data/processed/agg_timeliness.rds")
  setDT(agg_time)
} else {
  agg_time <- NULL
}

agg_prov <- agg_prov[!is.na(prov_) & prov_ != ""]

latest_date <- max(agg_nat$date)
month_start <- floor_date(latest_date, "month")
prev_month  <- month_start - months(1)
```

## Key Figures — `r format(month_start, "%B %Y")`

```{r}
#| label: kpi-cards
#| results: asis
this_month <- agg_nat[date >= month_start, .(cases = sum(count))]$cases
last_month <- agg_nat[date >= prev_month & date < month_start, .(cases = sum(count))]$cases
n_cond     <- agg_nat[date >= month_start & count > 0, uniqueN(condition)]
n_prov     <- agg_prov[date >= month_start & count > 0, uniqueN(prov_)]

pct_change <- ifelse(last_month > 0,
                     round((this_month - last_month) / last_month * 100, 1),
                     NA_real_)
arrow <- ifelse(!is.na(pct_change) & pct_change >= 0, "↑", "↓")

cat(sprintf(
'
::: {layout-ncol=4}

::: {.card .bg-primary .text-white .p-3 .text-center}
**Total Cases**

### %s
:::

::: {.card .bg-info .text-white .p-3 .text-center}
**Month Δ**

### %s %s%%
:::

::: {.card .bg-success .text-white .p-3 .text-center}
**Conditions Reported**

### %s
:::

::: {.card .bg-warning .text-dark .p-3 .text-center}
**Provinces Reporting**

### %s / 9
:::

:::
',
  format(this_month, big.mark = ","),
  arrow, abs(pct_change),
  n_cond, n_prov
))
```

---

## Top Conditions — Current Month {.tabset}

### Bar Chart

```{r}
#| label: fig-top-conditions
#| fig-cap: "Top 10 conditions by case count this month"
top10 <- agg_nat[date >= month_start, .(cases = sum(count)), by = condition]
setorder(top10, -cases)
top10 <- head(top10, 10)

plot_ly(top10, x = ~reorder(condition, cases), y = ~cases,
        type = "bar", orientation = "h",
        marker = list(color = "#1d4ed8"),
        hoverinfo = "text",
        text = ~paste0(condition, ": ", format(cases, big.mark = ","))) %>%
  layout(xaxis = list(title = "Cases"),
         yaxis = list(title = ""),
         margin = list(l = 250))
```

### Table

```{r}
#| label: tbl-top-conditions
all_conds <- agg_nat[date >= month_start, .(Cases = sum(count)), by = .(Condition = condition)]
setorder(all_conds, -Cases)
DT::datatable(all_conds, rownames = FALSE,
              options = list(pageLength = 15, dom = "ftp"))
```

---

## National Epicurve — Weekly

```{r}
#| label: fig-national-epicurve
#| fig-cap: "All notifiable conditions combined, weekly"
agg_nat[, week := floor_date(date, "week")]
weekly_all <- agg_nat[, .(cases = sum(count)), by = week]

plot_ly(weekly_all, x = ~week, y = ~cases, type = "bar",
        marker = list(color = "#0b3a75"),
        hoverinfo = "text",
        text = ~paste0("Week of ", format(week, "%d %b %Y"),
                       "<br>Cases: ", format(cases, big.mark = ","))) %>%
  layout(xaxis = list(title = "Week starting"),
         yaxis = list(title = "Weekly cases"))
```

---

## Surveillance Performance

::: {.panel-tabset}

### Cases by Source

```{r}
#| label: fig-source
#| fig-cap: "Notifications by Case Type (Clinical vs Lab etc.)"

if (!is.null(agg_source) && nrow(agg_source) > 0) {
    # Aggregate across conditions for a simpler overview chart
    # Or stack by condition + case_type? Just stack by `case_type` over time.
    agg_src_total <- agg_source[, .(count = sum(count)), by = .(month, case_type)]
    setorder(agg_src_total, month)

    plot_ly(agg_src_total, x = ~month, y = ~count, color = ~case_type,
            type = 'bar', 
            colors = "Set2") %>%
      layout(barmode = 'stack',
             xaxis = list(title = "Month"),
             yaxis = list(title = "Notifications"))
} else {
    cat("Data for Cases by Source not available.")
}
```

### Timeliness (Median Days)

```{r}
#| label: fig-timeliness
#| fig-cap: "Median days to notification by condition (last 12 months)"

if (!is.null(agg_time) && nrow(agg_time) > 0) {
    # Filter to last 12 months for readability
    subset_time <- agg_time[month >= (max(month) - years(1))]
    
    # We can do a heatmap or a line chart
    # Let's do a heatmap: X=Month, Y=Condition, Z=Median Days
    # Only show conditions with significant activity
    
    top_conds <- agg_nat[date >= (max(date) - years(1)), .(total = sum(count)), by=condition][order(-total)][1:15, condition]
    
    subset_time <- subset_time[condition %in% top_conds]
    
    plot_ly(subset_time, x = ~month, y = ~condition, z = ~median_days,
            type = "heatmap", colors = "Reds") %>%
      layout(xaxis = list(title = "Month"),
             yaxis = list(title = "", automargin=TRUE))

} else {
    cat("Data for Timeliness not available.")
}
```

:::

---

::: {.callout-tip}
## Drill Down Further
- [By Province](by-province/index.html) — detailed provincial breakdowns with district drilldown
- [By Condition](by-condition/index.html) — condition-level national trends and hotspots
- [Trending](by-trending/index.html) — signal detection for emerging conditions
- [Explorer](explorer/index.html) — custom queries and data downloads
- [Data Quality](quality.html) — reporting completeness and timeliness metrics
:::
