---
title: "Data Quality"
subtitle: "Completeness, timeliness & validation checks on NMC surveillance data"
format:
  html:
    toc: true
---

::: {.callout-note collapse="true"}
## Development Roadmap — Data Quality

| Phase | Feature | Status |
|-------|---------|--------|
| **v0.1** | Reporting completeness by province (% days with ≥ 1 case) | ✅ Done |
| **v0.1** | Weekly submission heatmap | ✅ Done |
| **v0.2** | Timeliness metrics (median days from onset to notification) | ✅ Done |
| **v0.2** | Linkage-field completeness trend (ID / folder / MRN) | ✅ Done |
| **v0.3** | Duplicate detection summary | 🔲 Planned |
| **v0.3** | Missing-field rates by condition (name, age, sex, district) | 🔲 Planned |
| **v0.3** | Facility-level reporting participation rates | 🔲 Planned |
| **v0.4** | Automated data-quality score card (red/amber/green) | 🔲 Planned |
| **v0.5** | Historic trend of completeness to track improvement | 🔲 Planned |

*Data pipeline:* `R/prepare_dashboard_data.r` → `data/processed/agg_province.rds`
:::

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(data.table)
  library(plotly)
  library(DT)
  library(lubridate)
  library(htmltools)
})

source("R/_load_all.r")

agg_prov <- readRDS("data/processed/agg_province.rds")
setDT(agg_prov)
agg_prov <- agg_prov[!is.na(prov_) & prov_ != ""]
agg_prov[, date := as.Date(date)]

latest_date <- max(agg_prov$date)
```

::: {.data-through-banner}
**Data through:** `r format(latest_date, "%d %B %Y")` <span class="muted">— provincial reporting completeness, weekly volume, timeliness & linkage</span>
:::

---

## Reporting Completeness by Province

Proportion of days in the last 90 days where at least one case was notified, aggregated across all conditions.

```{r}
#| label: fig-completeness
#| fig-cap: "Reporting completeness (last 90 days)"
last90 <- agg_prov[date >= (latest_date - 90)]
total_days <- as.integer(difftime(latest_date, latest_date - 90, units = "days"))

completeness <- last90[count > 0, .(
  days_reported = uniqueN(date)
), by = .(Province = prov_)]
completeness[, pct := round(days_reported / total_days * 100, 1)]
completeness[, Province := prov_label(Province)]

plot_ly(completeness, x = ~Province, y = ~pct,
        type = "bar",
        marker = list(color = ~ifelse(pct >= 80, "#065f46", ifelse(pct >= 50, "#b45309", "#b91c1c"))),
        hoverinfo = "text",
        text = ~paste0(Province, ": ", pct, "% (", days_reported, "/", total_days, " days)")) %>%
  layout(yaxis = list(title = "% Days Reporting", range = c(0, 100)),
         xaxis = list(title = "Province"),
         shapes = list(
           list(type = "line", x0 = -0.5, x1 = 8.5,
                y0 = 80, y1 = 80, line = list(dash = "dash", color = "#065f46"))))
```

---

## Weekly Submission Volume

```{r}
#| label: fig-heatmap
#| fig-cap: "Weekly case volume by province (last 52 weeks)"
last_year <- agg_prov[date >= (latest_date - 365)]
last_year[, week := floor_date(date, "week")]
heat_dt <- last_year[, .(cases = sum(count)), by = .(week, Province = prov_)]
heat_dt[, Province := prov_label(Province)]

plot_ly(heat_dt, x = ~week, y = ~Province, z = ~cases,
        type = "heatmap",
        colorscale = list(c(0, "#e6f2ff"), c(1, "#1d4ed8")),
        hoverinfo = "text",
        text = ~paste0(Province, " — wk ", format(week, "%d %b %Y"),
                       "<br>Cases: ", format(cases, big.mark = ","))) %>%
  layout(xaxis = list(title = "Week"),
         yaxis = list(title = ""))
```

---

## Completeness Table

```{r}
#| label: tbl-completeness
DT::datatable(completeness, rownames = FALSE,
              options = list(pageLength = 10, dom = "t"))
```

---

## Notification Timeliness

Median days from symptom onset (or earliest available date) to NMC
notification, by condition and month. Lower is better; sustained increases
flag a slowdown in the reporting chain.

```{r}
#| label: timeliness-load
#| message: false
#| warning: false
agg_time_path <- "data/processed/agg_timeliness.rds"
agg_time <- if (file.exists(agg_time_path)) {
  as.data.table(readRDS(agg_time_path))
} else {
  data.table()
}
```

```{r}
#| label: fig-timeliness
#| fig-cap: "Median onset-to-notification delay (days) — last 24 months"
if (nrow(agg_time) == 0L) {
  htmltools::tags$p(
    "Timeliness data not available — re-run R/prepare_dashboard_data.r ",
    "with a `time_to_notification` column to populate this section.",
    style = "color:#475569;font-style:italic;"
  )
} else {
  agg_time[, month := as.Date(month)]
  cutoff_24 <- max(agg_time$month) - months(24)
  recent <- agg_time[month >= cutoff_24 & n_cases >= 5]

  # Top 12 conditions by total case volume in the window
  top_cond <- recent[, .(tot = sum(n_cases)), by = condition][
    order(-tot)][seq_len(min(12L, .N)), condition]

  plot_ly(recent[condition %chin% top_cond],
          x = ~month, y = ~median_days,
          color = ~condition, split = ~condition,
          type = "scatter", mode = "lines+markers",
          marker = list(size = 5),
          hoverinfo = "text",
          text = ~paste0(condition, "<br>", format(month, "%b %Y"),
                         "<br>Median: ", median_days, " d",
                         "<br>IQR: ", q25, "–", q75,
                         "<br>n = ", n_cases)) |>
    layout(xaxis  = list(title = "Month"),
           yaxis  = list(title = "Median delay (days)"),
           legend = list(orientation = "h", y = -0.15),
           shapes = list(list(type = "line",
                              x0 = min(recent$month), x1 = max(recent$month),
                              y0 = 7, y1 = 7,
                              line = list(dash = "dash", color = "#65a30d"))))
}
```

```{r}
#| label: tbl-timeliness-current
if (nrow(agg_time) > 0L) {
  current_month <- max(agg_time$month)
  curr <- agg_time[month == current_month & n_cases >= 1]
  setorder(curr, -median_days)
  display <- curr[, .(
    Condition          = condition,
    Month              = format(month, "%b %Y"),
    `Median (days)`    = round(median_days, 1),
    `IQR (days)`       = paste0(round(q25, 1), "–", round(q75, 1)),
    `Cases`            = n_cases
  )]
  DT::datatable(display, rownames = FALSE,
                caption = paste0("Current-month timeliness — ",
                                 format(current_month, "%B %Y")),
                options = list(pageLength = 10, dom = "ftp"))
}
```

---

## Linkage-Field Completeness — Last 12 Months

Proportion of notifications with a non-missing patient identifier (SA ID
number, hospital folder number, MRN) — required for record linkage to lab
results, vital registration and other surveillance systems.

```{r}
#| label: linkage-completeness
#| fig-cap: "Linkage-field completeness (%) by province"
lnk_path <- "data/processed/agg_linkage_prov.rds"
lnk <- if (file.exists(lnk_path)) as.data.table(readRDS(lnk_path)) else data.table()

if (nrow(lnk) == 0L) {
  htmltools::tags$p(
    "Linkage completeness data not available.",
    style = "color:#475569;font-style:italic;"
  )
} else {
  lnk[, month := as.Date(month)]
  cutoff_12 <- max(lnk$month) - months(12)
  lnk12 <- lnk[month >= cutoff_12, .(
    n_total          = sum(n_total),
    n_id_present     = sum(n_id_present),
    n_folder_present = sum(n_folder_present),
    n_mrn_present    = sum(n_mrn_present)
  ), by = prov_]
  lnk12[, `:=`(
    `SA ID`        = round(n_id_present     / n_total * 100, 1),
    `Folder no.`   = round(n_folder_present / n_total * 100, 1),
    `MRN`          = round(n_mrn_present    / n_total * 100, 1)
  )]
  lnk12[, Province := prov_label(prov_)]

  long <- melt(lnk12[, .(Province, `SA ID`, `Folder no.`, MRN)],
               id.vars = "Province",
               variable.name = "Field", value.name = "pct")

  plot_ly(long, x = ~Province, y = ~pct, color = ~Field,
          type = "bar",
          hoverinfo = "text",
          text = ~paste0(Province, " — ", Field, ": ", pct, "%")) |>
    layout(barmode = "group",
           yaxis = list(title = "% complete", range = c(0, 100)),
           xaxis = list(title = ""),
           legend = list(orientation = "h", y = -0.15))
}
```

---

::: {.callout-tip}
## Related Pages
- [Methods](methods.html) — full technical documentation of quality checks and indicator definitions
- [Overview](overview.html) — national-level surveillance summary
- [By Province](by-province/index.html) — province-level data that depends on reporting quality
- [Glossary](glossary.html) — abbreviations and term definitions
:::
