---
title: "Data Quality"
subtitle: "Completeness, timeliness & validation checks on NMC surveillance data"
format:
  html:
    toc: true
---

::: {.callout-note collapse="true"}
## Development Roadmap — Data Quality

| Phase | Feature | Status |
|-------|---------|--------|
| **v0.1** | Reporting completeness by province (% days with ≥ 1 case) | ✅ Done |
| **v0.1** | Weekly submission heatmap | ✅ Done |
| **v0.2** | Timeliness metrics (median days from onset to notification) | 🔲 Planned |
| **v0.2** | Duplicate detection summary | 🔲 Planned |
| **v0.3** | Missing-field rates by condition (name, age, sex, district) | 🔲 Planned |
| **v0.3** | Facility-level reporting participation rates | 🔲 Planned |
| **v0.4** | Automated data-quality score card (red/amber/green) | 🔲 Planned |
| **v0.5** | Historic trend of completeness to track improvement | 🔲 Planned |

*Data pipeline:* `R/prepare_dashboard_data.r` → `data/processed/agg_province.rds`
:::

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(data.table)
  library(plotly)
  library(DT)
  library(lubridate)
})

agg_prov <- readRDS("data/processed/agg_province.rds")
setDT(agg_prov)
agg_prov <- agg_prov[!is.na(prov_) & prov_ != ""]
agg_prov[, date := as.Date(date)]

latest_date <- max(agg_prov$date)
```

**Data through:** `r format(latest_date, "%d %B %Y")`

---

## Reporting Completeness by Province

Proportion of days in the last 90 days where at least one case was notified, aggregated across all conditions.

```{r}
#| label: fig-completeness
#| fig-cap: "Reporting completeness (last 90 days)"
last90 <- agg_prov[date >= (latest_date - 90)]
total_days <- as.integer(difftime(latest_date, latest_date - 90, units = "days"))

completeness <- last90[count > 0, .(
  days_reported = uniqueN(date)
), by = .(Province = prov_)]
completeness[, pct := round(days_reported / total_days * 100, 1)]

plot_ly(completeness, x = ~Province, y = ~pct,
        type = "bar",
        marker = list(color = ~ifelse(pct >= 80, "#065f46", ifelse(pct >= 50, "#b45309", "#b91c1c"))),
        hoverinfo = "text",
        text = ~paste0(Province, ": ", pct, "% (", days_reported, "/", total_days, " days)")) %>%
  layout(yaxis = list(title = "% Days Reporting", range = c(0, 100)),
         xaxis = list(title = "Province"),
         shapes = list(
           list(type = "line", x0 = -0.5, x1 = 8.5,
                y0 = 80, y1 = 80, line = list(dash = "dash", color = "#065f46"))))
```

---

## Weekly Submission Volume

```{r}
#| label: fig-heatmap
#| fig-cap: "Weekly case volume by province (last 52 weeks)"
last_year <- agg_prov[date >= (latest_date - 365)]
last_year[, week := floor_date(date, "week")]
heat_dt <- last_year[, .(cases = sum(count)), by = .(week, Province = prov_)]

plot_ly(heat_dt, x = ~week, y = ~Province, z = ~cases,
        type = "heatmap",
        colorscale = list(c(0, "#e6f2ff"), c(1, "#1d4ed8")),
        hoverinfo = "text",
        text = ~paste0(Province, " — wk ", format(week, "%d %b %Y"),
                       "<br>Cases: ", format(cases, big.mark = ","))) %>%
  layout(xaxis = list(title = "Week"),
         yaxis = list(title = ""))
```

---

## Completeness Table

```{r}
#| label: tbl-completeness
DT::datatable(completeness, rownames = FALSE,
              options = list(pageLength = 10, dom = "t"))
```

---

::: {.callout-tip}
## Related Pages
- [Methods](methods.html) — full technical documentation of quality checks and indicator definitions
- [Overview](overview.html) — national-level surveillance summary
- [By Province](by-province/index.html) — province-level data that depends on reporting quality
- [Glossary](glossary.html) — abbreviations and term definitions
:::
