---
title: "By Province"
subtitle: "Provincial summary of notifiable conditions — counts, incidence & district drilldown"
format:
  html:
    page-layout: full
    toc: true
    toc-depth: 2
---

::: {.callout-note collapse="true"}
## Development Roadmap — By Province

| Phase | Feature | Status |
|-------|---------|--------|
| **v0.1** | Monthly case counts by province × condition (interactive bar chart) | ✅ Done |
| **v0.1** | Incidence per 100 000 dot-plot, filterable by condition | ✅ Done |
| **v0.1** | Searchable & downloadable data table (CSV / Excel) | ✅ Done |
| **v0.2** | Leaflet choropleth map — province-level counts | 🔲 Planned |
| **v0.2** | District drilldown table per province | 🔲 Planned |
| **v0.3** | Click-to-zoom district maps with Leaflet + lightbox | 🔲 Planned |
| **v0.3** | Time-series sparklines per province | 🔲 Planned |
| **v0.4** | Link to per-province sub-pages (EC, GP, …) | 🔲 Planned |
| **v0.5** | EpiNow2 Rt estimates per province fed from `agg_province.rds` | 🔲 Planned |

*Data pipeline:* `R/prepare_dashboard_data.r` → `data/processed/agg_province.rds`
:::

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(data.table)
  library(plotly)
  library(crosstalk)
  library(DT)
  library(lubridate)
  library(htmltools)
})

# ── Load processed data (execute-dir: project ⇒ paths from project root) ──
agg_prov <- readRDS("data/processed/agg_province.rds")
setDT(agg_prov)

# Safety: drop any NA provinces produced by CJ expansion
agg_prov <- agg_prov[!is.na(prov_) & prov_ != ""]

latest_date <- max(agg_prov$date)
month_start <- floor_date(latest_date, "month")

# Monthly summary: province × condition
prov_month <- agg_prov[date >= month_start, .(
  cases          = sum(count),
  incidence_100k = round(sum(count) / first(pop) * 1e5, 2)
), by = .(Province = prov_, Condition = condition)]

# Shared data for linked filtering
sd <- SharedData$new(prov_month)
```

## Filters

Use these controls to filter all plots and the table below.

```{r}
#| label: filters
div(
  style = "display:flex; gap:2rem; flex-wrap:wrap; margin-bottom:1.5rem;",
  div(style = "min-width:220px;", filter_select("cond", "Condition", sd, ~Condition)),
  div(style = "min-width:180px;", filter_slider("cases", "Cases ≥", sd, ~cases, min = 0))
)
```

**Data through:** `r format(latest_date, "%d %B %Y")`

---

## Cases by Province {.tabset}

### Stacked Bar Chart

```{r}
#| label: fig-prov-bar
#| fig-cap: "Monthly case counts stacked by condition (current month)"
plot_ly(sd, x = ~Province, y = ~cases, color = ~Condition,
        type = "bar", hoverinfo = "text",
        text = ~paste0(Condition, "<br>", Province, ": ", cases, " cases")) %>%
  layout(barmode = "stack",
         xaxis = list(title = "Province", categoryorder = "total descending"),
         yaxis = list(title = "Cases"),
         legend = list(orientation = "h", y = -0.2))
```

### Incidence Dot-plot

```{r}
#| label: fig-prov-incidence
#| fig-cap: "Incidence per 100 000 population by province (current month)"
plot_ly(sd, x = ~Province, y = ~incidence_100k, color = ~Condition,
        type = "scatter", mode = "markers",
        marker = list(size = 10, opacity = 0.8),
        hoverinfo = "text",
        text = ~paste0(Condition, "<br>", Province,
                       ": ", incidence_100k, " per 100k")) %>%
  layout(yaxis = list(title = "Incidence per 100 000"),
         xaxis = list(title = "Province"),
         legend = list(orientation = "h", y = -0.2))
```

### Data Table

```{r}
#| label: tbl-prov-data
DT::datatable(
  sd,
  extensions = "Buttons",
  filter   = "top",
  rownames = FALSE,
  options  = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel", "pdf"),
    pageLength = 20,
    scrollX    = TRUE
  )
)
```

---

## Province Sub-pages

Individual province dashboards with district-level drilldowns:

::: {layout-ncol=3}
[Eastern Cape](ec/index.html){.btn .btn-outline-primary .btn-sm}
[Free State](fs/index.html){.btn .btn-outline-primary .btn-sm}
[Gauteng](gp/index.html){.btn .btn-outline-primary .btn-sm}
[KwaZulu-Natal](kzn/index.html){.btn .btn-outline-primary .btn-sm}
[Limpopo](lp/index.html){.btn .btn-outline-primary .btn-sm}
[Mpumalanga](mp/index.html){.btn .btn-outline-primary .btn-sm}
[Northern Cape](nc/index.html){.btn .btn-outline-primary .btn-sm}
[North West](nw/index.html){.btn .btn-outline-primary .btn-sm}
[Western Cape](wc/index.html){.btn .btn-outline-primary .btn-sm}
:::

---

::: {.callout-tip}
## Related Sections
- [By Condition](../by-condition/index.html) — condition-first view with national trends
- [Trending & Signals](../by-trending/index.html) — signal detection for emerging conditions
- [Burden of Disease](../burden-of-disease/index.html) — YLL, DALYs & age-standardised rates
- [Forecasts](../by-forecasts/index.html) — short-term predictive models
- [Explorer](../explorer/index.html) — custom data queries and downloads
- [Data Quality](../quality.html) — reporting completeness underpinning this data
:::
