---
title: "By Province"
subtitle: "Provincial surveillance organised by NMC notification category — Category 1 for immediate action, Category 2 for trend monitoring"
format:
  html:
    page-layout: full
    toc: true
    toc-depth: 3
---

::: {.callout-note collapse="true"}
## Development Roadmap — By Province

| Phase | Feature | Status |
|-------|---------|--------|
| **v0.1** | Monthly case counts by province × condition (interactive bar chart) | ✅ Done |
| **v0.1** | Incidence per 100 000 dot-plot, filterable by condition | ✅ Done |
| **v0.1** | Searchable & downloadable data table (CSV / Excel) | ✅ Done |
| **v0.1** | NMC category organisation (Cat 1–4 tabs) | ✅ Done |
| **v0.2** | Leaflet choropleth map — province-level counts | 🔲 Planned |
| **v0.2** | District drilldown table per province | 🔲 Planned |
| **v0.3** | Click-to-zoom district maps with Leaflet + lightbox | 🔲 Planned |
| **v0.3** | Time-series sparklines per province | 🔲 Planned |
| **v0.4** | Link to per-province sub-pages (EC, GP, …) | 🔲 Planned |
| **v0.5** | EpiNow2 Rt estimates per province fed from `agg_province.rds` | 🔲 Planned |

*Data pipeline:* `R/prepare_dashboard_data.r` → `data/processed/agg_province.rds`
:::

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(data.table)
  library(plotly)
  library(crosstalk)
  library(DT)
  library(lubridate)
  library(htmltools)
})

source("R/nmc_categories.r")

# ── Load processed data ──────────────────────────────────────────────────────
agg_prov <- readRDS("data/processed/agg_province.rds")
setDT(agg_prov)
agg_prov <- agg_prov[!is.na(prov_) & prov_ != ""]
agg_prov[, date := as.Date(date)]

# Assign NMC categories
agg_prov <- assign_nmc_category(agg_prov)

latest_date <- max(agg_prov$date)
month_start <- floor_date(latest_date, "month")

# ── Category colour palette ──────────────────────────────────────────────────
cat_colours <- c(
  "1" = "#d9534f", "2" = "#5bc0de", "3" = "#5cb85c",
  "4" = "#f0ad4e", "0" = "#999999"
)

# ── Monthly summary: province × condition × category ─────────────────────────
prov_month <- agg_prov[date >= month_start, .(
  cases          = sum(count, na.rm = TRUE),
  incidence_100k = round(sum(count, na.rm = TRUE) / first(pop) * 1e5, 2)
), by = .(Province = prov_, Condition = condition, nmc_category)]
prov_month[, Category := nmc_category_short(nmc_category)]

# ── Category totals per province ─────────────────────────────────────────────
cat_prov_totals <- prov_month[, .(
  total_cases = sum(cases),
  n_conditions = uniqueN(Condition[cases > 0])
), by = .(Province, nmc_category, Category)]
setorder(cat_prov_totals, nmc_category, -total_cases)
```

Provincial surveillance data organised by NMC notification category. **Category 1 conditions appear first** — they require immediate public health action. Category 2 supports trend monitoring for higher-volume routine conditions.

**Data through:** `r format(latest_date, "%d %B %Y")`

---

## Category 1 — Immediate Notification {#cat1}

::: {.callout-important}
## Requires immediate public health action
Category 1 conditions must be notified within **24 hours**. Includes cholera, measles, malaria, meningococcal disease, VHFs, and other nationally significant threats. **Use this section for rapid situational awareness across provinces.**
:::

```{r}
#| label: cat1-data
#| include: false
cat1_prov <- prov_month[nmc_category == 1]
sd_cat1   <- SharedData$new(cat1_prov)
```

### Filters

```{r}
#| label: cat1-filters
div(
  style = "display:flex; gap:2rem; flex-wrap:wrap; margin-bottom:1.5rem;",
  div(style = "min-width:260px;", filter_select("cat1cond", "Condition", sd_cat1, ~Condition)),
  div(style = "min-width:180px;", filter_slider("cat1cases", "Cases ≥", sd_cat1, ~cases, min = 0))
)
```

### Cases by Province {.tabset}

#### Stacked Bar Chart

```{r}
#| label: fig-cat1-bar
plot_ly(sd_cat1, x = ~Province, y = ~cases, color = ~Condition,
        type = "bar", hoverinfo = "text",
        text = ~paste0(Condition, "<br>", Province, ": ", cases, " cases")) %>%
  layout(barmode = "stack",
         xaxis = list(title = "Province", categoryorder = "total descending"),
         yaxis = list(title = "Cases"),
         legend = list(orientation = "h", y = -0.2),
         title = list(text = "Category 1 — Immediate Notification"))
```

#### Incidence Dot-plot

```{r}
#| label: fig-cat1-incidence
plot_ly(sd_cat1, x = ~Province, y = ~incidence_100k, color = ~Condition,
        type = "scatter", mode = "markers",
        marker = list(size = 10, opacity = 0.8),
        hoverinfo = "text",
        text = ~paste0(Condition, "<br>", Province,
                       ": ", incidence_100k, " per 100k")) %>%
  layout(yaxis = list(title = "Incidence per 100 000"),
         xaxis = list(title = "Province"),
         legend = list(orientation = "h", y = -0.2))
```

#### Data Table

```{r}
#| label: tbl-cat1-data
DT::datatable(
  sd_cat1,
  extensions = "Buttons",
  filter   = "top",
  rownames = FALSE,
  options  = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel", "pdf"),
    pageLength = 30,
    scrollX    = TRUE
  )
)
```

---

## Category 2 — Routine Notification (Trend Monitoring) {#cat2}

::: {.callout-note}
## Higher-volume conditions for quarterly trend reports
Category 2 conditions are notified within **7 days**. Includes TB (all forms), hepatitis A–E, schistosomiasis (bilharzia), congenital syphilis, poisonings, and maternal deaths. **Use this section for routine monitoring and quarterly reporting.**
:::

```{r}
#| label: cat2-data
#| include: false
cat2_prov <- prov_month[nmc_category == 2]
sd_cat2   <- SharedData$new(cat2_prov)
```

### Filters

```{r}
#| label: cat2-filters
div(
  style = "display:flex; gap:2rem; flex-wrap:wrap; margin-bottom:1.5rem;",
  div(style = "min-width:260px;", filter_select("cat2cond", "Condition", sd_cat2, ~Condition)),
  div(style = "min-width:180px;", filter_slider("cat2cases", "Cases ≥", sd_cat2, ~cases, min = 0))
)
```

### Cases by Province {.tabset}

#### Stacked Bar Chart

```{r}
#| label: fig-cat2-bar
plot_ly(sd_cat2, x = ~Province, y = ~cases, color = ~Condition,
        type = "bar", hoverinfo = "text",
        text = ~paste0(Condition, "<br>", Province, ": ", cases, " cases")) %>%
  layout(barmode = "stack",
         xaxis = list(title = "Province", categoryorder = "total descending"),
         yaxis = list(title = "Cases"),
         legend = list(orientation = "h", y = -0.2),
         title = list(text = "Category 2 — Routine Notification"))
```

#### Incidence Dot-plot

```{r}
#| label: fig-cat2-incidence
plot_ly(sd_cat2, x = ~Province, y = ~incidence_100k, color = ~Condition,
        type = "scatter", mode = "markers",
        marker = list(size = 10, opacity = 0.8),
        hoverinfo = "text",
        text = ~paste0(Condition, "<br>", Province,
                       ": ", incidence_100k, " per 100k")) %>%
  layout(yaxis = list(title = "Incidence per 100 000"),
         xaxis = list(title = "Province"),
         legend = list(orientation = "h", y = -0.2))
```

#### Data Table

```{r}
#| label: tbl-cat2-data
DT::datatable(
  sd_cat2,
  extensions = "Buttons",
  filter   = "top",
  rownames = FALSE,
  options  = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel", "pdf"),
    pageLength = 30,
    scrollX    = TRUE
  )
)
```

---

## Category 3 — Sentinel Surveillance {#cat3}

::: {.callout-tip collapse="true"}
## Sentinel site monitoring
Category 3 covers arboviral diseases (Chikungunya, Sindbis, West Nile, Zika, Dengue), non-typhoidal Salmonellosis, STEC, Shigellosis, and resistant Gonorrhoea.
:::

```{r}
#| label: cat3-data
#| include: false
cat3_prov <- prov_month[nmc_category == 3]
sd_cat3   <- SharedData$new(cat3_prov)
```

### Cases by Province {.tabset}

#### Stacked Bar Chart

```{r}
#| label: fig-cat3-bar
if (nrow(cat3_prov) > 0) {
  plot_ly(sd_cat3, x = ~Province, y = ~cases, color = ~Condition,
          type = "bar", hoverinfo = "text",
          text = ~paste0(Condition, "<br>", Province, ": ", cases)) %>%
    layout(barmode = "stack",
           xaxis = list(title = "Province", categoryorder = "total descending"),
           yaxis = list(title = "Cases"),
           legend = list(orientation = "h", y = -0.2),
           title = list(text = "Category 3 — Sentinel Surveillance"))
} else {
  htmltools::p("No Category 3 data this month.", style = "color:gray;")
}
```

#### Data Table

```{r}
#| label: tbl-cat3-data
if (nrow(cat3_prov) > 0) {
  DT::datatable(sd_cat3, extensions = "Buttons", filter = "top",
    rownames = FALSE,
    options = list(dom = "Bfrtip", buttons = c("csv", "excel"),
                   pageLength = 20, scrollX = TRUE))
} else {
  htmltools::p("No Category 3 data to display.", style = "color:gray;")
}
```

---

## Category 4 — AMR Organisms {#cat4}

::: {.callout-warning collapse="true"}
## Antimicrobial resistance monitoring
Category 4 covers CPE, VRE, GISA/hGISA, colistin-resistant *Pseudomonas*/*Acinetobacter*, and *C. difficile*.
:::

```{r}
#| label: cat4-data
#| include: false
cat4_prov <- prov_month[nmc_category == 4]
sd_cat4   <- SharedData$new(cat4_prov)
```

### Cases by Province {.tabset}

#### Stacked Bar Chart

```{r}
#| label: fig-cat4-bar
if (nrow(cat4_prov) > 0) {
  plot_ly(sd_cat4, x = ~Province, y = ~cases, color = ~Condition,
          type = "bar", hoverinfo = "text",
          text = ~paste0(Condition, "<br>", Province, ": ", cases)) %>%
    layout(barmode = "stack",
           xaxis = list(title = "Province", categoryorder = "total descending"),
           yaxis = list(title = "Cases"),
           legend = list(orientation = "h", y = -0.2),
           title = list(text = "Category 4 — AMR Organisms"))
} else {
  htmltools::p("No Category 4 data this month.", style = "color:gray;")
}
```

#### Data Table

```{r}
#| label: tbl-cat4-data
if (nrow(cat4_prov) > 0) {
  DT::datatable(sd_cat4, extensions = "Buttons", filter = "top",
    rownames = FALSE,
    options = list(dom = "Bfrtip", buttons = c("csv", "excel"),
                   pageLength = 20, scrollX = TRUE))
} else {
  htmltools::p("No Category 4 data to display.", style = "color:gray;")
}
```

---

## All Categories — Summary Table

```{r}
#| label: tbl-all-prov
all_prov <- prov_month[order(nmc_category, Province, -cases)]
DT::datatable(
  all_prov[, .(Category, Province, Condition, Cases = cases,
               `Incidence /100k` = incidence_100k)],
  extensions = "Buttons",
  filter = "top",
  rownames = FALSE,
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel", "pdf"),
    pageLength = 30,
    scrollX = TRUE
  ),
  caption = paste0("All conditions by NMC category and province — ", format(month_start, "%B %Y"))
)
```

---

## Province Sub-pages

Individual province dashboards with district-level drilldowns organised by NMC category:

::: {layout-ncol=3}
[Eastern Cape](ec/index.html){.btn .btn-outline-primary .btn-sm}
[Free State](fs/index.html){.btn .btn-outline-primary .btn-sm}
[Gauteng](gp/index.html){.btn .btn-outline-primary .btn-sm}
[KwaZulu-Natal](kzn/index.html){.btn .btn-outline-primary .btn-sm}
[Limpopo](lp/index.html){.btn .btn-outline-primary .btn-sm}
[Mpumalanga](mp/index.html){.btn .btn-outline-primary .btn-sm}
[Northern Cape](nc/index.html){.btn .btn-outline-primary .btn-sm}
[North West](nw/index.html){.btn .btn-outline-primary .btn-sm}
[Western Cape](wc/index.html){.btn .btn-outline-primary .btn-sm}
:::

---

::: {.callout-tip}
## Related Sections
- [By Condition](../by-condition/index.html) — condition-first view with national trends
- [Trending & Signals](../by-trending/index.html) — signal detection for emerging conditions
- [Burden of Disease](../burden-of-disease/index.html) — YLL, DALYs & age-standardised rates
- [Forecasts](../by-forecasts/index.html) — short-term predictive models
- [Explorer](../explorer/index.html) — custom data queries and downloads
- [Data Quality](../quality.html) — reporting completeness underpinning this data
:::
