---
title: "Vaccine-Preventable Diseases"
subtitle: "Epidemiological Description of VPD Notifications · NMC Surveillance, South Africa"
date: today
sidebar: special-reports
execute:
  echo: false
  warning: false
  message: false
  error: false
  cache: false
---

::: {.callout-important appearance="minimal"}
**Priority surveillance** — Vaccine-preventable diseases (VPDs) are critical indicators of immunisation programme performance. This report describes the epidemiology of notifiable VPDs based on NMC surveillance data, 2020–present.
:::

```{r setup}
#| label: setup
#| include: false
#| cache: false

pacman::p_load(
  ggplot2, gtsummary, tidyverse, magrittr,
  conflicted, scales, data.table, lubridate,
  flextable, gt, ggrepel, zoo,
  janitor
)

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::lag)
conflicted::conflicts_prefer(tidyr::extract)
conflicted::conflicts_prefer(dplyr::first)
conflicted::conflicts_prefer(lubridate::year)
conflicted::conflicts_prefer(lubridate::month)
conflicted::conflicts_prefer(lubridate::quarter)

library(NMCleaner)

# ── Load master dataset ───────────────────────────────────────────────────────
load("~/Desktop/SAFETP/CLA/NMC_database/master/new_master.rda")

# ── Define VPD conditions ─────────────────────────────────────────────────────
# Core EPI-targeted VPDs
VPD_EPI <- c(

  "Measles",
  "Rubella",
  "Congenital rubella syndrome",
  "Diphtheria",
  "Pertussis",
  "Tetanus",
  "Poliomyelitis",
  "Acute flaccid paralysis",
  "Haemophilus influenzae type B"
)

# Other vaccine-preventable conditions
VPD_OTHER <- c(
  "Hepatitis A",
  "Hepatitis B",
  "Rabies",
  "Yellow fever"
)

VPD_CONDITIONS <- c(VPD_EPI, VPD_OTHER)

# Categorisation for analysis
VPD_CATEGORY <- c(
  "Measles"                      = "Measles/Rubella",
  "Rubella"                      = "Measles/Rubella",
  "Congenital rubella syndrome"  = "Measles/Rubella",
  "Diphtheria"                   = "DTP diseases",
  "Pertussis"                    = "DTP diseases",
  "Tetanus"                      = "DTP diseases",
  "Poliomyelitis"                = "Polio surveillance",
  "Acute flaccid paralysis"      = "Polio surveillance",
  "Haemophilus influenzae type B" = "Hib",
  "Hepatitis A"                  = "Hepatitis",
  "Hepatitis B"                  = "Hepatitis",
  "Rabies"                       = "Rabies",
  "Yellow fever"                 = "Yellow fever"
)

# ── Filter to VPD conditions only ─────────────────────────────────────────────
new_master %>%
  filter(
    condition %in% VPD_CONDITIONS,
    as.integer(Year_notification) %in% 2020:year(Sys.Date()),
    !is.na(prov_)
  ) %>%
  mutate(
    Year_notification = as.integer(Year_notification),
    patient_vital_status = if_else(
      patient_vital_status %like% "Unknown", "Alive", patient_vital_status
    ),
    Age_years = as.numeric(Age_years),
    age_group = cut(
      Age_years,
      breaks = c(-Inf, 1, 5, 10, 15, 20, 30, 40, 50, 65, Inf),
      labels = c("<1", "1–4", "5–9", "10–14", "15–19",
                 "20–29", "30–39", "40–49", "50–64", "65+"),
      right  = FALSE
    ),
    # Pediatric age groups for EPI analysis
    age_group_peds = cut(
      Age_years,
      breaks = c(-Inf, 1, 5, 10, 15, Inf),
      labels = c("<1 year", "1–4 years", "5–9 years", "10–14 years", "15+ years"),
      right  = FALSE
    ),
    sex = case_when(
      patient_gender %in% c("Male", "M")   ~ "Male",
      patient_gender %in% c("Female", "F") ~ "Female",
      TRUE                                  ~ NA_character_
    ),
    deceased = patient_vital_status %in% "Deceased",
    vpd_category = recode(condition, !!!VPD_CATEGORY),
    is_epi_vpd   = condition %in% VPD_EPI
  ) -> vpd

# ── Shared theme ──────────────────────────────────────────────────────────────
theme_nmc <- function(...) {
  theme_classic() +
    theme(
      text              = element_text(family = "Century Gothic", size = 10, color = "#333333"),
      panel.background  = element_rect(fill = "#F8F8F8"),
      plot.background   = element_rect(fill = "#F8F8F8"),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      axis.ticks        = element_line(linewidth = 0.15, color = "grey40"),
      legend.background = element_rect(fill = "#F8F8F8", linewidth = 0.25,
                                       linetype = "solid", colour = "grey50"),
      legend.key        = element_blank(),
      strip.background  = element_rect(color = NA, fill = NA),
      ...
    )
}

# VPD colour palette
vpd_cols <- c(
  "Measles/Rubella"    = "#E53935",
  "DTP diseases"       = "#1E88E5",
  "Polio surveillance" = "#43A047",
  "Hib"                = "#8E24AA",
  "Hepatitis"          = "#FB8C00",
  "Rabies"             = "#546E7A",
  "Yellow fever"       = "#FDD835"
)

condition_cols <- c(
  "Measles"                       = "#E53935",
  "Rubella"                       = "#F48FB1",
  "Congenital rubella syndrome"   = "#AD1457",
  "Diphtheria"                    = "#1565C0",
  "Pertussis"                     = "#42A5F5",
  "Tetanus"                       = "#0D47A1",
  "Poliomyelitis"                 = "#2E7D32",
  "Acute flaccid paralysis"       = "#81C784",
  "Haemophilus influenzae type B" = "#8E24AA",
  "Hepatitis A"                   = "#FB8C00",
  "Hepatitis B"                   = "#EF6C00",
  "Rabies"                        = "#546E7A",
  "Yellow fever"                  = "#FDD835"
)
```

::: {.callout-note appearance="minimal" collapse="true"}
## Report metadata
| | |
|---|---|
| **Conditions** | `r paste(VPD_CONDITIONS, collapse = "; ")` |
| **Years** | 2020 – `r lubridate::year(Sys.Date())` |
| **Data source** | NMC Notifiable Medical Conditions surveillance database |
| **Generated** | `r format(Sys.Date(), "%d %B %Y")` |
:::

---

# Situation Update for Policy Makers {#sec-executive}

```{r exec-setup}
#| label: exec-setup
#| include: false

# ── Current year and comparison calculations ──────────────────────────────────
current_year <- year(Sys.Date())
prev_year    <- current_year - 1
current_month <- month(Sys.Date())

# Year-to-date for current and previous year
ytd_current <- vpd %>%
  filter(Year_notification == current_year,
         month(as_date(notification_date)) <= current_month)
ytd_prev <- vpd %>%
  filter(Year_notification == prev_year,
         month(as_date(notification_date)) <= current_month)

n_ytd_current   <- nrow(ytd_current)
n_ytd_prev      <- nrow(ytd_prev)
deaths_ytd_curr <- sum(ytd_current$deceased, na.rm = TRUE)
deaths_ytd_prev <- sum(ytd_prev$deceased, na.rm = TRUE)

pct_change_n      <- round((n_ytd_current - n_ytd_prev) / max(n_ytd_prev, 1) * 100)
pct_change_deaths <- round((deaths_ytd_curr - deaths_ytd_prev) / max(deaths_ytd_prev, 1) * 100)

arrow_n      <- if (pct_change_n > 0) "↑" else if (pct_change_n < 0) "↓" else "→"
arrow_deaths <- if (pct_change_deaths > 0) "↑" else if (pct_change_deaths < 0) "↓" else "→"
col_n        <- if (pct_change_n > 10) "#C62828" else if (pct_change_n < -10) "#2E7D32" else "#F57C00"
col_deaths   <- if (pct_change_deaths > 10) "#C62828" else if (pct_change_deaths < -10) "#2E7D32" else "#F57C00"

# Key condition metrics
measles_ytd_curr  <- sum(ytd_current$condition == "Measles")
measles_ytd_prev  <- sum(ytd_prev$condition == "Measles")
pertussis_ytd_curr <- sum(ytd_current$condition == "Pertussis")
pertussis_ytd_prev <- sum(ytd_prev$condition == "Pertussis")
afp_ytd_curr      <- sum(ytd_current$condition == "Acute flaccid paralysis")
afp_ytd_prev      <- sum(ytd_prev$condition == "Acute flaccid paralysis")

pct_change_measles  <- round((measles_ytd_curr - measles_ytd_prev) / max(measles_ytd_prev, 1) * 100)
pct_change_pertussis <- round((pertussis_ytd_curr - pertussis_ytd_prev) / max(pertussis_ytd_prev, 1) * 100)
arrow_measles    <- if (pct_change_measles > 0) "↑" else if (pct_change_measles < 0) "↓" else "→"
arrow_pertussis  <- if (pct_change_pertussis > 0) "↑" else if (pct_change_pertussis < 0) "↓" else "→"

# Under 5 burden
under5_ytd_curr <- ytd_current %>% filter(Age_years < 5)
under5_ytd_prev <- ytd_prev %>% filter(Age_years < 5)
pct_under5_curr <- round(nrow(under5_ytd_curr) / max(n_ytd_current, 1) * 100, 1)

# Top 3 provinces
top3_prov <- ytd_current %>%
  count(prov_, sort = TRUE) %>%
  slice(1:3) %>%
  pull(prov_) %>%
  paste(collapse = ", ")
```

::: {.panel-tabset}

## Key Metrics

```{r exec-metrics}
#| label: tbl-exec-metrics
#| tbl-cap: !expr 'paste0("Year-to-date comparison: Jan–", format(Sys.Date(), "%b %Y"), " vs same period ", prev_year)'

tibble(
  Indicator = c(
    "All VPD notifications (YTD)",
    "Deaths (YTD)",
    "Measles cases",
    "Pertussis cases",
    "AFP cases (polio surveillance)",
    "Children <5 years (%)",
    "Top 3 provinces"
  ),
  `Current Year` = c(
    scales::comma(n_ytd_current),
    scales::comma(deaths_ytd_curr),
    scales::comma(measles_ytd_curr),
    scales::comma(pertussis_ytd_curr),
    scales::comma(afp_ytd_curr),
    paste0(pct_under5_curr, "%"),
    top3_prov
  ),
  `Same Period Last Year` = c(
    scales::comma(n_ytd_prev),
    scales::comma(deaths_ytd_prev),
    scales::comma(measles_ytd_prev),
    scales::comma(pertussis_ytd_prev),
    scales::comma(afp_ytd_prev),
    "—",
    "—"
  ),
  `Change` = c(
    paste0(arrow_n, " ", ifelse(pct_change_n >= 0, "+", ""), pct_change_n, "%"),
    paste0(arrow_deaths, " ", ifelse(pct_change_deaths >= 0, "+", ""), pct_change_deaths, "%"),
    paste0(arrow_measles, " ", ifelse(pct_change_measles >= 0, "+", ""), pct_change_measles, "%"),
    paste0(arrow_pertussis, " ", ifelse(pct_change_pertussis >= 0, "+", ""), pct_change_pertussis, "%"),
    "—",
    "—",
    "—"
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = 1:2, j = 1) %>%
  color(i = 1, j = 4, color = col_n) %>%
  color(i = 2, j = 4, color = col_deaths) %>%
  bg(i = 3:5, bg = "#FFF3E0") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 11, part = "all") %>%
  align(j = 2:4, align = "center", part = "all") %>%
  add_footer_lines(paste0("Data extracted: ", format(Sys.Date(), "%d %B %Y"),
                          ". YTD = Year-to-date (Jan 1 – ", format(Sys.Date(), "%d %b"), ")."))
```

## 12-Month Trend by Category

```{r exec-trend}
#| label: fig-exec-trend
#| fig-cap: "Monthly VPD notification trend by disease category — last 12 months."
#| fig-height: 5

# Last 12 complete months
end_date   <- floor_date(Sys.Date(), "month") - days(1)
start_date <- floor_date(end_date - months(11), "month")

vpd %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  filter(month_date >= start_date, month_date <= end_date) %>%
  count(month_date, vpd_category) %>%
  complete(month_date, vpd_category, fill = list(n = 0)) -> trend12_df

ggplot(trend12_df, aes(x = month_date, y = n, fill = vpd_category)) +
  geom_col(width = 22) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "1 month", expand = expansion(add = c(10, 10))) +
  scale_y_continuous(name = "Notifications", expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = vpd_cols, name = "Category") +
  labs(x = NULL) +
  theme_nmc() +
  theme(axis.text.x = element_text(size = 8), legend.position = "bottom",
        legend.key.size = unit(0.4, "cm"))
```

## Provincial Burden — Measles

```{r exec-prov-measles}
#| label: fig-exec-prov-measles
#| fig-cap: "Measles notifications by province — quarterly trend, last 2 years. Zero cases shaded grey."
#| fig-height: 5

# Get provincial population
prov_pop <- NMCleaner::pop %>%
  filter(as.integer(Year) == year(Sys.Date())) %>%
  group_by(prov) %>%
  summarise(population = sum(Population), .groups = "drop")

# Measles by province and quarter
vpd %>%
  filter(condition == "Measles", !is.na(notification_date)) %>%
  mutate(
    nd  = as_date(notification_date),
    qtr = paste0(year(nd), " Q", quarter(nd))
  ) %>%
  filter(nd >= (Sys.Date() - years(2))) %>%
  group_by(prov_, qtr) %>%
  summarise(n = n(), .groups = "drop") %>%
  complete(prov_, qtr, fill = list(n = 0)) -> measles_qtr

# Order quarters
qtr_order <- measles_qtr %>% distinct(qtr) %>% arrange(qtr) %>% pull(qtr)
measles_qtr$qtr <- factor(measles_qtr$qtr, levels = qtr_order)

ggplot(measles_qtr, aes(x = qtr, y = prov_, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = n), size = 3, colour = "black") +
  scale_fill_gradient(low = "#FFEBEE", high = "#B71C1C", name = "Cases",
                      na.value = "#E0E0E0") +
  labs(x = "Quarter", y = NULL) +
  theme_nmc() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "right",
    panel.grid = element_blank()
  )
```

:::

---

# Summary Statistics {#sec-summary}

```{r summary-stats}
#| label: tbl-summary
#| tbl-cap: "Summary of VPD notifications by condition, South Africa, 2020–present."

vpd %>%
  group_by(condition) %>%
  summarise(
    n        = n(),
    deaths   = sum(deceased, na.rm = TRUE),
    cfr      = round(deaths / n * 100, 1),
    under5   = sum(Age_years < 5, na.rm = TRUE),
    pct_u5   = round(under5 / n * 100, 1),
    .groups  = "drop"
  ) %>%
  arrange(desc(n)) %>%
  mutate(
    Condition  = condition,
    `Cases`    = scales::comma(n),
    `Deaths`   = deaths,
    `CFR (%)`  = cfr,
    `<5 yrs`   = scales::comma(under5),
    `% <5 yrs` = pct_u5
  ) %>%
  select(Condition, Cases, Deaths, `CFR (%)`, `<5 yrs`, `% <5 yrs`) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 10, part = "all") %>%
  add_footer_lines("CFR = Case fatality ratio. <5 yrs = children under 5 years of age.")
```

---

# Temporal Trends {#sec-trends}

## Annual notifications by condition

```{r annual-trend}
#| label: fig-annual-trend
#| fig-cap: "Annual VPD notifications by condition, South Africa, 2020–present."
#| fig-height: 6

vpd %>%
  count(Year_notification, condition) %>%
  complete(Year_notification, condition, fill = list(n = 0)) -> annual_cond

# Top 8 conditions by total volume for cleaner display
top_conds <- vpd %>% count(condition, sort = TRUE) %>% slice(1:8) %>% pull(condition)

annual_cond %>%
  filter(condition %in% top_conds) %>%
  ggplot(aes(x = factor(Year_notification), y = n, fill = condition)) +
  geom_col(position = "dodge", width = 0.8) +
  scale_fill_manual(values = condition_cols, name = "Condition") +
  labs(x = "Year", y = "Notifications (n)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## Monthly epicurve — key conditions

```{r epicurve-monthly}
#| label: fig-epicurve-monthly
#| fig-cap: "Monthly notification counts for key VPDs (Measles, Pertussis, AFP), South Africa, 2020–present."
#| fig-height: 5

key_conds <- c("Measles", "Pertussis", "Acute flaccid paralysis")

vpd %>%
  filter(condition %in% key_conds, !is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  count(month_date, condition) %>%
  complete(month_date, condition, fill = list(n = 0)) -> epicurve_df

ggplot(epicurve_df, aes(x = month_date, y = n, colour = condition)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months") +
  scale_y_continuous(name = "Notifications (n)") +
  scale_colour_manual(values = condition_cols, name = "Condition") +
  labs(x = "Date") +
  facet_wrap(~condition, ncol = 1, scales = "free_y") +
  theme_nmc() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

## Seasonal pattern — Measles

```{r seasonal-measles}
#| label: fig-seasonal-measles
#| fig-cap: "Epiweek × year heatmap of Measles notifications, South Africa."
#| fig-height: 4.5

vpd %>%
  filter(condition == "Measles", !is.na(notification_date)) %>%
  mutate(
    nd      = as_date(notification_date),
    yr      = lubridate::year(nd),
    epiweek = lubridate::isoweek(nd)
  ) %>%
  count(year = yr, epiweek) %>%
  complete(year = 2020:year(Sys.Date()), epiweek = 1:52, fill = list(n = 0)) -> heat_df

ggplot(heat_df, aes(x = epiweek,
                    y = factor(year, levels = rev(sort(unique(year)))),
                    fill = n)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "white", mid = "#FFCDD2", high = "#B71C1C",
    midpoint = max(1, median(heat_df$n[heat_df$n > 0], na.rm = TRUE)),
    name = "Cases"
  ) +
  scale_x_continuous(breaks = c(1, seq(10, 52, 10)), expand = c(0, 0)) +
  labs(x = "Epiweek", y = "Year", title = "Measles") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.width = unit(1.5, "cm"))
```

---

# Age & Sex Demographics {#sec-demographics}

## Age distribution by condition

```{r age-by-condition}
#| label: fig-age-by-condition
#| fig-cap: "Age distribution of VPD notifications by condition (top 6 by volume), South Africa, 2020–present."
#| fig-height: 6

top6_conds <- vpd %>% count(condition, sort = TRUE) %>% slice(1:6) %>% pull(condition)

vpd %>%
  filter(condition %in% top6_conds, !is.na(age_group_peds)) %>%
  count(condition, age_group_peds) %>%
  group_by(condition) %>%
  mutate(pct = n / sum(n) * 100) -> age_cond_df

ggplot(age_cond_df, aes(x = age_group_peds, y = pct, fill = condition)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = condition_cols, name = "Condition") +
  labs(x = "Age group", y = "% within condition") +
  facet_wrap(~condition, scales = "free_y") +
  theme_nmc() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "none", strip.text = element_text(face = "bold", size = 9))
```

## Under-5 burden by condition

```{r under5-burden}
#| label: fig-under5-burden
#| fig-cap: "Proportion of VPD notifications in children under 5 years by condition, South Africa, 2020–present."
#| fig-height: 5

vpd %>%
  group_by(condition) %>%
  summarise(
    n_total  = n(),
    n_under5 = sum(Age_years < 5, na.rm = TRUE),
    pct_u5   = n_under5 / n_total * 100,
    .groups  = "drop"
  ) %>%
  filter(n_total >= 10) %>%
  mutate(condition = fct_reorder(condition, pct_u5)) -> u5_df

ggplot(u5_df, aes(x = condition, y = pct_u5, fill = pct_u5)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(round(pct_u5, 0), "%")), hjust = -0.2, size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "#BBDEFB", high = "#1565C0", guide = "none") +
  scale_y_continuous(limits = c(0, 105)) +
  labs(x = NULL, y = "% of cases in children <5 years") +
  theme_nmc()
```

## Population pyramid — all VPDs

```{r age-sex-pyramid}
#| label: fig-age-sex-pyramid
#| fig-cap: "Age–sex population pyramid of all VPD notifications, South Africa, 2020–present."
#| fig-height: 5

vpd %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  count(age_group, sex) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyramid_df

max_n <- max(abs(pyramid_df$n))
step  <- ifelse(max_n > 500, 200, ifelse(max_n > 200, 50, 25))
br    <- seq(0, ceiling(max_n / step) * step, step)

ggplot(pyramid_df, aes(x = age_group, y = n_plot, fill = sex)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  geom_hline(yintercept = 0, colour = "white", linewidth = 0.5) +
  labs(x = "Age group (years)", y = "Number of notifications") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Case fatality by condition

```{r cfr-condition}
#| label: fig-cfr-condition
#| fig-cap: "Case fatality ratio (%) by VPD condition (conditions with ≥10 cases), South Africa, 2020–present."
#| fig-height: 5

vpd %>%
  group_by(condition) %>%
  summarise(
    n      = n(),
    deaths = sum(deceased, na.rm = TRUE),
    cfr    = deaths / n * 100,
    .groups = "drop"
  ) %>%
  filter(n >= 10) %>%
  mutate(condition = fct_reorder(condition, cfr)) -> cfr_cond

ggplot(cfr_cond, aes(x = condition, y = cfr, fill = cfr)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(round(cfr, 1), "% (", deaths, "/", n, ")")),
            hjust = -0.05, size = 2.8) +
  coord_flip() +
  scale_fill_gradient(low = "#FFF9C4", high = "#C62828", guide = "none") +
  scale_y_continuous(limits = c(0, max(cfr_cond$cfr, na.rm = TRUE) * 1.4)) +
  labs(x = NULL, y = "Case fatality ratio (%)") +
  theme_nmc()
```

---

# Geographical Distribution {#sec-geo}

## Provincial incidence — Measles

```{r prov-incidence-measles}
#| label: fig-prov-incidence-measles
#| fig-cap: "Measles incidence per 100,000 population by province and year, South Africa."
#| fig-height: 5

NMCleaner::pop %>%
  mutate(Year = as.integer(Year)) %>%
  group_by(prov, Year) %>%
  summarise(population = sum(Population), .groups = "drop") -> pop_prov

vpd %>%
  filter(condition == "Measles") %>%
  group_by(prov_, year = Year_notification) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(pop_prov, by = c("prov_" = "prov", "year" = "Year")) %>%
  mutate(incidence = n / population * 1e5) -> measles_inc

ggplot(measles_inc, aes(x = year, y = incidence, colour = prov_, group = prov_)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  scale_x_continuous(breaks = unique(measles_inc$year)) +
  labs(x = "Year", y = "Incidence per 100,000 population",
       title = "Measles incidence by province") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## Top districts by VPD burden

```{r top-districts}
#| label: tbl-top-districts
#| tbl-cap: "Top 20 districts by total VPD notification count, South Africa, 2020–present."

NMCleaner::pop %>%
  filter(as.integer(Year) %in% 2020:year(Sys.Date())) %>%
  group_by(district_standard) %>%
  summarise(pop_total = sum(Population), .groups = "drop") -> pop_dist

vpd %>%
  filter(!district %in% c("Unknown", "Isazi Import", "Not Applicable"),
         !is.na(district)) %>%
  mutate_district_name("district") %>%
  group_by(Province = prov_, district_standard) %>%
  summarise(
    Notifications = n(),
    Deaths        = sum(deceased, na.rm = TRUE),
    CFR           = round(Deaths / Notifications * 100, 1),
    n_measles     = sum(condition == "Measles"),
    n_pertussis   = sum(condition == "Pertussis"),
    .groups = "drop"
  ) %>%
  left_join(pop_dist, by = "district_standard") %>%
  mutate(
    Incidence = round(Notifications / pop_total * 1e5, 1),
    pop_total = round(pop_total / 1e3, 1)
  ) %>%
  arrange(desc(Notifications)) %>%
  slice(1:20) %>%
  rename(District = district_standard, Measles = n_measles, Pertussis = n_pertussis,
         `Incidence per 100k` = Incidence, `Pop (k)` = pop_total) %>%
  select(Province, District, Notifications, Deaths, `CFR (%)` = CFR,
         Measles, Pertussis, `Incidence per 100k`, `Pop (k)`) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 9, part = "all")
```

---

# Polio Surveillance (AFP) {#sec-polio}

::: {.callout-note}
## AFP Surveillance Context

**Acute flaccid paralysis (AFP)** surveillance is the global standard for detecting poliovirus circulation. South Africa maintains high AFP surveillance sensitivity as part of polio eradication efforts. The target is **≥2 AFP cases per 100,000 children under 15 years**.
:::

```{r afp-analysis}
#| label: fig-afp-analysis
#| fig-cap: "AFP surveillance performance — annual cases and rate per 100,000 children <15 years, South Africa."
#| fig-height: 4.5

# Under-15 population
pop_u15 <- NMCleaner::pop %>%
  filter(as.integer(Year) %in% 2020:year(Sys.Date())) %>%
  filter(Age < 15) %>%
  group_by(Year = as.integer(Year)) %>%
  summarise(pop_u15 = sum(Population), .groups = "drop")

vpd %>%
  filter(condition == "Acute flaccid paralysis") %>%
  group_by(year = Year_notification) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(pop_u15, by = c("year" = "Year")) %>%
  mutate(rate = n / pop_u15 * 1e5) -> afp_rate

ggplot(afp_rate, aes(x = factor(year))) +
  geom_col(aes(y = n), fill = "#43A047", alpha = 0.7, width = 0.6) +
  geom_hline(yintercept = 2, linetype = "dashed", colour = "#C62828", linewidth = 0.8) +
  geom_text(aes(y = n, label = paste0(n, "\n(", round(rate, 1), ")")),
            vjust = -0.2, size = 3) +
  annotate("text", x = 0.6, y = 2.2, label = "Target ≥2/100k", hjust = 0,
           colour = "#C62828", size = 3, fontface = "italic") +
  scale_y_continuous(name = "AFP cases", expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Year", caption = "Numbers in bars: cases (rate per 100,000 <15 yrs)") +
  theme_nmc()
```

```{r afp-by-prov}
#| label: fig-afp-by-prov
#| fig-cap: "AFP cases by province and year, South Africa, 2020–present."
#| fig-height: 5

vpd %>%
  filter(condition == "Acute flaccid paralysis") %>%
  count(Year_notification, prov_) %>%
  complete(Year_notification, prov_, fill = list(n = 0)) -> afp_prov

ggplot(afp_prov, aes(x = factor(Year_notification), y = prov_, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = n), size = 3) +
  scale_fill_gradient(low = "#E8F5E9", high = "#1B5E20", name = "Cases") +
  labs(x = "Year", y = NULL, title = "AFP cases by province") +
  theme_nmc() +
  theme(legend.position = "right")
```

---

# Mortality Profile {#sec-mortality}

## Deaths by condition and year

```{r deaths-condition}
#| label: fig-deaths-condition
#| fig-cap: "Annual VPD deaths by condition, South Africa."
#| fig-height: 5

vpd %>%
  filter(deceased) %>%
  count(year = Year_notification, condition) -> deaths_cond

ggplot(deaths_cond, aes(x = factor(year), y = n, fill = condition)) +
  geom_col(position = "stack", width = 0.75) +
  scale_fill_manual(values = condition_cols, name = "Condition") +
  labs(x = "Year", y = "Deaths (n)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

---

# Reporting & Facility Analysis {#sec-facilities}

## Unique reporting facilities per year

```{r facilities-trend}
#| label: fig-facilities
#| fig-cap: "Number of unique health facilities reporting at least one VPD notification per year, South Africa."
#| fig-height: 4

vpd %>%
  filter(!is.na(facility)) %>%
  group_by(Year_notification) %>%
  summarise(n_facilities = n_distinct(facility), .groups = "drop") -> fac_df

ggplot(fac_df, aes(x = factor(Year_notification))) +
  geom_col(aes(y = n_facilities, fill = "Unique facilities"), width = 0.6, alpha = 0.85) +
  geom_text(aes(y = n_facilities, label = n_facilities), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Unique facilities" = "#4CAF50"), name = NULL) +
  labs(x = "Year", y = "Unique reporting facilities (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

---

::: {.callout-note}
## Interpretation Notes

**EPI-targeted VPDs:**

- **Measles/Rubella**: Core EPI vaccines; rising cases may indicate immunisation gaps
- **DTP diseases** (Diphtheria, Pertussis, Tetanus): Rare in well-immunised populations
- **Polio surveillance**: AFP cases should be investigated for poliovirus; target ≥2/100k <15 yrs
- **Hib**: Haemophilus influenzae type B; vaccine introduced in 1999

**Surveillance implications:**

- Clusters of measles cases warrant immunisation response
- Any diphtheria or poliomyelitis case is a public health emergency
- High CFR in certain VPDs may reflect late presentation or case ascertainment bias

See [Data Quality](../../quality.html) for reporting completeness and [Methods](../../methods.html) for full definitions.
:::

---

*Report generated: `r format(Sys.Date(), "%d %B %Y")` | NMC Surveillance Unit*
