---
title: "Trending Conditions"
subtitle: "Signal detection, emerging hotspots & alert log"
format:
  html:
    toc: true
---

::: {.callout-note collapse="true"}
## Development Roadmap — Trending / Signal Detection

| Phase | Feature | Status |
|-------|---------|--------|
| **v0.1** | Recent surge table — conditions with biggest % increase (last 14 days vs prior 14) | ✅ Done |
| **v0.1** | Interactive trend sparklines for flagged conditions | ✅ Done |
| **v0.2** | EARS-C2 + sliding-window CUSUM signal detection | ✅ Done |
| **v0.2** | Province × condition signal heatmap (drill-down) | ✅ Done |
| **v0.2** | Persistent alert log (`data/processed/alerts.rds`) | ✅ Done |
| **v0.3** | Analyst notes & manual triage layer | 🔲 Planned |
| **v0.3** | Automated alert email / webhook on threshold breach | 🔲 Planned |
| **v0.4** | Auto-trigger EpiNow2 forecasts for flagged conditions | 🔲 Planned |

*Data pipeline:* `R/prepare_dashboard_data.r` → `data/processed/agg_national.rds`
*Algorithms:* `R/signal_detection.r` (`detect_signals_batch`, `cusum_sliding`, `build_alert_log`)
:::

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(data.table)
  library(plotly)
  library(DT)
  library(lubridate)
  library(ggplot2)
  library(htmltools)
})

source("R/_load_all.r")

agg_nat  <- readRDS("data/processed/agg_national.rds")
agg_prov <- readRDS("data/processed/agg_province.rds")
setDT(agg_nat); setDT(agg_prov)
agg_nat[, date := as.Date(date)]
agg_prov[, date := as.Date(date)]
agg_prov <- agg_prov[!is.na(prov_) & prov_ != ""]

# Column harmonisation: signal_detection.r expects `n` (daily count)
agg_nat[, n := count]
agg_prov[, n := count]

latest_date <- max(agg_nat$date)

# Reporting-delay window — last 14 days are provisional and biased low.
TRUNC_DAYS <- 14L

agg_nat_eval  <- agg_nat[date <= (latest_date - TRUNC_DAYS)]
agg_prov_eval <- agg_prov[date <= (latest_date - TRUNC_DAYS)]
```

::: {.data-through-banner}
**Data through:** `r format(latest_date, "%d %B %Y")` &nbsp;·&nbsp;
**Signal-detection cutoff:** `r format(latest_date - TRUNC_DAYS, "%d %B %Y")`
<span class="muted">— the last `r TRUNC_DAYS` days are excluded from algorithms (reporting incomplete)</span>
:::

---

## National Signal Log — Last 28 Days

EARS-C2 (rolling 14-day z-score, threshold = 3) and one-sided CUSUM
(slack k = 0.5, threshold = 5) applied to each condition's daily series.
Signals are de-duplicated to one row per (condition × method × episode) and
persisted to `data/processed/alerts.rds`.

::: {.callout-note collapse="true"}
## Stabilising z-scores for rare conditions

The baseline SD used in the EARS-C2 z-score is replaced by
$\tilde s = \max\!\bigl(s,\,\sqrt{\max(\bar n,\,1)}\bigr)$ — i.e. at least the
Poisson-implied SD given the baseline mean (`sd_floor = "poisson"`). EARS
signals are additionally suppressed when the baseline mean is below
0.5 cases/day (`min_baseline_mean = 0.5`); these rows are still computed but
shown only via CUSUM, which is more stable on sparse counts. See
[Methods → Signal Detection](../methods.html#signal-detection) for the full
rationale.
:::

```{r}
#| label: signal-detect-national
sig_nat <- detect_signals_batch(
  agg_nat_eval[, .(condition, date, n)],
  group_cols        = "condition",
  baseline          = 14L,
  z_thresh          = 3,
  k                 = 0.5,
  cusum_thresh      = 5,
  sd_floor          = "poisson",
  min_baseline_mean = 0.5
)

alert_nat <- build_alert_log(sig_nat, group_cols = "condition", recent_days = 28L)

invisible(persist_alerts(alert_nat,
                         group_cols = "condition",
                         path       = "data/processed/alerts.rds"))
```

```{r}
#| label: tbl-alerts
if (nrow(alert_nat) == 0L) {
  htmltools::tags$p(
    "No EARS or CUSUM signals in the last 28 days.",
    style = "color:#475569;font-style:italic;"
  )
} else {
  display <- alert_nat[, .(
    Condition      = condition,
    Method         = method,
    `First signal` = format(first_signal, "%d %b %Y"),
    `Last signal`  = format(last_signal,  "%d %b %Y"),
    `Signal days`  = n_signal_days,
    `Peak score`   = round(peak_value, 2),
    `Latest count` = latest_n
  )]
  DT::datatable(display, rownames = FALSE,
                options = list(pageLength = 15, dom = "ftp",
                               order = list(list(3, "desc"))))
}
```

---

## Province × Condition Signal Heatmap

Conditions on the y-axis, provinces on the x-axis; cell colour indicates the
peak CUSUM score within the last 28 days. Empty cells = no signal.

```{r}
#| label: signal-prov
sig_prov <- detect_signals_batch(
  agg_prov_eval[, .(condition, prov_, date, n)],
  group_cols        = c("condition", "prov_"),
  baseline          = 14L,
  z_thresh          = 3,
  k                 = 0.5,
  cusum_thresh      = 5,
  sd_floor          = "poisson",
  min_baseline_mean = 0.5
)

alert_prov <- build_alert_log(sig_prov,
                              group_cols  = c("condition", "prov_"),
                              recent_days = 28L)

if (nrow(alert_prov) > 0L) {
  alert_prov[, Province := prov_label(prov_)]
  heat <- alert_prov[method == "CUSUM",
                     .(score = max(peak_value, na.rm = TRUE)),
                     by = .(Condition = condition, Province)]
} else {
  heat <- data.table(Condition = character(), Province = character(),
                     score = numeric())
}
```

```{r}
#| label: fig-signal-heatmap
#| fig-height: 5
if (nrow(heat) == 0L) {
  htmltools::tags$p(
    "No province-level CUSUM signals in the last 28 days.",
    style = "color:#475569;font-style:italic;"
  )
} else {
  plot_ly(heat,
          x = ~Province, y = ~Condition, z = ~score,
          type = "heatmap",
          colorscale = list(c(0, "#fee2e2"), c(1, "#7f1d1d")),
          hoverinfo = "text",
          text = ~paste0(Condition, " — ", Province,
                         "<br>Peak CUSUM: ", round(score, 2))) |>
    layout(xaxis = list(title = ""),
           yaxis = list(title = ""),
           margin = list(l = 160))
}
```

---

## Recent 14-day Surge — Quick Triage

Conditions ordered by absolute change in the most recent 14-day window vs the
preceding 14 days. Use the **signal table above** as the primary trigger; this
view is a complementary, formula-free sanity check and is **not** delay-corrected.

```{r}
#| label: surge-calc
recent_start <- latest_date - 13
prior_start  <- latest_date - 27
prior_end    <- latest_date - 14

recent <- agg_nat[date >= recent_start, .(recent_cases = sum(count)), by = condition]
prior  <- agg_nat[date >= prior_start & date <= prior_end,
                  .(prior_cases = sum(count)), by = condition]

surge <- merge(recent, prior, by = "condition", all = TRUE)
surge[is.na(recent_cases), recent_cases := 0L]
surge[is.na(prior_cases),  prior_cases  := 0L]
surge[, abs_change := recent_cases - prior_cases]
surge[, pct_change := fifelse(
  prior_cases > 0,
  round((recent_cases - prior_cases) / prior_cases * 100, 1),
  fifelse(recent_cases > 0, Inf, 0)
)]
surge[, low_count := recent_cases < 10]
setorder(surge, -abs_change)

flagged <- surge[recent_cases >= 5 & pct_change > 0]
```

```{r}
#| label: tbl-surge
display <- flagged[, .(
  Condition       = condition,
  `Last 14 days`  = recent_cases,
  `Prior 14 days` = prior_cases,
  `Abs Δ`         = abs_change,
  `% Δ`           = fifelse(is.infinite(pct_change), "New", paste0(pct_change, "%")),
  Caveat          = fifelse(low_count, "low count", "")
)]

DT::datatable(display, rownames = FALSE,
              options = list(pageLength = 15, dom = "ftp",
                             order = list(list(3, "desc"))))
```

---

## Seasonal Context — Top Flagged Condition

Current epiyear (solid red) overlaid on the historical median + IQR + 5–95%
ribbon from the prior five years. The dotted segment marks provisional weeks
inside the reporting-delay window.

```{r}
#| label: fig-seasonal
#| fig-cap: "Weekly notifications vs 5-year historical baseline"
#| fig-height: 5
top_cond <- if (nrow(alert_nat) > 0L) {
              alert_nat$condition[1]
            } else if (nrow(flagged) > 0L) {
              flagged$condition[1]
            } else {
              agg_nat[, .(tot = sum(count)), by = condition][order(-tot)][1, condition]
            }

cond_dt <- agg_nat[condition == top_cond, .(date, count)]

plot_epicurve_seasonal(
  cond_dt,
  title         = top_cond,
  subtitle      = "Current year vs historical median + IQR + 5–95% baseline",
  truncate_days = TRUNC_DAYS
)
```

---

## Trend Sparklines — Top Flagged Conditions

```{r}
#| label: fig-sparklines
#| fig-height: 8
top_flagged <- if (nrow(alert_nat) > 0L) {
                 head(unique(alert_nat$condition), 8)
               } else {
                 head(flagged$condition, 8)
               }

if (length(top_flagged) > 0L) {
  spark_data <- agg_nat[condition %chin% top_flagged]
  spark_data[, week := mmwr_week_start(date)]
  spark_weekly <- spark_data[week >= (latest_date - 26 * 7),
                             .(cases = sum(count)), by = .(condition, week)]

  plot_ly(spark_weekly,
          x = ~week, y = ~cases,
          color = ~condition, split = ~condition,
          type = "scatter", mode = "lines+markers",
          marker = list(size = 4)) |>
    layout(xaxis  = list(title = "Week (MMWR start)"),
           yaxis  = list(title = "Weekly cases"),
           legend = list(orientation = "h", y = -0.15),
           title  = "26-week trend — flagged conditions")
} else {
  htmltools::tags$p("No flagged conditions to plot.",
                    style = "color:#475569;font-style:italic;")
}
```

---

::: {.callout-tip}
## Related Sections
- [Forecasts & Scenarios](../by-forecasts/index.html) — short-term predictive models fed by signals
- [Burden of Disease](../burden-of-disease/index.html) — quantified health impact of conditions
- [By Condition](../by-condition/index.html) — condition-level national trends
- [By Province](../by-province/index.html) — province-level surveillance detail
- [Methods](../methods.html) — signal detection methodology documentation
:::
