---
title: "Trending Conditions"
subtitle: "Signal detection, emerging hotspots & alert log"
format:
  html:
    toc: true
---

::: {.callout-note collapse="true"}
## Development Roadmap — Trending / Signal Detection

| Phase | Feature | Status |
|-------|---------|--------|
| **v0.1** | Recent surge table — conditions with biggest % increase (last 14 days vs prior 14) | ✅ Done |
| **v0.1** | Interactive trend sparklines for flagged conditions | ✅ Done |
| **v0.2** | CUSUM / EARS algorithm signal detection integration | 🔲 Planned |
| **v0.2** | Province → district hotspot drill-down for flagged conditions | 🔲 Planned |
| **v0.3** | Analyst notes & alert log (manual triage layer) | 🔲 Planned |
| **v0.3** | Automated alert email / webhook on threshold breach | 🔲 Planned |
| **v0.4** | Link to Forecasts page — auto-trigger EpiNow2 for flagged conditions | 🔲 Planned |

*Data pipeline:* `R/prepare_dashboard_data.r` → `data/processed/agg_national.rds`
Forecasting extension: `by-forecasts/index.qmd` (uses `EpiNow2`)
:::

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(data.table)
  library(plotly)
  library(DT)
  library(lubridate)
})

agg_nat <- readRDS("data/processed/agg_national.rds")
setDT(agg_nat)
agg_nat[, date := as.Date(date)]

latest_date <- max(agg_nat$date)
```

**Data through:** `r format(latest_date, "%d %B %Y")`

---

## Emerging Signals — Last 14 Days

Conditions with the largest percentage increase comparing the most recent 14-day window to the preceding 14 days.

```{r}
#| label: surge-calc
recent_start <- latest_date - 13
prior_start  <- latest_date - 27
prior_end    <- latest_date - 14

recent <- agg_nat[date >= recent_start, .(recent_cases = sum(count)), by = condition]
prior  <- agg_nat[date >= prior_start & date <= prior_end, .(prior_cases = sum(count)), by = condition]

surge <- merge(recent, prior, by = "condition", all = TRUE)
surge[is.na(recent_cases), recent_cases := 0L]
surge[is.na(prior_cases), prior_cases := 0L]
surge[, pct_change := fifelse(
  prior_cases > 0,
  round((recent_cases - prior_cases) / prior_cases * 100, 1),
  fifelse(recent_cases > 0, Inf, 0)
)]
surge[, abs_change := recent_cases - prior_cases]
setorder(surge, -pct_change)

# Flag conditions with material increase
flagged <- surge[recent_cases >= 5 & pct_change > 0]
```

### Surge Table

```{r}
#| label: tbl-surge
display <- flagged[, .(
  Condition      = condition,
  `Last 14 days` = recent_cases,
  `Prior 14 days`= prior_cases,
  `Abs Δ`        = abs_change,
  `% Δ`          = fifelse(is.infinite(pct_change), "New", paste0(pct_change, "%"))
)]

DT::datatable(display, rownames = FALSE,
              options = list(pageLength = 15, dom = "ftp", order = list(list(3, "desc"))))
```

---

## Trend Sparklines — Flagged Conditions

Weekly trends for the top flagged conditions over the last 26 weeks.

```{r}
#| label: fig-sparklines
#| fig-height: 8
top_flagged <- head(flagged$condition, 8)
spark_data  <- agg_nat[condition %chin% top_flagged]
spark_data[, week := floor_date(date, "week")]
spark_weekly <- spark_data[week >= (latest_date - 26*7),
                           .(cases = sum(count)), by = .(condition, week)]

plot_ly() %>%
  {
    p <- .
    for (cond in top_flagged) {
      d <- spark_weekly[condition == cond]
      p <- add_trace(p, data = d, x = ~week, y = ~cases,
                     type = "scatter", mode = "lines+markers",
                     name = cond, marker = list(size = 4))
    }
    p
  } %>%
  layout(xaxis = list(title = "Week"),
         yaxis = list(title = "Weekly Cases"),
         legend = list(orientation = "h", y = -0.15),
         title = "26-Week Trend for Flagged Conditions")
```

---

## Next Steps

::: {.callout-tip}
### Future: CUSUM / EARS Integration

The trending page will integrate formal signal-detection algorithms (CUSUM, EARS C1/C2/C3) from the `surveillance` R package. Detected signals will be highlighted automatically and linked to the **Forecasts** tab for EpiNow2-based short-term projections.

See [Forecasts →](../by-forecasts/index.html)
:::

---

::: {.callout-tip}
## Related Sections
- [Forecasts & Scenarios](../by-forecasts/index.html) — short-term predictive models fed by signals
- [Burden of Disease](../burden-of-disease/index.html) — quantified health impact of conditions
- [By Condition](../by-condition/index.html) — condition-level national trends
- [By Province](../by-province/index.html) — province-level surveillance detail
- [Methods](../methods.html) — signal detection methodology documentation
:::
