---
title: "Agricultural & Stock Remedy Poisoning"
subtitle: "Epidemiological Description, South Africa · NMC Surveillance"
date: today
sidebar: special-reports
execute:
  echo: false
  warning: false
  message: false
  error: false
  cache: false
---

::: {.callout-important appearance="minimal"}
**Priority condition** — Agricultural and stock remedy poisoning is a current public health priority in South Africa. This report describes the epidemiology based on NMC notifiable conditions surveillance data, 2020–present.
:::

```{r setup}
#| label: setup
#| include: false
#| cache: false

pacman::p_load(
  ggplot2, gtsummary, tidyverse, magrittr,
  conflicted, scales, data.table, lubridate,
  flextable, gt, ggrepel, zoo,
  janitor, binom, broom
)

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::lag)
conflicted::conflicts_prefer(tidyr::extract)
conflicted::conflicts_prefer(dplyr::first)
conflicted::conflicts_prefer(lubridate::year)
conflicted::conflicts_prefer(lubridate::month)
conflicted::conflicts_prefer(lubridate::quarter)

library(NMCleaner)

# ── Load master dataset ───────────────────────────────────────────────────────


new_master <- arrow::read_feather("~/Desktop/SAFETP/CLA/NMC_database/master/new_master.feather")

# ── Filter to Agricultural poisoning only ────────────────────────────────────
CONDITION <- "Agricultural or stock remedy poisoning"

new_master %>%
  filter(
    condition %in% CONDITION,
    as.integer(Year_notification) %in% 2020:year(Sys.Date()),
    !is.na(prov_)
  ) %>%
  mutate(
    Year_notification = as.integer(Year_notification),
    patient_vital_status = if_else(
      patient_vital_status %like% "Unknown", "Alive", patient_vital_status
    ),
    Age_years = as.numeric(Age_years),
    age_group = cut(
      Age_years,
      breaks = c(-Inf, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, Inf),
      labels = c("<1", "1\u20134", "5\u20139", "10\u201314", "15\u201319",
                 "20\u201324", "25\u201329", "30\u201334", "35\u201339",
                 "40\u201344", "45\u201349", "50\u201354", "55\u201359",
                 "60\u201364", "65+"),
      right  = FALSE
    ),
    sex = case_when(
      patient_gender %in% c("Male", "M")   ~ "Male",
      patient_gender %in% c("Female", "F") ~ "Female",
      TRUE                                  ~ NA_character_
    ),
    deceased = patient_vital_status %in% "Deceased"
  ) -> poison

# ── Province code → full name lookup ─────────────────────────────────────────
prov_labels <- c(
  "EC"  = "Eastern Cape",  "FS"  = "Free State",    "GP"  = "Gauteng",
  "KZN" = "KwaZulu-Natal", "LP"  = "Limpopo",       "MP"  = "Mpumalanga",
  "NC"  = "Northern Cape", "NW"  = "North West",    "WC"  = "Western Cape"
)
poison <- poison %>%
  mutate(Province = factor(prov_labels[prov_], levels = unname(prov_labels)))

# ── Shared theme ──────────────────────────────────────────────────────────────
theme_nmc <- function(...) {
  theme_classic() +
    theme(
      text              = element_text(family = "sans", size = 10, color = "#333333"),
      panel.background  = element_rect(fill = "#F8F8F8"),
      plot.background   = element_rect(fill = "#F8F8F8"),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      axis.ticks        = element_line(linewidth = 0.15, color = "grey40"),
      legend.background = element_rect(fill = "#F8F8F8", linewidth = 0.25,
                                       linetype = "solid", colour = "grey50"),
      legend.key        = element_blank(),
      legend.title      = element_text(face = "bold", size = 9, color = "#333333"),
      legend.text       = element_text(size = 9, color = "#333333"),
      strip.background  = element_rect(color = NA, fill = NA),
      ...
    )
}
```

::: {.callout-note appearance="minimal" collapse="true"}
## Report metadata
| | |
|---|---|
| **Condition** | `r CONDITION` |
| **Years** | 2020 – `r lubridate::year(Sys.Date())` |
| **Data source** | NMC Notifiable Medical Conditions surveillance database |
| **Generated** | `r format(Sys.Date(), "%d %B %Y")` |
:::

---

# Situation Update for Policy Makers {#sec-executive}

```{r exec-setup}
#| label: exec-setup
#| include: false

# ── Current year and comparison calculations ──────────────────────────────────
current_year <- year(Sys.Date())
prev_year    <- current_year - 1
current_month <- month(Sys.Date())

# Year-to-date (Jan 1 to current date) for current and previous year
ytd_current <- poison %>%
 filter(Year_notification == current_year,
         month(as_date(notification_date)) <= current_month)
ytd_prev <- poison %>%
  filter(Year_notification == prev_year,
         month(as_date(notification_date)) <= current_month)

n_ytd_current   <- nrow(ytd_current)
n_ytd_prev      <- nrow(ytd_prev)
deaths_ytd_curr <- sum(ytd_current$deceased, na.rm = TRUE)
deaths_ytd_prev <- sum(ytd_prev$deceased, na.rm = TRUE)
cfr_ytd_curr    <- round(deaths_ytd_curr / max(n_ytd_current, 1) * 100, 1)
cfr_ytd_prev    <- round(deaths_ytd_prev / max(n_ytd_prev, 1) * 100, 1)

pct_change_n      <- round((n_ytd_current - n_ytd_prev) / max(n_ytd_prev, 1) * 100)
pct_change_deaths <- round((deaths_ytd_curr - deaths_ytd_prev) / max(deaths_ytd_prev, 1) * 100)

# Direction arrows and colours
arrow_n      <- if (pct_change_n > 0) "↑" else if (pct_change_n < 0) "↓" else "→"
arrow_deaths <- if (pct_change_deaths > 0) "↑" else if (pct_change_deaths < 0) "↓" else "→"
col_n        <- if (pct_change_n > 10) "#C62828" else if (pct_change_n < -10) "#2E7D32" else "#F57C00"
col_deaths   <- if (pct_change_deaths > 10) "#C62828" else if (pct_change_deaths < -10) "#2E7D32" else "#F57C00"

# Children under 5 burden — current and previous year
under5_ytd_curr <- ytd_current %>% filter(Age_years < 5)
under5_ytd_prev <- ytd_prev %>% filter(Age_years < 5)
n_under5_curr   <- nrow(under5_ytd_curr)
n_under5_prev   <- nrow(under5_ytd_prev)
deaths_under5_curr <- sum(under5_ytd_curr$deceased, na.rm = TRUE)
deaths_under5_prev <- sum(under5_ytd_prev$deceased, na.rm = TRUE)
pct_change_under5 <- round((n_under5_curr - n_under5_prev) / max(n_under5_prev, 1) * 100)
pct_change_deaths_under5 <- round((deaths_under5_curr - deaths_under5_prev) / max(deaths_under5_prev, 1) * 100)
arrow_under5 <- if (pct_change_under5 > 0) "↑" else if (pct_change_under5 < 0) "↓" else "→"
arrow_deaths_under5 <- if (pct_change_deaths_under5 > 0) "↑" else if (pct_change_deaths_under5 < 0) "↓" else "→"

# Children under 12 burden — current and previous year
under12_ytd_curr <- ytd_current %>% filter(Age_years < 12)
under12_ytd_prev <- ytd_prev %>% filter(Age_years < 12)
n_under12_curr   <- nrow(under12_ytd_curr)
n_under12_prev   <- nrow(under12_ytd_prev)
deaths_under12_curr <- sum(under12_ytd_curr$deceased, na.rm = TRUE)
deaths_under12_prev <- sum(under12_ytd_prev$deceased, na.rm = TRUE)
pct_change_under12 <- round((n_under12_curr - n_under12_prev) / max(n_under12_prev, 1) * 100)
pct_change_deaths_under12 <- round((deaths_under12_curr - deaths_under12_prev) / max(deaths_under12_prev, 1) * 100)
arrow_under12 <- if (pct_change_under12 > 0) "↑" else if (pct_change_under12 < 0) "↓" else "→"
arrow_deaths_under12 <- if (pct_change_deaths_under12 > 0) "↑" else if (pct_change_deaths_under12 < 0) "↓" else "→"

# Top 3 provinces this year
top3_prov <- ytd_current %>%
  count(Province, sort = TRUE) %>%
  slice(1:3) %>%
  pull(Province) %>%
  paste(collapse = ", ")
```

::: {.panel-tabset}

## Key Metrics

```{r exec-metrics}
#| label: tbl-exec-metrics
#| tbl-cap: !expr 'paste0("Year-to-date comparison: Jan–", format(Sys.Date(), "%b %Y"), " vs same period ", prev_year)'

tibble(
  Indicator = c(
    "Notifications (YTD)",
    "Deaths (YTD)",
    "Case Fatality Ratio",
    "Children <5 (notifications)",
    "Children <5 (deaths)",
    "Children <12 (notifications)",
    "Children <12 (deaths)",
    "Top 3 provinces (by case count)"
  ),
  `Current Year` = c(
    scales::comma(n_ytd_current),
    scales::comma(deaths_ytd_curr),
    paste0(cfr_ytd_curr, "%"),
    scales::comma(n_under5_curr),
    scales::comma(deaths_under5_curr),
    scales::comma(n_under12_curr),
    scales::comma(deaths_under12_curr),
    top3_prov
  ),
  `Same Period Last Year` = c(
    scales::comma(n_ytd_prev),
    scales::comma(deaths_ytd_prev),
    paste0(cfr_ytd_prev, "%"),
    scales::comma(n_under5_prev),
    scales::comma(deaths_under5_prev),
    scales::comma(n_under12_prev),
    scales::comma(deaths_under12_prev),
    "—"
  ),
  `Change` = c(
    paste0(arrow_n, " ", ifelse(pct_change_n >= 0, "+", ""), pct_change_n, "%"),
    paste0(arrow_deaths, " ", ifelse(pct_change_deaths >= 0, "+", ""), pct_change_deaths, "%"),
    paste0(ifelse(cfr_ytd_curr >= cfr_ytd_prev, "↑", "↓"), " ",
           ifelse(cfr_ytd_curr - cfr_ytd_prev >= 0, "+", ""),
           round(cfr_ytd_curr - cfr_ytd_prev, 1), " pp"),
    paste0(arrow_under5, " ", ifelse(pct_change_under5 >= 0, "+", ""), pct_change_under5, "%"),
    paste0(arrow_deaths_under5, " ", ifelse(pct_change_deaths_under5 >= 0, "+", ""), pct_change_deaths_under5, "%"),
    paste0(arrow_under12, " ", ifelse(pct_change_under12 >= 0, "+", ""), pct_change_under12, "%"),
    paste0(arrow_deaths_under12, " ", ifelse(pct_change_deaths_under12 >= 0, "+", ""), pct_change_deaths_under12, "%"),
    "—"
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = 1:3, j = 1) %>%
  color(i = 1, j = 4, color = col_n) %>%
  color(i = 2, j = 4, color = col_deaths) %>%
  bg(i = 4:7, bg = "#FFF8E1") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 11, part = "all") %>%
  align(j = 2:4, align = "center", part = "all") %>%
  add_footer_lines(paste0("Data extracted: ", format(Sys.Date(), "%d %B %Y"),
                          ". YTD = Year-to-date (Jan 1 – ", format(Sys.Date(), "%d %b"), "). pp = percentage points."))
```

## 12-Month Trend

```{r exec-trend}
#| label: fig-exec-trend
#| fig-cap: "Monthly notification trend — last 12 months with 3-month rolling average. Shaded area highlights the Sep–Nov seasonal peak period."
#| fig-height: 4

# Last 12 complete months
end_date   <- floor_date(Sys.Date(), "month") - days(1)
start_date <- floor_date(end_date - months(11), "month")

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  #filter(month_date >= start_date, month_date <= end_date) %>%
  count(month_date) %>%
  arrange(month_date) %>%
  mutate(
    rolling3 = zoo::rollmean(n, k = 3, fill = NA, align = "right"),
    is_peak  = month(month_date) %in% c(9, 10, 11)  # Sep-Nov peak season
  ) -> trend12_df

# Highlight rectangle for peak season months in this 12-month window
peak_rects <- trend12_df %>%
  filter(is_peak) %>%
  summarise(xmin = min(month_date, na.rm = TRUE) - days(15),
            xmax = max(month_date, na.rm = TRUE) + days(15)) %>%
  slice(1)

ggplot(trend12_df, aes(x = month_date)) +
  {if (nrow(peak_rects) > 0) annotate("rect",
    xmin = peak_rects$xmin, xmax = peak_rects$xmax,
    ymin = -Inf, ymax = Inf, fill = "#FFECB3", alpha = 0.5)} +
  geom_col(aes(y = n), fill = "#90CAF9", colour = "#64B5F6", width = 22) +
  geom_line(aes(y = rolling3), colour = "#C62828", linewidth = 1.3, na.rm = TRUE) +
  geom_point(aes(y = rolling3), colour = "#C62828", size = 2.5, na.rm = TRUE) +
  geom_text(aes(y = n, label = n), vjust = -0.5, size = 3.2, colour = "#333333") +
  scale_x_date(
    date_labels      = "%b %Y",
    date_breaks      = "1 month",
    expand           = expansion(mult = c(0.03, 0.03))
  ) +
  scale_y_continuous(name = "Notifications", expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL,
       caption = "Red line = 3-month rolling average. Yellow shading = Sep–Nov seasonal peak.") +
  theme_nmc() +
  theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1, vjust = 1))
```

## Provincial Burden

```{r exec-prov-heatmap}
#| label: fig-exec-prov
#| fig-cap: "Provincial burden heatmap — quarterly notifications, deaths, and incidence per 100,000, last 2 years. Darker = higher incidence."
#| fig-height: 5.5

# Get provincial population for incidence calculation
prov_pop <- NMCleaner::pop %>%
  filter(as.integer(Year) == year(Sys.Date())) %>%
  group_by(prov) %>%
  summarise(population = sum(Population), .groups = "drop")

# Last 8 quarters with incidence
poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(
    nd = as_date(notification_date),
    qtr = paste0(year(nd), " Q", quarter(nd))
  ) %>%
  filter(nd >= (Sys.Date() - years(2))) %>%
  group_by(prov_, qtr) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") %>%
  complete(prov_, qtr, fill = list(n = 0, deaths = 0)) %>%
  left_join(prov_pop, by = c("prov_" = "prov")) %>%
  mutate(
    Province = factor(prov_labels[prov_], levels = unname(prov_labels)),
    # Quarterly incidence annualized (multiply by 4) per 100k
    inc = round(n / population * 1e5 * 4, 1)
  ) -> prov_qtr_df

# Order quarters chronologically
qtr_order <- prov_qtr_df %>% distinct(qtr) %>% arrange(qtr) %>% pull(qtr)
prov_qtr_df$qtr <- factor(prov_qtr_df$qtr, levels = qtr_order)

ggplot(prov_qtr_df, aes(x = qtr, y = Province, fill = inc)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(n, ifelse(deaths > 0, paste0(" (", deaths, "†)"), ""),
                               "\n", inc)),
            size = 2.5, colour = "black", lineheight = 0.85) +
  scale_fill_gradient(low = "#E3F2FD", high = "#0D47A1", name = "Incidence\n(per 100k/yr)") +
  labs(x = "Quarter", y = NULL,
       caption = "Top number = notifications (†deaths); bottom = annualized incidence per 100,000.") +
  theme_nmc() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "right",
    panel.grid = element_blank()
  )
```

:::

---

# Summary Statistics {#sec-summary}

```{r summary-stats}
#| label: tbl-summary
#| tbl-cap: "Summary of agricultural or stock remedy poisoning notifications, South Africa, 2020–present."

total_n      <- nrow(poison)
total_deaths <- sum(poison$deceased, na.rm = TRUE)
overall_cfr  <- round(total_deaths / total_n * 100, 1)

tibble(
  Metric = c(
    "Total notifications",
    "Total deaths reported",
    "Overall case fatality ratio",
    "Provinces affected",
    "Districts affected",
    "Unique reporting facilities",
    "Date range"
  ),
  Value = c(
    scales::comma(total_n),
    scales::comma(total_deaths),
    paste0(overall_cfr, "%"),
    as.character(poison$prov_ %>% na.omit() %>% n_distinct()),
    as.character(poison$district %>% na.omit() %>% n_distinct()),
    as.character(poison$facility %>% na.omit() %>% n_distinct()),
    paste(
      format(as_date(min(poison$notification_date, na.rm = TRUE)), "%d %b %Y"),
      "to",
      format(as_date(max(poison$notification_date, na.rm = TRUE)), "%d %b %Y")
    )
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 10, part = "all")
```

---

# Temporal Trends {#sec-trends}

## Annual notifications and CFR

```{r annual-trend}
#| label: fig-annual-trend
#| fig-cap: "Annual count of agricultural or stock remedy poisoning notifications and case fatality ratio (CFR), South Africa, 2020–present."

poison %>%
  group_by(year = Year_notification) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") -> annual_df

sf_ax <- max(annual_df$n) / max(annual_df$cfr, na.rm = TRUE) * 1.1

ggplot(annual_df, aes(x = year)) +
  geom_col(aes(y = n, fill = "Notifications"), alpha = 0.8, width = 0.6) +
  geom_line(aes(y = cfr * sf_ax, colour = "CFR (%)"), linewidth = 1.2, group = 1) +
  geom_point(aes(y = cfr * sf_ax, colour = "CFR (%)"), size = 3) +
  geom_text(aes(y = n + max(n) * 0.02, label = scales::comma(n)),
            size = 3.2, vjust = 0, color = "#333333") +
  scale_y_continuous(
    name     = "Notifications (n)",
    sec.axis = sec_axis(~ . / sf_ax, name = "CFR (%)", breaks = pretty_breaks(4))
  ) +
  scale_fill_manual(values   = c("Notifications" = "#90CAF9"), name = NULL) +
  scale_colour_manual(values = c("CFR (%)" = "#C62828"), name = NULL) +
  scale_x_continuous(breaks = annual_df$year) +
  labs(x = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Monthly epicurve

```{r epicurve-monthly}
#| label: fig-epicurve-monthly
#| fig-cap: "Monthly notification counts for agricultural or stock remedy poisoning, South Africa, 2020–present. Line = 3-month rolling average."

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  count(month_date) %>%
  arrange(month_date) %>%
  mutate(rolling3 = zoo::rollmean(n, k = 3, fill = NA, align = "right")) -> epicurve_df

ggplot(epicurve_df, aes(x = month_date)) +
  geom_col(aes(y = n), fill = "#90CAF9", colour = "#64B5F6", width = 25) +
  geom_line(aes(y = rolling3), colour = "#C62828", linewidth = 1.1, na.rm = TRUE) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months",
               date_minor_breaks = "3 months", expand = expansion(add = c(0, 0))) +
  scale_y_continuous(name = "Notifications (n)", breaks = pretty_breaks(5)) +
  labs(x = "Date",
       caption = "Red line = 3-month rolling average") +
  theme_nmc() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8))
```

## Seasonal pattern — epiweek heatmap

```{r seasonal-heatmap}
#| label: fig-seasonal-heatmap
#| fig-cap: "Epiweek × year heatmap of agricultural or stock remedy poisoning notification counts, South Africa. The Sep–Nov seasonal peak is clearly visible."
#| fig-height: 5

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(
    nd      = as_date(notification_date),
    yr      = lubridate::year(nd),
    epiweek = lubridate::isoweek(nd)
  ) %>%
  count(year = yr, epiweek) %>%
  complete(year = 2020:year(Sys.Date()), epiweek = 1:52, fill = list(n = 0)) -> heat_df

ggplot(heat_df, aes(x = epiweek,
                    y = factor(year, levels = rev(sort(unique(year)))),
                    fill = n)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "white", mid = "#FFF9C4", high = "#B71C1C",
    midpoint = median(heat_df$n[heat_df$n > 0], na.rm = TRUE),
    name = "Notifications (n)"
  ) +
  scale_x_continuous(breaks = c(1, seq(5, 52, 5)), expand = c(0, 0)) +
  labs(x = "Epiweek", y = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.width = unit(1.5, "cm"),
        panel.grid.major = element_blank())
```

---

# Age & Sex Demographics {#sec-demographics}

## Population pyramid

```{r age-sex-pyramid}
#| label: fig-age-sex-pyramid
#| fig-cap: "Age–sex population pyramid of agricultural or stock remedy poisoning notifications, South Africa, 2020–present."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  count(age_group, sex) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyramid_df

max_n <- max(abs(pyramid_df$n))
step  <- ifelse(max_n > 500, 200, ifelse(max_n > 200, 50, 25))
br    <- seq(0, ceiling(max_n / step) * step, step)

ggplot(pyramid_df, aes(x = age_group, y = n_plot, fill = sex)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  geom_hline(yintercept = 0, colour = "white", linewidth = 0.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y = -max_n * 0.5, label = "Male",   color = "#1565C0", fontface = "bold", size = 3.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y =  max_n * 0.5, label = "Female", color = "#C62828", fontface = "bold", size = 3.5) +
  labs(x = "Age group (years)", y = "Number of notifications") +
  theme_nmc() +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(color = "#DDDDDD", linewidth = 0.3))
```

## Population pyramid by vital status

```{r pyramid-by-vital}
#| label: fig-pyramid-vital
#| fig-cap: "Age–sex pyramid overlaid with vital status. Solid fill = alive; faded = deceased."
#| fig-height: 5.5

poison %>%
  filter(!is.na(age_group), !is.na(sex),
         patient_vital_status %in% c("Alive", "Deceased")) %>%
  count(age_group, sex, patient_vital_status) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyr_vital_df

ggplot(pyr_vital_df,
       aes(x = age_group, y = n_plot, fill = sex, alpha = patient_vital_status)) +
  geom_col(width = 0.8, position = "stack") +
  coord_flip() +
  scale_alpha_manual(values = c("Alive" = 0.9, "Deceased" = 0.35), name = "Vital status") +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = "Sex") +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  geom_hline(yintercept = 0, colour = "white") +
  labs(x = "Age group (years)", y = "Notifications (n)") +
  theme_nmc() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(order = 1), alpha = guide_legend(order = 2))
```

## Age group distribution over time

```{r age-time}
#| label: fig-age-time
#| fig-cap: "Annual proportion of agricultural or stock remedy poisoning notifications by age group, South Africa."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group)) %>%
  count(Year_notification, age_group) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) -> age_time_df

ggplot(age_time_df, aes(x = factor(Year_notification), y = pct, fill = age_group)) +
  geom_col(position = "fill", width = 0.75) +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_fill_viridis_d(option = "turbo", direction = -1, name = "Age group") +
  labs(x = "Year", y = "Proportion of notifications (%)") +
  theme_nmc() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))
```

## Sex distribution over time

```{r sex-time}
#| label: fig-sex-time
#| fig-cap: "Annual proportion of agricultural or stock remedy poisoning notifications by sex, South Africa."
#| fig-height: 4.5

poison %>%
  filter(!is.na(sex)) %>%
  count(Year_notification, sex) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) -> sex_time_df

ggplot(sex_time_df, aes(x = factor(Year_notification), y = pct, fill = sex)) +
  geom_col(position = "fill", width = 0.75) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_fill(vjust = 0.5),
            size = 3.2, color = "white", fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  labs(x = "Year", y = "Proportion of notifications (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Case fatality ratio by age group and sex

```{r cfr-age-sex}
#| label: fig-cfr-age-sex
#| fig-cap: "Case fatality ratio (%) by age group and sex for agricultural or stock remedy poisoning, South Africa, 2020–present."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  group_by(age_group, sex) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") -> cfr_df

ggplot(cfr_df, aes(x = age_group, y = cfr, fill = sex)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = paste0(round(cfr, 1), "%")),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3, color = "#333333") +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  scale_y_continuous(limits = c(0, max(cfr_df$cfr, na.rm = TRUE) * 1.25)) +
  labs(x = "Age group (years)", y = "Case fatality ratio (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Summary demographics table

```{r demo-table}
#| label: tbl-demographics
#| tbl-cap: "Summary demographics of agricultural or stock remedy poisoning notifications by year and vital status, South Africa."

poison %>%
  mutate(year_chr = factor(Year_notification)) %>%
  select(year_chr, sex, age_group, Age_years, patient_vital_status, Province) %>%
  tbl_strata(
    strata   = year_chr,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by        = patient_vital_status,
        missing   = "no",
        label     = list(sex ~ "Sex", age_group ~ "Age group",
                         Age_years ~ "Age (years)", Province ~ "Province"),
        type      = list(Age_years ~ "continuous"),
        statistic = list(
          all_categorical() ~ "{n}/{N} ({p}%)",
          Age_years ~ "{median} ({p25}\u2013{p75})"
        )
      ) %>%
      bold_labels()
  ) %>%
  as_flex_table()
```

---

# Geographical Distribution {#sec-geo}

```{r geo-setup}
#| include: false

# ── Provincial incidence ──────────────────────────────────────────────────────
NMCleaner::pop %>%
  mutate(Year = as.integer(Year)) %>%
  group_by(prov, Year) %>%
  summarise(population = sum(Population), .groups = "drop") -> pop_prov

poison %>%
  group_by(prov_, Province, year = Year_notification) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(pop_prov, by = c("prov_" = "prov", "year" = "Year")) %>%
  mutate(incidence = n / population * 1e5) -> incidence_prov_df

# ── District incidence ────────────────────────────────────────────────────────
NMCleaner::pop %>%
  filter(as.integer(Year) %in% 2020:year(Sys.Date())) %>%
  group_by(district_standard) %>%
  summarise(pop_total = sum(Population), .groups = "drop") -> pop_dist

poison %>%
  filter(!district %in% c("Unknown", "Isazi Import", "Not Applicable"),
         !is.na(district)) %>%
  mutate_district_name("district") %>%
  group_by(Province, district_standard) %>%
  summarise(
    Notifications = n(),
    Deaths        = sum(deceased, na.rm = TRUE),
    CFR           = round(Deaths / Notifications * 100, 1),
    .groups = "drop"
  ) %>%
  left_join(pop_dist, by = "district_standard") %>%
  mutate(
    Incidence = round(Notifications / pop_total * 1e5, 1),
    pop_total = round(pop_total / 1e3, 1)          # display in thousands
  ) -> incidence_dist_df
```

## Provincial incidence over time

```{r prov-incidence-line}
#| label: fig-prov-incidence-line
#| fig-cap: "Incidence of agricultural or stock remedy poisoning notifications per 100,000 population by province and year, South Africa."
#| fig-height: 5.5

incidence_prov_df %>%
  group_by(Province) %>%
  filter(incidence == max(incidence, na.rm = TRUE)) %>%
  ungroup() -> peak_prov_df

ggplot(incidence_prov_df, aes(x = year, y = incidence, colour = Province, group = Province)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  ggrepel::geom_label_repel(data = peak_prov_df, aes(label = Province),
    size = 2.8, show.legend = FALSE, label.size = 0.15, max.overlaps = 25) +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  scale_x_continuous(breaks = unique(incidence_prov_df$year)) +
  labs(x = "Year", y = "Incidence per 100,000 population") +
  theme_nmc() +
  theme(legend.position = "none")
```

## Provincial incidence table

```{r prov-table}
#| label: tbl-prov-incidence
#| tbl-cap: "Notifications and incidence per 100,000 by province and year, South Africa."

incidence_prov_df %>%
  mutate(year = as.character(year)) %>%
  pivot_wider(id_cols = Province, names_from = year,
              values_from = c(n, incidence), names_glue = "{.value}_{year}") %>%
  gt(rowname_col = "Province") %>%
  tab_spanner_delim(delim = "_") %>%
  fmt_number(starts_with("n_"),         decimals = 0, use_seps = TRUE) %>%
  fmt_number(starts_with("incidence_"), decimals = 1) %>%
  tab_header(title = md("**Provincial notifications and incidence per 100,000**")) %>%
  tab_source_note("Incidence per 100,000 using mid-year provincial population estimates.")
```

## Top 20 districts by notification count

```{r top-districts}
#| label: tbl-top-districts
#| tbl-cap: "Top 20 districts by total notification count, South Africa, 2020–present. Pink = CFR > 5%."

incidence_dist_df %>%
  arrange(desc(Notifications)) %>%
  slice(1:20) %>%
  rename(District = district_standard, `Pop (k)` = pop_total) %>%
  select(Province, District, Notifications, Deaths, `CFR (%)` = CFR,
         `Incidence per 100k` = Incidence, `Pop (k)`) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  bg(i = ~ `Incidence per 100k` > median(incidence_dist_df$Incidence, na.rm = TRUE) * 2,
     j = "Incidence per 100k", bg = "#FFF3E0") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 9, part = "all")
```

## Top 30 districts by incidence rate

```{r district-incidence-lollipop}
#| label: fig-district-incidence
#| fig-cap: "Top 30 districts ranked by incidence per 100,000 population, agricultural or stock remedy poisoning, South Africa, 2020–present. Districts with high incidence but low absolute counts are often missed by count-only rankings."
#| fig-height: 8

incidence_dist_df %>%
  filter(!is.na(Incidence)) %>%
  arrange(desc(Incidence)) %>%
  slice(1:30) %>%
  mutate(
    district_standard = fct_reorder(district_standard, Incidence),
    label = paste0(Province, " · ", round(Incidence, 1))
  ) -> top_inc_df

ggplot(top_inc_df, aes(x = Incidence, y = district_standard)) +
  geom_segment(aes(x = 0, xend = Incidence,
                   y = district_standard, yend = district_standard),
               colour = "grey75", linewidth = 0.5) +
  geom_point(aes(size = Notifications, colour = Province), alpha = 0.85) +
  geom_text(aes(label = round(Incidence, 1)), hjust = -0.5, size = 2.8,
            colour = "#333333") +
  scale_size_continuous(range = c(2, 9), name = "Notifications (n)") +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  labs(
    x = "Incidence per 100,000 population",
    y = NULL,
    caption = "Point size proportional to number of notifications. Colour = province."
  ) +
  theme_nmc() +
  theme(legend.position = "bottom",
        panel.grid.major.x = element_line(colour = "#EEEEEE", linewidth = 0.3))
```

## Count vs incidence — identifying overlooked high-burden districts

```{r district-scatter}
#| label: fig-district-scatter
#| fig-cap: "Notification count versus incidence rate per 100,000 population for all districts, South Africa, 2020–present. Quadrant lines = national median count and median incidence. Districts in the top-left quadrant have high incidence relative to their case count (small, high-burden populations)."
#| fig-height: 6

med_n   <- median(incidence_dist_df$Notifications, na.rm = TRUE)
med_inc <- median(incidence_dist_df$Incidence,     na.rm = TRUE)

incidence_dist_df %>%
  filter(!is.na(Incidence)) %>%
  mutate(quadrant = case_when(
    Notifications > med_n   & Incidence > med_inc ~ "High count, high incidence",
    Notifications <= med_n  & Incidence > med_inc ~ "Low count, high incidence",
    Notifications > med_n   & Incidence <= med_inc ~ "High count, low incidence",
    TRUE                                           ~ "Low count, low incidence"
  )) -> scatter_df

# label top 10 by incidence + top 10 by count
top_labels <- scatter_df %>%
  arrange(desc(Incidence)) %>% slice(1:10) %>%
  bind_rows(scatter_df %>% arrange(desc(Notifications)) %>% slice(1:10)) %>%
  distinct(district_standard, .keep_all = TRUE)

ggplot(scatter_df, aes(x = Notifications, y = Incidence)) +
  geom_hline(yintercept = med_inc, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_vline(xintercept = med_n,   linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = Province, size = pop_total), alpha = 0.75) +
  ggrepel::geom_label_repel(
    data = top_labels,
    aes(label = district_standard),
    size = 2.4, label.size = 0.1, max.overlaps = 25,
    fill = "white", colour = "#333333"
  ) +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  scale_size_continuous(range = c(1.5, 7), name = "Pop (thousands)") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    x = "Total notifications (n)",
    y = "Incidence per 100,000 population",
    caption = "Dashed lines = national medians. Top 10 districts by incidence and by count labelled."
  ) +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Full district incidence table

```{r tbl-district-incidence}
#| label: tbl-district-incidence
#| tbl-cap: "All districts with at least 5 notifications ranked by incidence per 100,000 population, South Africa, 2020–present."

incidence_dist_df %>%
  filter(Notifications >= 5, !is.na(Incidence)) %>%
  arrange(desc(Incidence)) %>%
  rename(District = district_standard, `Pop (k)` = pop_total,
         `Incidence per 100k` = Incidence, `CFR (%)` = CFR) %>%
  select(Province, District, Notifications, Deaths, `CFR (%)`,
         `Incidence per 100k`, `Pop (k)`) %>%
  flextable() %>%
  bold(part = "header") %>%
  color(i = ~ `Incidence per 100k` > quantile(incidence_dist_df$Incidence,
                                               0.9, na.rm = TRUE),
        j = "Incidence per 100k", color = "#C62828") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 9, part = "all") %>%
  add_footer_lines("Red = top 10% incidence districts. Pink = CFR > 5%.")
```

---

# Mortality Profile {#sec-mortality}

## Deaths by province and year

```{r deaths-prov}
#| label: fig-deaths-prov
#| fig-cap: "Annual deaths from agricultural or stock remedy poisoning by province, South Africa."
#| fig-height: 5

poison %>%
  filter(deceased) %>%
  count(year = Year_notification, Province) %>%
  ggplot(aes(x = factor(year), y = n, fill = Province)) +
  geom_col(position = "stack", width = 0.75) +
  scale_fill_brewer(palette = "Paired", name = "Province") +
  labs(x = "Year", y = "Deaths (n)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## CFR by province and year

```{r cfr-prov}
#| label: fig-cfr-prov
#| fig-cap: "Case fatality ratio (%) by province and year for agricultural or stock remedy poisoning, South Africa."
#| fig-height: 5

poison %>%
  group_by(year = Year_notification, Province) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") %>%
  ggplot(aes(x = factor(year), y = cfr, fill = Province)) +
  geom_col(position = "dodge", width = 0.85) +
  scale_fill_brewer(palette = "Paired", name = "Province") +
  labs(x = "Year", y = "CFR (%)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## Deaths in children under 12

```{r under12-deaths}
#| label: tbl-under12-deaths
#| tbl-cap: "Vital status of agricultural or stock remedy poisoning notifications in children under 12 years, by year, South Africa."

poison %>%
  filter(Year_notification %in% 2023:year(Sys.Date())) %>%
  mutate(
    under_12 = factor(if_else(Age_years <= 12, "Under 12 years", "12 years and over"),
                      levels = c("Under 12 years", "12 years and over"))
  ) %>%
  select(year = Year_notification, under_12, patient_vital_status, Age_years, sex, prov_) %>%
  tbl_strata(
    strata   = year,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by      = patient_vital_status, missing = "no",
        label   = list(under_12 ~ "Age category", Age_years ~ "Age (years)",
                       sex ~ "Sex", prov_ ~ "Province"),
        type    = list(Age_years ~ "continuous"), percent = "row"
      ) %>% bold_labels()
  ) %>%
  as_flex_table()%>%save_as_docx(path = "/Users/briday/Desktop/under12_deaths.docx")
```

---

# Reporting & Facility Analysis {#sec-facilities}

## Unique reporting facilities per year

```{r facilities-trend}
#| label: fig-facilities
#| fig-cap: "Number of unique health facilities reporting at least one agricultural or stock remedy poisoning notification per year, South Africa."
#| fig-height: 4

poison %>%
  filter(!is.na(facility)) %>%
  group_by(Year_notification) %>%
  summarise(n_facilities = n_distinct(facility), .groups = "drop") -> fac_df

ggplot(fac_df, aes(x = factor(Year_notification))) +
  geom_col(aes(y = n_facilities, fill = "Unique facilities"), width = 0.6, alpha = 0.85) +
  geom_text(aes(y = n_facilities, label = n_facilities), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Unique facilities" = "#4CAF50"), name = NULL) +
  labs(x = "Year", y = "Unique reporting facilities (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

---

# Post-Ban Burden: Cabinet Recommendation to Gazetted Ban {#sec-terbufos}

::: {.callout-important appearance="minimal"}
**Policy window** — The Cabinet recommended a ban on terbufos on **12 June 2025**. The ban was formally gazetted on **8 May 2026**. This section describes the poisoning and mortality burden observed during this 330-day interim period.
:::

```{r terbufos-setup}
#| label: terbufos-setup
#| include: false

ban_recommend <- as_date("2025-06-12")   # Cabinet recommendation date
ban_gazette   <- as_date("2026-05-08")   # Gazetted ban date

# Post-recommendation period subset
poison_ban <- poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(nd = as_date(notification_date)) %>%
  filter(nd >= ban_recommend, nd <= ban_gazette)

# Pre-recommendation equivalent period (same window duration, one year prior)
window_days  <- as.integer(ban_gazette - ban_recommend)
pre_start    <- ban_recommend - years(1)
pre_end      <- pre_start + days(window_days)

poison_pre <- poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(nd = as_date(notification_date)) %>%
  filter(nd >= pre_start, nd <= pre_end)

# Summary statistics
ban_n      <- nrow(poison_ban)
ban_deaths <- sum(poison_ban$deceased, na.rm = TRUE)
ban_cfr    <- round(ban_deaths / max(ban_n, 1) * 100, 1)
pre_n      <- nrow(poison_pre)
pre_deaths <- sum(poison_pre$deceased, na.rm = TRUE)
pre_cfr    <- round(pre_deaths / max(pre_n, 1) * 100, 1)
pct_chg_n  <- round((ban_n - pre_n) / max(pre_n, 1) * 100)
pct_chg_d  <- round((ban_deaths - pre_deaths) / max(pre_deaths, 1) * 100)
```

## Summary statistics: 12 June 2025 – 8 May 2026

```{r terbufos-summary-table}
#| label: tbl-terbufos-summary
#| tbl-cap: !expr 'paste0("Agricultural and stock remedy poisoning burden during the post-Cabinet-recommendation window (12 Jun 2025 – 08 May 2026) compared with the equivalent prior-year window (", format(pre_start, \"%d %b %Y\"), " – ", format(pre_end, \"%d %b %Y\"), ").")'

tibble(
  Indicator = c(
    "Period start", "Period end", "Duration (days)",
    "Total notifications", "Total deaths",
    "Case fatality ratio (CFR)",
    "Weekly notification rate (mean ± SD)",
    "Top province by notifications",
    "Children <12 (notifications)",
    "Children <12 (deaths)"
  ),
  `Post-recommendation window\n(12 Jun 2025 – 8 May 2026)` = c(
    format(ban_recommend, "%d %b %Y"),
    format(ban_gazette,   "%d %b %Y"),
    as.character(window_days),
    scales::comma(ban_n),
    scales::comma(ban_deaths),
    paste0(ban_cfr, "%"),
    {
      wk <- poison_ban %>%
        mutate(wk = floor_date(nd, "week")) %>%
        count(wk)
      paste0(round(mean(wk$n), 1), " ± ", round(sd(wk$n), 1))
    },
    {
      poison_ban %>% count(Province, sort = TRUE) %>%
        slice(1) %>% pull(Province) %>% as.character()
    },
    scales::comma(sum(poison_ban$Age_years < 12, na.rm = TRUE)),
    scales::comma(sum(poison_ban$deceased[poison_ban$Age_years < 12], na.rm = TRUE))
  ),
  `Equivalent prior-year window\n(12 Jun 2024 – 8 May 2025)` = c(
    format(pre_start, "%d %b %Y"),
    format(pre_end,   "%d %b %Y"),
    as.character(window_days),
    scales::comma(pre_n),
    scales::comma(pre_deaths),
    paste0(pre_cfr, "%"),
    {
      wk <- poison_pre %>%
        mutate(wk = floor_date(nd, "week")) %>%
        count(wk)
      paste0(round(mean(wk$n), 1), " ± ", round(sd(wk$n), 1))
    },
    {
      poison_pre %>% count(Province, sort = TRUE) %>%
        slice(1) %>% pull(Province) %>% as.character()
    },
    scales::comma(sum(poison_pre$Age_years < 12, na.rm = TRUE)),
    scales::comma(sum(poison_pre$deceased[poison_pre$Age_years < 12], na.rm = TRUE))
  ),
  `% Change` = c(
    "—", "—", "—",
    paste0(ifelse(pct_chg_n >= 0, "+", ""), pct_chg_n, "%"),
    paste0(ifelse(pct_chg_d >= 0, "+", ""), pct_chg_d, "%"),
    paste0(ifelse(ban_cfr - pre_cfr >= 0, "+", ""), round(ban_cfr - pre_cfr, 1), " pp"),
    "—", "—", "—", "—"
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = 4:6, j = 1) %>%
  bg(i = 4:6, bg = "#FFF3E0") %>%
  color(i = 4, j = 4,
        color = if (pct_chg_n > 10) "#C62828" else if (pct_chg_n < -10) "#2E7D32" else "#F57C00") %>%
  color(i = 5, j = 4,
        color = if (pct_chg_d > 10) "#C62828" else if (pct_chg_d < -10) "#2E7D32" else "#F57C00") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 10, part = "all") %>%
  add_footer_lines("pp = percentage points. CFR calculated as deaths ÷ total notifications × 100. Children <12 based on recorded age in years.")
```

## Weekly trend within the policy window

```{r terbufos-weekly-trend}
#| label: fig-terbufos-weekly
#| fig-cap: "Weekly notifications and deaths for agricultural and stock remedy poisoning during the post-Cabinet-recommendation period (12 June 2025 – 8 May 2026). The dashed vertical line marks the gazetted ban date (8 May 2026)."
#| fig-height: 4.5

poison_ban %>%
  mutate(week = floor_date(nd, "week")) %>%
  group_by(week) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") -> ban_weekly

ggplot(ban_weekly, aes(x = week)) +
  geom_col(aes(y = n),      fill = "#90CAF9", colour = "#64B5F6", width = 5.5, alpha = 0.9) +
  geom_col(aes(y = deaths), fill = "#EF9A9A", colour = "#E57373", width = 5.5, alpha = 0.9) +
  geom_line(aes(y = zoo::rollmean(n, k = 4, fill = NA, align = "right")),
            colour = "#1565C0", linewidth = 1.1, na.rm = TRUE) +
  geom_vline(xintercept = as.numeric(ban_gazette),
             linetype = "dashed", colour = "#B71C1C", linewidth = 0.8) +
  annotate("text", x = ban_gazette + days(3),
           y = max(ban_weekly$n, na.rm = TRUE) * 0.92,
           label = "Gazetted ban\n8 May 2026",
           colour = "#B71C1C", size = 3, hjust = 0) +
  scale_x_date(date_labels = "%d %b\n%Y", date_breaks = "4 weeks",
               expand = expansion(mult = c(0.01, 0.05))) +
  scale_y_continuous(name = "Notifications / deaths (n)", breaks = pretty_breaks(5)) +
  labs(x = "Week start",
       caption = "Blue bars = all notifications; red bars = deaths; blue line = 4-week rolling average.") +
  theme_nmc() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
```

## Provincial breakdown during the policy window

```{r terbufos-prov-breakdown}
#| label: tbl-terbufos-prov
#| tbl-cap: "Notifications, deaths, and CFR by province during the post-Cabinet-recommendation period (12 June 2025 – 8 May 2026)."

poison_ban %>%
  group_by(Province) %>%
  summarise(
    Notifications = n(),
    Deaths        = sum(deceased, na.rm = TRUE),
    CFR           = round(Deaths / Notifications * 100, 1),
    `Children <12 (n)` = sum(Age_years < 12, na.rm = TRUE),
    `Median age (years)` = round(median(Age_years, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(Notifications)) %>%
  adorn_totals("row", fill = "—", na.rm = TRUE,
               cols = c(Notifications, Deaths, `Children <12 (n)`)) %>%
  mutate(CFR = round(Deaths / Notifications * 100, 1)) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = ~ Province == "Total") %>%
  bg(i = ~ Province == "Total", bg = "#E3F2FD") %>%
  bg(i = ~ CFR > 10 & Province != "Total", j = "CFR", bg = "#FFCDD2") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 10, part = "all") %>%
  add_footer_lines("Pink = CFR > 10%. CFR not calculated for total row from adorn_totals; recalculated from column sums.")
```

---

# Case Fatality Rate: Detailed Profile {#sec-cfr-detail}

```{r cfr-detail-setup}
#| include: false

# Overall CFR across full dataset
cfr_overall <- round(sum(poison$deceased, na.rm = TRUE) / nrow(poison) * 100, 1)

# CFR with exact 95% CI using Wilson method
cfr_ci <- binom::binom.wilson(
  x = sum(poison$deceased, na.rm = TRUE),
  n = nrow(poison)
)
cfr_lo <- round(cfr_ci$lower * 100, 1)
cfr_hi <- round(cfr_ci$upper * 100, 1)
```

::: {.callout-note appearance="minimal"}
Overall CFR for the full surveillance period (2020–present): **`r cfr_overall`%** (95% CI: `r cfr_lo`–`r cfr_hi`%).
:::

## Annual CFR with confidence intervals

```{r cfr-annual-ci}
#| label: fig-cfr-annual-ci
#| fig-cap: "Annual case fatality ratio (%) with 95% Wilson confidence intervals, agricultural and stock remedy poisoning, South Africa, 2020–present. Numbers above bars = annual notification count."
#| fig-height: 4.5

poison %>%
  group_by(Year_notification) %>%
  summarise(
    n      = n(),
    deaths = sum(deceased, na.rm = TRUE),
    cfr    = deaths / n * 100,
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    lo = binom::binom.wilson(deaths, n)$lower * 100,
    hi = binom::binom.wilson(deaths, n)$upper * 100
  ) %>%
  ungroup() -> cfr_annual_ci

ggplot(cfr_annual_ci, aes(x = factor(Year_notification), y = cfr)) +
  geom_col(fill = "#EF9A9A", colour = "#E57373", width = 0.65, alpha = 0.9) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.25,
                colour = "#B71C1C", linewidth = 0.8) +
  geom_text(aes(y = hi + 0.3, label = scales::comma(n)),
            size = 3.2, vjust = 0, colour = "#333333") +
  scale_y_continuous(name = "Case fatality ratio (%)", expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Year",
       caption = "Error bars = 95% Wilson confidence intervals. Numbers = total notifications per year.") +
  theme_nmc()
```

## CFR by province and year — heatmap

```{r cfr-prov-heatmap}
#| label: fig-cfr-prov-heatmap
#| fig-cap: "Case fatality ratio (%) by province and year, agricultural and stock remedy poisoning, South Africa. Cell text = CFR (notifications count in parentheses). Grey = no data."
#| fig-height: 5.5

poison %>%
  group_by(Province, Year_notification) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = round(deaths / n * 100, 1), .groups = "drop") %>%
  complete(Province, Year_notification, fill = list(n = NA_real_, deaths = NA_real_, cfr = NA_real_)) -> cfr_heat_df

ggplot(cfr_heat_df, aes(x = factor(Year_notification), y = Province, fill = cfr)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(!is.na(cfr),
                               paste0(cfr, "%\n(n=", scales::comma(n), ")"), "—")),
            size = 2.6, lineheight = 0.85) +
  scale_fill_gradient(low = "#FFF9C4", high = "#B71C1C", na.value = "grey88",
                      name = "CFR (%)") +
  labs(x = "Year", y = NULL) +
  theme_nmc() +
  theme(legend.position = "right", panel.grid = element_blank())
```

## CFR by age group — forest plot

```{r cfr-age-forest}
#| label: fig-cfr-age-forest
#| fig-cap: "Case fatality ratio by age group with 95% Wilson confidence intervals, agricultural and stock remedy poisoning, South Africa, 2020–present. Reference line = overall CFR."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group)) %>%
  group_by(age_group) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") %>%
  rowwise() %>%
  mutate(
    cfr = binom::binom.wilson(deaths, n)$mean  * 100,
    lo  = binom::binom.wilson(deaths, n)$lower * 100,
    hi  = binom::binom.wilson(deaths, n)$upper * 100
  ) %>%
  ungroup() -> cfr_age_df

ggplot(cfr_age_df, aes(x = cfr, y = fct_rev(age_group))) +
  geom_vline(xintercept = cfr_overall, linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.3,
                 colour = "#B71C1C", linewidth = 0.8) +
  geom_point(aes(size = n), colour = "#B71C1C", shape = 18) +
  geom_text(aes(x = hi + 0.3, label = paste0(round(cfr, 1), "%")),
            hjust = 0, size = 3, colour = "#333333") +
  scale_size_continuous(range = c(2, 8), name = "Notifications (n)") +
  scale_x_continuous(name = "Case fatality ratio (%)",
                     expand = expansion(mult = c(0, 0.15))) +
  annotate("text", x = cfr_overall + 0.1,
           y = nrow(cfr_age_df) + 0.5, label = paste0("Overall CFR = ", cfr_overall, "%"),
           hjust = 0, size = 3, colour = "grey40") +
  labs(y = "Age group (years)") +
  theme_nmc() +
  theme(legend.position = "bottom",
        panel.grid.major.x = element_line(colour = "#EEEEEE", linewidth = 0.3))
```

## CFR summary table by key strata

```{r cfr-strata-table}
#| label: tbl-cfr-strata
#| tbl-cap: "Case fatality ratio (%) by key demographic and geographic strata, agricultural and stock remedy poisoning, South Africa, 2020–present."

bind_rows(
  # Overall
  poison %>%
    summarise(Stratum = "Overall", n = n(),
              deaths = sum(deceased, na.rm = TRUE)) %>%
    mutate(group = "Overall"),
  # By sex
  poison %>% filter(!is.na(sex)) %>%
    group_by(Stratum = sex) %>%
    summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") %>%
    mutate(group = "Sex"),
  # By age group
  poison %>% filter(!is.na(age_group)) %>%
    group_by(Stratum = as.character(age_group)) %>%
    summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") %>%
    mutate(group = "Age group"),
  # By province
  poison %>%
    group_by(Stratum = as.character(Province)) %>%
    summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") %>%
    mutate(group = "Province")
) %>%
  rowwise() %>%
  mutate(
    CFR = round(binom::binom.wilson(deaths, n)$mean  * 100, 1),
    Lo  = round(binom::binom.wilson(deaths, n)$lower * 100, 1),
    Hi  = round(binom::binom.wilson(deaths, n)$upper * 100, 1)
  ) %>%
  ungroup() %>%
  mutate(`CFR (95% CI)` = paste0(CFR, "% (", Lo, "–", Hi, "%)")) %>%
  select(group, Stratum, `Notifications (n)` = n, `Deaths (n)` = deaths, `CFR (95% CI)`) %>%
  flextable() %>%
  merge_v(j = "group") %>%
  bold(part = "header") %>%
  bold(i = ~ group == "Overall") %>%
  bg(i = ~ group == "Overall", bg = "#E3F2FD") %>%
  theme_vanilla() %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 10, part = "all") %>%
  add_footer_lines("95% CI calculated using the Wilson score method.")
```

---

# Facility Profile {#sec-facility-profile}

::: {.callout-note appearance="minimal"}
This section describes **which types of health facilities** are reporting agricultural and stock remedy poisoning notifications, indicating care-seeking pathways and potential surveillance gaps.
:::

```{r facility-setup}
#| include: false

# Standardise facility_type to title case and collapse rare types to "Other"
poison %>%
  filter(!is.na(facility_type)) %>%
  mutate(
    fac_type = str_to_title(str_trim(facility_type)),
    fac_type = fct_lump_n(fac_type, n = 8, other_level = "Other")
  ) -> poison_fac

poison %>%
  filter(!is.na(facility_sector)) %>%
  mutate(fac_sector = str_to_title(str_trim(facility_sector))) -> poison_sec
```

## Notifications by facility type

```{r facility-type-bar}
#| label: fig-facility-type
#| fig-cap: "Number of agricultural and stock remedy poisoning notifications by reporting facility type, South Africa, 2020–present. Facility types that account for <1% combined are grouped as 'Other'."
#| fig-height: 5

poison_fac %>%
  count(fac_type, sort = TRUE) %>%
  mutate(
    pct      = n / sum(n) * 100,
    fac_type = fct_reorder(fac_type, n)
  ) -> fac_type_df

ggplot(fac_type_df, aes(x = n, y = fac_type)) +
  geom_col(fill = "#5C6BC0", alpha = 0.88, width = 0.7) +
  geom_text(aes(label = paste0(scales::comma(n), " (", round(pct, 1), "%)")),
            hjust = -0.08, size = 3.1, colour = "#333333") +
  scale_x_continuous(name = "Notifications (n)",
                     expand = expansion(mult = c(0, 0.2)),
                     labels = scales::comma) +
  labs(y = NULL) +
  theme_nmc() +
  theme(panel.grid.major.x = element_line(colour = "#EEEEEE", linewidth = 0.3))
```

## Facility type by vital status

```{r facility-vital}
#| label: fig-facility-vital
#| fig-cap: "Proportion of cases alive vs deceased by reporting facility type, agricultural and stock remedy poisoning, South Africa, 2020–present. Facilities with ≥10 notifications included. Red = deceased."
#| fig-height: 5.5

poison_fac %>%
  filter(!is.na(patient_vital_status),
         patient_vital_status %in% c("Alive", "Deceased")) %>%
  group_by(fac_type) %>%
  filter(n() >= 10) %>%
  ungroup() %>%
  count(fac_type, patient_vital_status) %>%
  group_by(fac_type) %>%
  mutate(
    total = sum(n),
    pct   = n / total * 100,
    fac_type = fct_reorder(fac_type, total)
  ) %>%
  ungroup() -> fac_vital_df

ggplot(fac_vital_df, aes(x = pct, y = fac_type, fill = patient_vital_status)) +
  geom_col(position = "fill", width = 0.7) +
  geom_text(data = fac_vital_df %>% filter(patient_vital_status == "Deceased"),
            aes(x = pct / 100, label = paste0(round(pct, 1), "%")),
            hjust = 1.1, size = 3, colour = "white", fontface = "bold") +
  scale_x_continuous(labels = scales::percent_format(scale = 100),
                     name = "Proportion of notifications (%)") +
  scale_fill_manual(values = c("Alive" = "#81C784", "Deceased" = "#E57373"), name = NULL) +
  labs(y = NULL) +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## CFR by facility type and sector

```{r facility-cfr-table}
#| label: tbl-facility-cfr
#| tbl-cap: "Notifications, deaths, and case fatality ratio by facility type, agricultural and stock remedy poisoning, South Africa, 2020–present."

poison_fac %>%
  group_by(`Facility Type` = fac_type) %>%
  summarise(
    Notifications = n(),
    Deaths        = sum(deceased, na.rm = TRUE),
    CFR           = round(Deaths / Notifications * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(Notifications)) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ CFR > 10, j = "CFR", bg = "#FFCDD2") %>%
  bg(i = ~ CFR <= 10 & CFR > 5, j = "CFR", bg = "#FFF9C4") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 10, part = "all") %>%
  add_footer_lines("Pink = CFR >10%; Yellow = CFR 5–10%.")
```

## Facility type trend over years

```{r facility-type-trend}
#| label: fig-facility-type-trend
#| fig-cap: "Annual proportion of notifications by reporting facility type, agricultural and stock remedy poisoning, South Africa. Shift in facility mix may reflect changes in care pathways or surveillance coverage."
#| fig-height: 5

poison_fac %>%
  count(Year_notification, fac_type) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  ggplot(aes(x = factor(Year_notification), y = pct, fill = fac_type)) +
  geom_col(position = "fill", width = 0.75) +
  scale_y_continuous(labels = scales::percent_format(scale = 100),
                     name = "Proportion of notifications (%)") +
  scale_fill_brewer(palette = "Set2", name = "Facility type") +
  labs(x = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))
```

## Public vs private sector split

```{r facility-sector-bar}
#| label: fig-facility-sector
#| fig-cap: "Notifications by facility sector (public vs private) and year, agricultural and stock remedy poisoning, South Africa."
#| fig-height: 4.5

poison %>%
  filter(!is.na(facility_sector)) %>%
  mutate(Sector = str_to_title(str_trim(facility_sector))) %>%
  count(Year_notification, Sector) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  ggplot(aes(x = factor(Year_notification), y = pct, fill = Sector)) +
  geom_col(position = "fill", width = 0.75) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_fill(vjust = 0.5),
            size = 3, colour = "white", fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(scale = 100),
                     name = "Proportion of notifications (%)") +
  scale_fill_manual(values = c("Public" = "#1565C0", "Private" = "#2E7D32",
                               "Unknown" = "grey70"), name = "Sector") +
  labs(x = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

---

# Odds of Death: ASRP vs Other Notifiable Conditions {#sec-odds-death}

::: {.callout-note appearance="minimal"}
Logistic regression comparing the odds of dying from agricultural and stock remedy poisoning versus all other notifiable conditions reported in the same period, adjusted for age group, sex, province, and year.
:::

```{r odds-setup}
#| label: odds-setup
#| include: false

# Load full master dataset for comparison conditions
new_master %>%
  filter(
    as.integer(Year_notification) %in% 2020:year(Sys.Date()),
    !is.na(prov_), !is.na(patient_vital_status)
  ) %>%
  mutate(
    deceased_bin = as.integer(patient_vital_status %in% "Deceased"),
    is_asrp      = as.integer(condition %in% CONDITION),
    Age_years    = as.numeric(Age_years),
    age_group_lr = cut(
      Age_years,
      breaks = c(-Inf, 5, 15, 30, 45, 60, Inf),
      labels = c("<5", "5–14", "15–29", "30–44", "45–59", "60+"),
      right  = FALSE
    ),
    sex_lr = case_when(
      patient_gender %in% c("Male",   "M") ~ "Male",
      patient_gender %in% c("Female", "F") ~ "Female",
      TRUE ~ NA_character_
    ),
    year_lr = as.factor(as.integer(Year_notification))
  ) %>%
  filter(!is.na(sex_lr), !is.na(age_group_lr), !is.na(prov_)) -> lr_data

# Logistic regression: deceased ~ is_asrp + age_group + sex + prov_ + year
lr_model <- glm(
  deceased_bin ~ is_asrp + age_group_lr + sex_lr + prov_ + year_lr,
  data   = lr_data,
  family = binomial(link = "logit")
)

# Extract OR and CI for is_asrp
lr_tidy <- broom::tidy(lr_model, exponentiate = TRUE, conf.int = TRUE)

asrp_or <- lr_tidy %>%
  filter(term == "is_asrp") %>%
  mutate(
    OR   = round(estimate,   2),
    lo95 = round(conf.low,   2),
    hi95 = round(conf.high,  2),
    p    = round(p.value,    4)
  )
```

## Crude vs adjusted odds of dying

```{r odds-table}
#| label: tbl-odds-asrp
#| tbl-cap: "Crude and adjusted odds of dying from agricultural and stock remedy poisoning compared with all other notifiable conditions, South Africa, 2020–present. Adjusted for age group, sex, province, and year."

# Crude OR
crude_model <- glm(deceased_bin ~ is_asrp, data = lr_data, family = binomial())
crude_tidy  <- broom::tidy(crude_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "is_asrp")

# All model terms formatted
lr_tidy %>%
  mutate(
    OR   = round(estimate,  2),
    lo95 = round(conf.low,  2),
    hi95 = round(conf.high, 2),
    p    = ifelse(p.value < 0.001, "<0.001", round(p.value, 3))
  ) %>%
  mutate(
    `OR (95% CI)` = paste0(OR, " (", lo95, "–", hi95, ")"),
    Predictor     = case_when(
      term == "is_asrp"                   ~ "ASRP vs other conditions",
      str_starts(term, "age_group_lr")    ~ paste0("Age group: ", str_remove(term, "age_group_lr")),
      str_starts(term, "sex_lr")          ~ paste0("Sex: ",       str_remove(term, "sex_lr")),
      str_starts(term, "prov_")           ~ paste0("Province: ",  str_remove(term, "prov_")),
      str_starts(term, "year_lr")         ~ paste0("Year: ",      str_remove(term, "year_lr")),
      term == "(Intercept)"               ~ "Intercept",
      TRUE ~ term
    )
  ) %>%
  filter(term != "(Intercept)") %>%
  select(Predictor, OR, `95% CI lower` = lo95, `95% CI upper` = hi95, `p-value` = p) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = ~ Predictor == "ASRP vs other conditions") %>%
  bg(i   = ~ Predictor == "ASRP vs other conditions", bg = "#FFF3E0") %>%
  color(i = ~ OR > 3 & Predictor != "Intercept", j = "OR", color = "#C62828") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 10, part = "all") %>%
  add_footer_lines(paste0(
    "Adjusted OR for ASRP vs other conditions: ",
    asrp_or$OR, " (95% CI: ", asrp_or$lo95, "–", asrp_or$hi95,
    ", p ", ifelse(asrp_or$p < 0.001, "<0.001", asrp_or$p), "). ",
    "Reference categories: age <5, Female, EC province, earliest year."
  ))
```

## Forest plot of adjusted odds ratios

```{r odds-forest}
#| label: fig-odds-forest
#| fig-cap: "Adjusted odds ratios (OR) for death from logistic regression model. Reference line at OR = 1.0. ASRP vs other conditions highlighted in red."
#| fig-height: 8

lr_tidy %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR    = round(estimate,  2),
    lo95  = round(conf.low,  2),
    hi95  = round(conf.high, 2),
    label = case_when(
      term == "is_asrp"                   ~ "ASRP vs other conditions",
      str_starts(term, "age_group_lr")    ~ paste0("Age: ",     str_remove(term, "age_group_lr")),
      str_starts(term, "sex_lr")          ~ paste0("Sex: ",     str_remove(term, "sex_lr")),
      str_starts(term, "prov_")           ~ paste0("Province: ",str_remove(term, "prov_")),
      str_starts(term, "year_lr")         ~ paste0("Year: ",    str_remove(term, "year_lr")),
      TRUE ~ term
    ),
    is_asrp_term = term == "is_asrp"
  ) %>%
  mutate(label = fct_reorder(label, OR)) -> forest_df

ggplot(forest_df, aes(x = OR, y = label,
                      colour = is_asrp_term,
                      size   = is_asrp_term)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = lo95, xmax = hi95),
                 height = 0.3, linewidth = 0.8) +
  geom_point(shape = 18) +
  geom_text(aes(x = hi95 + 0.05,
                label = paste0(OR, " (", lo95, "–", hi95, ")")),
            hjust = 0, size = 2.8, colour = "#333333") +
  scale_colour_manual(values = c("TRUE" = "#B71C1C", "FALSE" = "#455A64"),
                      guide = "none") +
  scale_size_manual(values = c("TRUE" = 5, "FALSE" = 3), guide = "none") +
  scale_x_continuous(name = "Adjusted odds ratio (OR)",
                     expand = expansion(mult = c(0.02, 0.3))) +
  labs(y = NULL,
       caption = "Reference: age <5, Female, Eastern Cape, earliest year in dataset.") +
  theme_nmc() +
  theme(panel.grid.major.x = element_line(colour = "#EEEEEE", linewidth = 0.3),
        axis.text.y = element_text(size = 8))
```

## CFR comparison across all notifiable conditions

```{r cfr-all-conditions}
#| label: fig-cfr-conditions
#| fig-cap: "Case fatality ratio (%) for all notifiable conditions with ≥50 notifications, South Africa, 2020–present. Agricultural and stock remedy poisoning highlighted in red. Points sized by notification count."
#| fig-height: 8

new_master %>%
  filter(
    as.integer(Year_notification) %in% 2020:year(Sys.Date()),
    !is.na(patient_vital_status)
  ) %>%
  group_by(condition) %>%
  summarise(
    n      = n(),
    deaths = sum(patient_vital_status %in% "Deceased", na.rm = TRUE),
    cfr    = deaths / n * 100,
    .groups = "drop"
  ) %>%
  filter(n >= 50) %>%
  mutate(
    is_asrp   = condition %in% CONDITION,
    condition = fct_reorder(condition, cfr)
  ) -> cfr_all_df

ggplot(cfr_all_df, aes(x = cfr, y = condition,
                        colour = is_asrp, size = n)) +
  geom_vline(xintercept = weighted.mean(cfr_all_df$cfr, cfr_all_df$n),
             linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_point(alpha = 0.85, shape = 16) +
  geom_text(data = cfr_all_df %>% filter(is_asrp),
            aes(label = paste0(round(cfr, 1), "% (n=", scales::comma(n), ")")),
            hjust = -0.15, size = 3.2, colour = "#B71C1C", fontface = "bold") +
  scale_colour_manual(values = c("TRUE" = "#B71C1C", "FALSE" = "#607D8B"),
                      guide = "none") +
  scale_size_continuous(range = c(2, 10), name = "Notifications (n)",
                        labels = scales::comma) +
  scale_x_continuous(name = "Case fatality ratio (%)",
                     expand = expansion(mult = c(0.01, 0.25))) +
  labs(y = NULL,
       caption = "Dashed line = weighted mean CFR across all conditions. Red = agricultural and stock remedy poisoning.") +
  theme_nmc() +
  theme(
    axis.text.y      = element_text(size = 7.5),
    legend.position  = "bottom",
    panel.grid.major.x = element_line(colour = "#EEEEEE", linewidth = 0.3)
  )
```

---

# Additional Epidemiological Analyses {#sec-additional-epi}

## Time from notification to death

```{r time-to-death}
#| label: fig-time-to-death
#| fig-cap: "Distribution of days between date of notification and date of death for agricultural and stock remedy poisoning fatalities, South Africa, 2020–present. Cases with negative values (death recorded before notification) are excluded. Median and IQR annotated."
#| fig-height: 4.5

poison %>%
  filter(deceased, !is.na(notification_date), !is.na(patient_death_date)) %>%
  mutate(
    nd  = as_date(notification_date),
    dod = as_date(patient_death_date),
    days_to_death = as.integer(dod - nd)
  ) %>%
  filter(days_to_death >= 0, days_to_death <= 60) -> ttd_df

if (nrow(ttd_df) >= 10) {
  med_ttd <- median(ttd_df$days_to_death, na.rm = TRUE)
  iqr_ttd <- IQR(ttd_df$days_to_death,   na.rm = TRUE)

  ggplot(ttd_df, aes(x = days_to_death)) +
    geom_histogram(binwidth = 1, fill = "#EF9A9A", colour = "#E57373", alpha = 0.9) +
    geom_vline(xintercept = med_ttd, linetype = "dashed",
               colour = "#B71C1C", linewidth = 0.9) +
    annotate("text", x = med_ttd + 1,
             y = max(table(ttd_df$days_to_death)) * 0.9,
             label = paste0("Median = ", med_ttd, " days\nIQR = ", round(iqr_ttd, 1), " days"),
             hjust = 0, size = 3.2, colour = "#B71C1C") +
    scale_x_continuous(name = "Days from notification to death",
                       breaks = pretty_breaks(8)) +
    scale_y_continuous(name = "Deaths (n)", breaks = pretty_breaks(5)) +
    labs(caption = "Limited to 0–60 days. Dashed line = median.") +
    theme_nmc()
} else {
  cat("Insufficient date-of-death data available to generate this plot.")
}
```

## Notification lag: date of illness onset to notification date

```{r notification-lag}
#| label: fig-notif-lag
#| fig-cap: "Distribution of days between reported date of illness onset and date of notification, agricultural and stock remedy poisoning, South Africa, 2020–present. Longer lags indicate delays in surveillance detection."
#| fig-height: 4.5

poison %>%
  filter(!is.na(notification_date), !is.na(symptom_date)) %>%
  mutate(
    nd    = as_date(notification_date),
    onset = as_date(symptom_date),
    lag   = as.integer(nd - onset)
  ) %>%
  filter(lag >= 0, lag <= 90) -> lag_df

if (nrow(lag_df) >= 10) {
  med_lag <- median(lag_df$lag, na.rm = TRUE)
  iqr_lag <- IQR(lag_df$lag,   na.rm = TRUE)

  ggplot(lag_df, aes(x = lag)) +
    geom_histogram(binwidth = 1, fill = "#90CAF9", colour = "#64B5F6", alpha = 0.9) +
    geom_vline(xintercept = med_lag, linetype = "dashed",
               colour = "#1565C0", linewidth = 0.9) +
    annotate("text", x = med_lag + 2,
             y = max(table(lag_df$lag)) * 0.88,
             label = paste0("Median = ", med_lag, " days\nIQR = ", round(iqr_lag, 1), " days"),
             hjust = 0, size = 3.2, colour = "#1565C0") +
    scale_x_continuous(name = "Days from onset to notification",
                       breaks = pretty_breaks(8)) +
    scale_y_continuous(name = "Notifications (n)", breaks = pretty_breaks(5)) +
    labs(caption = "Limited to 0–90 days.") +
    theme_nmc()
} else {
  cat("Insufficient onset-date data available to generate this plot.")
}
```

## Missing data profile

```{r missing-data}
#| label: tbl-missing
#| tbl-cap: "Completeness of key surveillance variables for agricultural and stock remedy poisoning notifications, South Africa, 2020–present. Stratified by year."

vars_of_interest <- c(
  "Age_years"            = "Age (years)",
  "sex"                  = "Sex",
  "notification_date"    = "Notification date",
  "symptom_date"         = "Onset date",
  "patient_death_date"   = "Date of death (deaths only)",
  "patient_vital_status" = "Vital status",
  "district"             = "District",
  "facility"             = "Reporting facility",
  "facility_type"        = "Facility type"
)

missing_wide <- purrr::map_dfr(names(vars_of_interest), function(v) {
  poison %>%
    group_by(Year = Year_notification) %>%
    summarise(
      Variable    = vars_of_interest[[v]],
      Total       = n(),
      `Complete`  = sum(!is.na(.data[[v]]) & .data[[v]] != ""),
      .groups = "drop"
    ) %>%
    mutate(`% Complete` = round(`Complete` / Total * 100, 1))
}) %>%
  select(Variable, Year, Total, `Complete`, `% Complete`) %>%
  pivot_wider(names_from = Year,
              values_from = c(`Complete`, `% Complete`),
              names_glue  = "{Year}_{.value}")

# pre-compute which rows have any year with completeness < 70%
pct_cols <- grep("% Complete", names(missing_wide), value = TRUE)
low_rows <- which(rowSums(missing_wide[pct_cols] < 70, na.rm = TRUE) > 0)

ft <- flextable(missing_wide) %>%
  bold(part = "header") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 9, part = "all") %>%
  add_footer_lines("Orange text = at least one year with completeness <70%. Date of death evaluated across all cases (not only deaths).")

if (length(low_rows) > 0) ft <- color(ft, i = low_rows, color = "#E65100")
ft
```

## Repeat reporting facilities — top contributors

```{r top-facilities-table}
#| label: tbl-top-facilities
#| tbl-cap: "Top 20 reporting facilities by total notification count, agricultural and stock remedy poisoning, South Africa, 2020–present."

poison %>%
  filter(!is.na(facility), facility != "") %>%
  group_by(facility, Province) %>%
  summarise(
    Notifications  = n(),
    Deaths         = sum(deceased, na.rm = TRUE),
    CFR            = round(Deaths / Notifications * 100, 1),
    `Years active` = n_distinct(Year_notification),
    `Facility type` = first(facility_type),
    .groups = "drop"
  ) %>%
  arrange(desc(Notifications)) %>%
  slice(1:20) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ CFR > 10, j = "CFR", bg = "#FFCDD2") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 9, part = "all") %>%
  add_footer_lines("Pink = CFR > 10%.")
```

## Annual CFR trend with policy annotations

```{r cfr-policy-timeline}
#| label: fig-cfr-policy
#| fig-cap: "Monthly case fatality ratio (%) for agricultural and stock remedy poisoning, 2020–present, with key policy milestones annotated. Shaded band = 95% Wilson confidence interval."
#| fig-height: 5

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  group_by(month_date) %>%
  summarise(
    n      = n(),
    deaths = sum(deceased, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n >= 5) %>%
  rowwise() %>%
  mutate(
    cfr = binom::binom.wilson(deaths, n)$mean  * 100,
    lo  = binom::binom.wilson(deaths, n)$lower * 100,
    hi  = binom::binom.wilson(deaths, n)$upper * 100
  ) %>%
  ungroup() -> cfr_monthly_df

policy_events <- tibble(
  date  = c(as_date("2025-06-12"), as_date("2026-05-08")),
  label = c("Cabinet ban\nrecommendation", "Gazetted ban"),
  y     = c(max(cfr_monthly_df$hi, na.rm = TRUE) * 0.85,
            max(cfr_monthly_df$hi, na.rm = TRUE) * 0.70)
)

ggplot(cfr_monthly_df, aes(x = month_date)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#FFCDD2", alpha = 0.4) +
  geom_line(aes(y = cfr), colour = "#B71C1C", linewidth = 1.1) +
  geom_point(aes(y = cfr, size = n), colour = "#B71C1C", alpha = 0.8) +
  geom_vline(data = policy_events, aes(xintercept = as.numeric(date)),
             linetype = "dashed", colour = "#1A237E", linewidth = 0.7) +
  geom_label(data = policy_events,
             aes(x = date, y = y, label = label),
             size = 2.8, colour = "#1A237E", fill = "white",
             label.size = 0.2, hjust = 0, nudge_x = 15) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months",
               expand = expansion(add = c(0, 0))) +
  scale_y_continuous(name = "Monthly CFR (%)", breaks = pretty_breaks(5)) +
  scale_size_continuous(range = c(1.5, 5), name = "n", guide = "none") +
  labs(x = "Month",
       caption = "Shaded band = 95% Wilson CI. Only months with ≥5 notifications shown.") +
  theme_nmc()
```

---

*Report generated: `r format(Sys.Date(), "%d %B %Y")` | NMC Surveillance Unit*
