---
title: "Agricultural & Stock Remedy Poisoning"
subtitle: "Epidemiological Description, South Africa · NMC Surveillance"
date: today
sidebar: special-reports
execute:
  echo: false
  warning: false
  message: false
  error: false
  cache: false
---

::: {.callout-important appearance="minimal"}
**Priority condition** — Agricultural and stock remedy poisoning is a current public health priority in South Africa. This report describes the epidemiology based on NMC notifiable conditions surveillance data, 2020–present.
:::

```{r setup}
#| label: setup
#| include: false
#| cache: false

pacman::p_load(
  ggplot2, gtsummary, tidyverse, magrittr,
  conflicted, scales, data.table, lubridate,
  flextable, gt, ggrepel, zoo,
  janitor
)

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::lag)
conflicted::conflicts_prefer(tidyr::extract)
conflicted::conflicts_prefer(dplyr::first)
conflicted::conflicts_prefer(lubridate::year)
conflicted::conflicts_prefer(lubridate::month)
conflicted::conflicts_prefer(lubridate::quarter)

library(NMCleaner)

# ── Load master dataset ───────────────────────────────────────────────────────


new_master <- arrow::read_feather("~/Desktop/SAFETP/CLA/NMC_database/master/new_master.feather")

# ── Filter to Agricultural poisoning only ────────────────────────────────────
CONDITION <- "Agricultural or stock remedy poisoning"

new_master %>%
  filter(
    condition %in% CONDITION,
    as.integer(Year_notification) %in% 2020:year(Sys.Date()),
    !is.na(prov_)
  ) %>%
  mutate(
    Year_notification = as.integer(Year_notification),
    patient_vital_status = if_else(
      patient_vital_status %like% "Unknown", "Alive", patient_vital_status
    ),
    Age_years = as.numeric(Age_years),
    age_group = cut(
      Age_years,
      breaks = c(-Inf, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, Inf),
      labels = c("<1", "1\u20134", "5\u20139", "10\u201314", "15\u201319",
                 "20\u201324", "25\u201329", "30\u201334", "35\u201339",
                 "40\u201344", "45\u201349", "50\u201354", "55\u201359",
                 "60\u201364", "65+"),
      right  = FALSE
    ),
    sex = case_when(
      patient_gender %in% c("Male", "M")   ~ "Male",
      patient_gender %in% c("Female", "F") ~ "Female",
      TRUE                                  ~ NA_character_
    ),
    deceased = patient_vital_status %in% "Deceased"
  ) -> poison

# ── Shared theme ──────────────────────────────────────────────────────────────
theme_nmc <- function(...) {
  theme_classic() +
    theme(
      text              = element_text(family = "Century Gothic", size = 10, color = "#333333"),
      panel.background  = element_rect(fill = "#F8F8F8"),
      plot.background   = element_rect(fill = "#F8F8F8"),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      axis.ticks        = element_line(linewidth = 0.15, color = "grey40"),
      legend.background = element_rect(fill = "#F8F8F8", linewidth = 0.25,
                                       linetype = "solid", colour = "grey50"),
      legend.key        = element_blank(),
      strip.background  = element_rect(color = NA, fill = NA),
      ...
    )
}
```

::: {.callout-note appearance="minimal" collapse="true"}
## Report metadata
| | |
|---|---|
| **Condition** | `r CONDITION` |
| **Years** | 2020 – `r lubridate::year(Sys.Date())` |
| **Data source** | NMC Notifiable Medical Conditions surveillance database |
| **Generated** | `r format(Sys.Date(), "%d %B %Y")` |
:::

---

# Situation Update for Policy Makers {#sec-executive}

```{r exec-setup}
#| label: exec-setup
#| include: false

# ── Current year and comparison calculations ──────────────────────────────────
current_year <- year(Sys.Date())
prev_year    <- current_year - 1
current_month <- month(Sys.Date())

# Year-to-date (Jan 1 to current date) for current and previous year
ytd_current <- poison %>%
 filter(Year_notification == current_year,
         month(as_date(notification_date)) <= current_month)
ytd_prev <- poison %>%
  filter(Year_notification == prev_year,
         month(as_date(notification_date)) <= current_month)

n_ytd_current   <- nrow(ytd_current)
n_ytd_prev      <- nrow(ytd_prev)
deaths_ytd_curr <- sum(ytd_current$deceased, na.rm = TRUE)
deaths_ytd_prev <- sum(ytd_prev$deceased, na.rm = TRUE)
cfr_ytd_curr    <- round(deaths_ytd_curr / max(n_ytd_current, 1) * 100, 1)
cfr_ytd_prev    <- round(deaths_ytd_prev / max(n_ytd_prev, 1) * 100, 1)

pct_change_n      <- round((n_ytd_current - n_ytd_prev) / max(n_ytd_prev, 1) * 100)
pct_change_deaths <- round((deaths_ytd_curr - deaths_ytd_prev) / max(deaths_ytd_prev, 1) * 100)

# Direction arrows and colours
arrow_n      <- if (pct_change_n > 0) "↑" else if (pct_change_n < 0) "↓" else "→"
arrow_deaths <- if (pct_change_deaths > 0) "↑" else if (pct_change_deaths < 0) "↓" else "→"
col_n        <- if (pct_change_n > 10) "#C62828" else if (pct_change_n < -10) "#2E7D32" else "#F57C00"
col_deaths   <- if (pct_change_deaths > 10) "#C62828" else if (pct_change_deaths < -10) "#2E7D32" else "#F57C00"

# Children under 5 burden — current and previous year
under5_ytd_curr <- ytd_current %>% filter(Age_years < 5)
under5_ytd_prev <- ytd_prev %>% filter(Age_years < 5)
n_under5_curr   <- nrow(under5_ytd_curr)
n_under5_prev   <- nrow(under5_ytd_prev)
deaths_under5_curr <- sum(under5_ytd_curr$deceased, na.rm = TRUE)
deaths_under5_prev <- sum(under5_ytd_prev$deceased, na.rm = TRUE)
pct_change_under5 <- round((n_under5_curr - n_under5_prev) / max(n_under5_prev, 1) * 100)
pct_change_deaths_under5 <- round((deaths_under5_curr - deaths_under5_prev) / max(deaths_under5_prev, 1) * 100)
arrow_under5 <- if (pct_change_under5 > 0) "↑" else if (pct_change_under5 < 0) "↓" else "→"
arrow_deaths_under5 <- if (pct_change_deaths_under5 > 0) "↑" else if (pct_change_deaths_under5 < 0) "↓" else "→"

# Children under 12 burden — current and previous year
under12_ytd_curr <- ytd_current %>% filter(Age_years < 12)
under12_ytd_prev <- ytd_prev %>% filter(Age_years < 12)
n_under12_curr   <- nrow(under12_ytd_curr)
n_under12_prev   <- nrow(under12_ytd_prev)
deaths_under12_curr <- sum(under12_ytd_curr$deceased, na.rm = TRUE)
deaths_under12_prev <- sum(under12_ytd_prev$deceased, na.rm = TRUE)
pct_change_under12 <- round((n_under12_curr - n_under12_prev) / max(n_under12_prev, 1) * 100)
pct_change_deaths_under12 <- round((deaths_under12_curr - deaths_under12_prev) / max(deaths_under12_prev, 1) * 100)
arrow_under12 <- if (pct_change_under12 > 0) "↑" else if (pct_change_under12 < 0) "↓" else "→"
arrow_deaths_under12 <- if (pct_change_deaths_under12 > 0) "↑" else if (pct_change_deaths_under12 < 0) "↓" else "→"

# Top 3 provinces this year
top3_prov <- ytd_current %>%
  count(prov_, sort = TRUE) %>%
  slice(1:3) %>%
  pull(prov_) %>%
  paste(collapse = ", ")
```

::: {.panel-tabset}

## Key Metrics

```{r exec-metrics}
#| label: tbl-exec-metrics
#| tbl-cap: !expr 'paste0("Year-to-date comparison: Jan–", format(Sys.Date(), "%b %Y"), " vs same period ", prev_year)'

tibble(
  Indicator = c(
    "Notifications (YTD)",
    "Deaths (YTD)",
    "Case Fatality Ratio",
    "Children <5 (notifications)",
    "Children <5 (deaths)",
    "Children <12 (notifications)",
    "Children <12 (deaths)",
    "Top 3 provinces (by case count)"
  ),
  `Current Year` = c(
    scales::comma(n_ytd_current),
    scales::comma(deaths_ytd_curr),
    paste0(cfr_ytd_curr, "%"),
    scales::comma(n_under5_curr),
    scales::comma(deaths_under5_curr),
    scales::comma(n_under12_curr),
    scales::comma(deaths_under12_curr),
    top3_prov
  ),
  `Same Period Last Year` = c(
    scales::comma(n_ytd_prev),
    scales::comma(deaths_ytd_prev),
    paste0(cfr_ytd_prev, "%"),
    scales::comma(n_under5_prev),
    scales::comma(deaths_under5_prev),
    scales::comma(n_under12_prev),
    scales::comma(deaths_under12_prev),
    "—"
  ),
  `Change` = c(
    paste0(arrow_n, " ", ifelse(pct_change_n >= 0, "+", ""), pct_change_n, "%"),
    paste0(arrow_deaths, " ", ifelse(pct_change_deaths >= 0, "+", ""), pct_change_deaths, "%"),
    paste0(ifelse(cfr_ytd_curr >= cfr_ytd_prev, "↑", "↓"), " ",
           ifelse(cfr_ytd_curr - cfr_ytd_prev >= 0, "+", ""),
           round(cfr_ytd_curr - cfr_ytd_prev, 1), " pp"),
    paste0(arrow_under5, " ", ifelse(pct_change_under5 >= 0, "+", ""), pct_change_under5, "%"),
    paste0(arrow_deaths_under5, " ", ifelse(pct_change_deaths_under5 >= 0, "+", ""), pct_change_deaths_under5, "%"),
    paste0(arrow_under12, " ", ifelse(pct_change_under12 >= 0, "+", ""), pct_change_under12, "%"),
    paste0(arrow_deaths_under12, " ", ifelse(pct_change_deaths_under12 >= 0, "+", ""), pct_change_deaths_under12, "%"),
    "—"
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  bold(i = 1:3, j = 1) %>%
  color(i = 1, j = 4, color = col_n) %>%
  color(i = 2, j = 4, color = col_deaths) %>%
  bg(i = 4:7, bg = "#FFF8E1") %>%
  set_table_properties(layout = "autofit", width = 1) %>%
  fontsize(size = 11, part = "all") %>%
  align(j = 2:4, align = "center", part = "all") %>%
  add_footer_lines(paste0("Data extracted: ", format(Sys.Date(), "%d %B %Y"),
                          ". YTD = Year-to-date (Jan 1 – ", format(Sys.Date(), "%d %b"), "). pp = percentage points."))
```

## 12-Month Trend

```{r exec-trend}
#| label: fig-exec-trend
#| fig-cap: "Monthly notification trend — last 12 months with 3-month rolling average. Shaded area highlights the Sep–Nov seasonal peak period."
#| fig-height: 4

# Last 12 complete months
end_date   <- floor_date(Sys.Date(), "month") - days(1)
start_date <- floor_date(end_date - months(11), "month")

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  #filter(month_date >= start_date, month_date <= end_date) %>%
  count(month_date) %>%
  arrange(month_date) %>%
  mutate(
    rolling3 = zoo::rollmean(n, k = 3, fill = NA, align = "right"),
    is_peak  = month(month_date) %in% c(9, 10, 11)  # Sep-Nov peak season
  ) -> trend12_df

# Highlight rectangle for peak season months in this 12-month window
peak_rects <- trend12_df %>%
  filter(is_peak) %>%
  summarise(xmin = min(month_date, na.rm = TRUE) - days(15),
            xmax = max(month_date, na.rm = TRUE) + days(15)) %>%
  slice(1)

ggplot(trend12_df, aes(x = month_date)) +
  {if (nrow(peak_rects) > 0) annotate("rect",
    xmin = peak_rects$xmin, xmax = peak_rects$xmax,
    ymin = -Inf, ymax = Inf, fill = "#FFECB3", alpha = 0.5)} +
  geom_col(aes(y = n), fill = "#90CAF9", colour = "#64B5F6", width = 22) +
  geom_line(aes(y = rolling3), colour = "#C62828", linewidth = 1.3, na.rm = TRUE) +
  geom_point(aes(y = rolling3), colour = "#C62828", size = 2.5, na.rm = TRUE) +
  geom_text(aes(y = n, label = n), vjust = -0.5, size = 3.2, colour = "#333333") +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "1 month", expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(name = "Notifications", expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL,
       caption = "Red line = 3-month rolling average. Yellow shading = Sep–Nov seasonal peak.") +
  theme_nmc() +
  theme(axis.text.x = element_text(size = 8))
```

## Provincial Burden

```{r exec-prov-heatmap}
#| label: fig-exec-prov
#| fig-cap: "Provincial burden heatmap — quarterly notifications, deaths, and incidence per 100,000, last 2 years. Darker = higher incidence."
#| fig-height: 5.5

# Get provincial population for incidence calculation
prov_pop <- NMCleaner::pop %>%
  filter(as.integer(Year) == year(Sys.Date())) %>%
  group_by(prov) %>%
  summarise(population = sum(Population), .groups = "drop")

# Last 8 quarters with incidence
poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(
    nd = as_date(notification_date),
    qtr = paste0(year(nd), " Q", quarter(nd))
  ) %>%
  filter(nd >= (Sys.Date() - years(2))) %>%
  group_by(prov_, qtr) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE), .groups = "drop") %>%
  complete(prov_, qtr, fill = list(n = 0, deaths = 0)) %>%
  left_join(prov_pop, by = c("prov_" = "prov")) %>%
  mutate(
    # Quarterly incidence annualized (multiply by 4) per 100k
    inc = round(n / population * 1e5 * 4, 1)
  ) -> prov_qtr_df

# Order quarters chronologically
qtr_order <- prov_qtr_df %>% distinct(qtr) %>% arrange(qtr) %>% pull(qtr)
prov_qtr_df$qtr <- factor(prov_qtr_df$qtr, levels = qtr_order)

ggplot(prov_qtr_df, aes(x = qtr, y = prov_, fill = inc)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(n, ifelse(deaths > 0, paste0(" (", deaths, "†)"), ""),
                               "\n", inc)),
            size = 2.5, colour = "black", lineheight = 0.85) +
  scale_fill_gradient(low = "#E3F2FD", high = "#0D47A1", name = "Incidence\n(per 100k/yr)") +
  labs(x = "Quarter", y = NULL,
       caption = "Top number = notifications (†deaths); bottom = annualized incidence per 100,000.") +
  theme_nmc() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "right",
    panel.grid = element_blank()
  )
```

:::

---

# Summary Statistics {#sec-summary}

```{r summary-stats}
#| label: tbl-summary
#| tbl-cap: "Summary of agricultural or stock remedy poisoning notifications, South Africa, 2020–present."

total_n      <- nrow(poison)
total_deaths <- sum(poison$deceased, na.rm = TRUE)
overall_cfr  <- round(total_deaths / total_n * 100, 1)

tibble(
  Metric = c(
    "Total notifications",
    "Total deaths reported",
    "Overall case fatality ratio",
    "Provinces affected",
    "Districts affected",
    "Unique reporting facilities",
    "Date range"
  ),
  Value = c(
    scales::comma(total_n),
    scales::comma(total_deaths),
    paste0(overall_cfr, "%"),
    as.character(poison$prov_ %>% na.omit() %>% n_distinct()),
    as.character(poison$district %>% na.omit() %>% n_distinct()),
    as.character(poison$facility %>% na.omit() %>% n_distinct()),
    paste(
      format(as_date(min(poison$notification_date, na.rm = TRUE)), "%d %b %Y"),
      "to",
      format(as_date(max(poison$notification_date, na.rm = TRUE)), "%d %b %Y")
    )
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 10, part = "all")
```

---

# Temporal Trends {#sec-trends}

## Annual notifications and CFR

```{r annual-trend}
#| label: fig-annual-trend
#| fig-cap: "Annual count of agricultural or stock remedy poisoning notifications and case fatality ratio (CFR), South Africa, 2020–present."

poison %>%
  group_by(year = Year_notification) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") -> annual_df

sf_ax <- max(annual_df$n) / max(annual_df$cfr, na.rm = TRUE) * 1.1

ggplot(annual_df, aes(x = year)) +
  geom_col(aes(y = n, fill = "Notifications"), alpha = 0.8, width = 0.6) +
  geom_line(aes(y = cfr * sf_ax, colour = "CFR (%)"), linewidth = 1.2, group = 1) +
  geom_point(aes(y = cfr * sf_ax, colour = "CFR (%)"), size = 3) +
  geom_text(aes(y = n + max(n) * 0.02, label = scales::comma(n)),
            size = 3.2, vjust = 0, color = "#333333") +
  scale_y_continuous(
    name     = "Notifications (n)",
    sec.axis = sec_axis(~ . / sf_ax, name = "CFR (%)", breaks = pretty_breaks(4))
  ) +
  scale_fill_manual(values   = c("Notifications" = "#90CAF9"), name = NULL) +
  scale_colour_manual(values = c("CFR (%)" = "#C62828"), name = NULL) +
  scale_x_continuous(breaks = annual_df$year) +
  labs(x = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Monthly epicurve

```{r epicurve-monthly}
#| label: fig-epicurve-monthly
#| fig-cap: "Monthly notification counts for agricultural or stock remedy poisoning, South Africa, 2020–present. Line = 3-month rolling average."

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  count(month_date) %>%
  arrange(month_date) %>%
  mutate(rolling3 = zoo::rollmean(n, k = 3, fill = NA, align = "right")) -> epicurve_df

ggplot(epicurve_df, aes(x = month_date)) +
  geom_col(aes(y = n), fill = "#90CAF9", colour = "#64B5F6", width = 25) +
  geom_line(aes(y = rolling3), colour = "#C62828", linewidth = 1.1, na.rm = TRUE) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months",
               date_minor_breaks = "3 months", expand = expansion(add = c(0, 0))) +
  scale_y_continuous(name = "Notifications (n)", breaks = pretty_breaks(5)) +
  labs(x = "Date",
       caption = "Red line = 3-month rolling average") +
  theme_nmc() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8))
```

## Seasonal pattern — epiweek heatmap

```{r seasonal-heatmap}
#| label: fig-seasonal-heatmap
#| fig-cap: "Epiweek × year heatmap of agricultural or stock remedy poisoning notification counts, South Africa. The Sep–Nov seasonal peak is clearly visible."
#| fig-height: 5

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(
    nd      = as_date(notification_date),
    yr      = lubridate::year(nd),
    epiweek = lubridate::isoweek(nd)
  ) %>%
  count(year = yr, epiweek) %>%
  complete(year = 2020:year(Sys.Date()), epiweek = 1:52, fill = list(n = 0)) -> heat_df

ggplot(heat_df, aes(x = epiweek,
                    y = factor(year, levels = rev(sort(unique(year)))),
                    fill = n)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "white", mid = "#FFF9C4", high = "#B71C1C",
    midpoint = median(heat_df$n[heat_df$n > 0], na.rm = TRUE),
    name = "Notifications (n)"
  ) +
  scale_x_continuous(breaks = c(1, seq(5, 52, 5)), expand = c(0, 0)) +
  labs(x = "Epiweek", y = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.width = unit(1.5, "cm"),
        panel.grid.major = element_blank())
```

---

# Age & Sex Demographics {#sec-demographics}

## Population pyramid

```{r age-sex-pyramid}
#| label: fig-age-sex-pyramid
#| fig-cap: "Age–sex population pyramid of agricultural or stock remedy poisoning notifications, South Africa, 2020–present."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  count(age_group, sex) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyramid_df

max_n <- max(abs(pyramid_df$n))
step  <- ifelse(max_n > 500, 200, ifelse(max_n > 200, 50, 25))
br    <- seq(0, ceiling(max_n / step) * step, step)

ggplot(pyramid_df, aes(x = age_group, y = n_plot, fill = sex)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  geom_hline(yintercept = 0, colour = "white", linewidth = 0.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y = -max_n * 0.5, label = "Male",   color = "#1565C0", fontface = "bold", size = 3.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y =  max_n * 0.5, label = "Female", color = "#C62828", fontface = "bold", size = 3.5) +
  labs(x = "Age group (years)", y = "Number of notifications") +
  theme_nmc() +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(color = "#DDDDDD", linewidth = 0.3))
```

## Population pyramid by vital status

```{r pyramid-by-vital}
#| label: fig-pyramid-vital
#| fig-cap: "Age–sex pyramid overlaid with vital status. Solid fill = alive; faded = deceased."
#| fig-height: 5.5

poison %>%
  filter(!is.na(age_group), !is.na(sex),
         patient_vital_status %in% c("Alive", "Deceased")) %>%
  count(age_group, sex, patient_vital_status) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyr_vital_df

ggplot(pyr_vital_df,
       aes(x = age_group, y = n_plot, fill = sex, alpha = patient_vital_status)) +
  geom_col(width = 0.8, position = "stack") +
  coord_flip() +
  scale_alpha_manual(values = c("Alive" = 0.9, "Deceased" = 0.35), name = "Vital status") +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = "Sex") +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  geom_hline(yintercept = 0, colour = "white") +
  labs(x = "Age group (years)", y = "Notifications (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Age group distribution over time

```{r age-time}
#| label: fig-age-time
#| fig-cap: "Annual proportion of agricultural or stock remedy poisoning notifications by age group, South Africa."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group)) %>%
  count(Year_notification, age_group) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) -> age_time_df

ggplot(age_time_df, aes(x = factor(Year_notification), y = pct, fill = age_group)) +
  geom_col(position = "fill", width = 0.75) +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_fill_viridis_d(option = "turbo", direction = -1, name = "Age group") +
  labs(x = "Year", y = "Proportion of notifications (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Sex distribution over time

```{r sex-time}
#| label: fig-sex-time
#| fig-cap: "Annual proportion of agricultural or stock remedy poisoning notifications by sex, South Africa."
#| fig-height: 4.5

poison %>%
  filter(!is.na(sex)) %>%
  count(Year_notification, sex) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) -> sex_time_df

ggplot(sex_time_df, aes(x = factor(Year_notification), y = pct, fill = sex)) +
  geom_col(position = "fill", width = 0.75) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_fill(vjust = 0.5),
            size = 3.2, color = "white", fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  labs(x = "Year", y = "Proportion of notifications (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Case fatality ratio by age group and sex

```{r cfr-age-sex}
#| label: fig-cfr-age-sex
#| fig-cap: "Case fatality ratio (%) by age group and sex for agricultural or stock remedy poisoning, South Africa, 2020–present."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  group_by(age_group, sex) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") -> cfr_df

ggplot(cfr_df, aes(x = age_group, y = cfr, fill = sex)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = paste0(round(cfr, 1), "%")),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3, color = "#333333") +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  scale_y_continuous(limits = c(0, max(cfr_df$cfr, na.rm = TRUE) * 1.25)) +
  labs(x = "Age group (years)", y = "Case fatality ratio (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Summary demographics table

```{r demo-table}
#| label: tbl-demographics
#| tbl-cap: "Summary demographics of agricultural or stock remedy poisoning notifications by year and vital status, South Africa."

poison %>%
  mutate(year_chr = factor(Year_notification)) %>%
  select(year_chr, sex, age_group, Age_years, patient_vital_status, prov_) %>%
  tbl_strata(
    strata   = year_chr,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by        = patient_vital_status,
        missing   = "no",
        label     = list(sex ~ "Sex", age_group ~ "Age group",
                         Age_years ~ "Age (years)", prov_ ~ "Province"),
        type      = list(Age_years ~ "continuous"),
        statistic = list(
          all_categorical() ~ "{n}/{N} ({p}%)",
          Age_years ~ "{median} ({p25}\u2013{p75})"
        )
      ) %>%
      bold_labels()
  ) %>%
  as_flex_table()
```

---

# Geographical Distribution {#sec-geo}

```{r geo-setup}
#| include: false

# ── Provincial incidence ──────────────────────────────────────────────────────
NMCleaner::pop %>%
  mutate(Year = as.integer(Year)) %>%
  group_by(prov, Year) %>%
  summarise(population = sum(Population), .groups = "drop") -> pop_prov

poison %>%
  group_by(prov_, year = Year_notification) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(pop_prov, by = c("prov_" = "prov", "year" = "Year")) %>%
  mutate(incidence = n / population * 1e5) -> incidence_prov_df

# ── District incidence ────────────────────────────────────────────────────────
NMCleaner::pop %>%
  filter(as.integer(Year) %in% 2020:year(Sys.Date())) %>%
  group_by(district_standard) %>%
  summarise(pop_total = sum(Population), .groups = "drop") -> pop_dist

poison %>%
  filter(!district %in% c("Unknown", "Isazi Import", "Not Applicable"),
         !is.na(district)) %>%
  mutate_district_name("district") %>%
  group_by(Province = prov_, district_standard) %>%
  summarise(
    Notifications = n(),
    Deaths        = sum(deceased, na.rm = TRUE),
    CFR           = round(Deaths / Notifications * 100, 1),
    .groups = "drop"
  ) %>%
  left_join(pop_dist, by = "district_standard") %>%
  mutate(
    Incidence = round(Notifications / pop_total * 1e5, 1),
    pop_total = round(pop_total / 1e3, 1)          # display in thousands
  ) -> incidence_dist_df
```

## Provincial incidence over time

```{r prov-incidence-line}
#| label: fig-prov-incidence-line
#| fig-cap: "Incidence of agricultural or stock remedy poisoning notifications per 100,000 population by province and year, South Africa."
#| fig-height: 5.5

incidence_prov_df %>%
  group_by(prov_) %>%
  filter(incidence == max(incidence, na.rm = TRUE)) %>%
  ungroup() -> peak_prov_df

ggplot(incidence_prov_df, aes(x = year, y = incidence, colour = prov_, group = prov_)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  ggrepel::geom_label_repel(data = peak_prov_df, aes(label = prov_),
    size = 2.8, show.legend = FALSE, label.size = 0.15, max.overlaps = 20) +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  scale_x_continuous(breaks = unique(incidence_prov_df$year)) +
  labs(x = "Year", y = "Incidence per 100,000 population") +
  theme_nmc() +
  theme(legend.position = "none")
```

## Provincial incidence table

```{r prov-table}
#| label: tbl-prov-incidence
#| tbl-cap: "Notifications and incidence per 100,000 by province and year, South Africa."

incidence_prov_df %>%
  mutate(year = as.character(year)) %>%
  pivot_wider(id_cols = prov_, names_from = year,
              values_from = c(n, incidence), names_glue = "{.value}_{year}") %>%
  rename(Province = prov_) %>%
  gt(rowname_col = "Province") %>%
  tab_spanner_delim(delim = "_") %>%
  fmt_number(starts_with("n_"),         decimals = 0, use_seps = TRUE) %>%
  fmt_number(starts_with("incidence_"), decimals = 1) %>%
  tab_header(title = md("**Provincial notifications and incidence per 100,000**")) %>%
  tab_source_note("Incidence per 100,000 using mid-year provincial population estimates.")
```

## Top 20 districts by notification count

```{r top-districts}
#| label: tbl-top-districts
#| tbl-cap: "Top 20 districts by total notification count, South Africa, 2020–present. Pink = CFR > 5%."

incidence_dist_df %>%
  arrange(desc(Notifications)) %>%
  slice(1:20) %>%
  rename(District = district_standard, `Pop (k)` = pop_total) %>%
  select(Province, District, Notifications, Deaths, `CFR (%)` = CFR,
         `Incidence per 100k` = Incidence, `Pop (k)`) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  bg(i = ~ `Incidence per 100k` > median(incidence_dist_df$Incidence, na.rm = TRUE) * 2,
     j = "Incidence per 100k", bg = "#FFF3E0") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 9, part = "all")
```

## Top 30 districts by incidence rate

```{r district-incidence-lollipop}
#| label: fig-district-incidence
#| fig-cap: "Top 30 districts ranked by incidence per 100,000 population, agricultural or stock remedy poisoning, South Africa, 2020–present. Districts with high incidence but low absolute counts are often missed by count-only rankings."
#| fig-height: 8

incidence_dist_df %>%
  filter(!is.na(Incidence)) %>%
  arrange(desc(Incidence)) %>%
  slice(1:30) %>%
  mutate(
    district_standard = fct_reorder(district_standard, Incidence),
    label = paste0(Province, " · ", round(Incidence, 1))
  ) -> top_inc_df

ggplot(top_inc_df, aes(x = Incidence, y = district_standard)) +
  geom_segment(aes(x = 0, xend = Incidence,
                   y = district_standard, yend = district_standard),
               colour = "grey75", linewidth = 0.5) +
  geom_point(aes(size = Notifications, colour = Province), alpha = 0.85) +
  geom_text(aes(label = round(Incidence, 1)), hjust = -0.5, size = 2.8,
            colour = "#333333") +
  scale_size_continuous(range = c(2, 9), name = "Notifications (n)") +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  labs(
    x = "Incidence per 100,000 population",
    y = NULL,
    caption = "Point size proportional to number of notifications. Colour = province."
  ) +
  theme_nmc() +
  theme(legend.position = "bottom",
        panel.grid.major.x = element_line(colour = "#EEEEEE", linewidth = 0.3))
```

## Count vs incidence — identifying overlooked high-burden districts

```{r district-scatter}
#| label: fig-district-scatter
#| fig-cap: "Notification count versus incidence rate per 100,000 population for all districts, South Africa, 2020–present. Quadrant lines = national median count and median incidence. Districts in the top-left quadrant have high incidence relative to their case count (small, high-burden populations)."
#| fig-height: 6

med_n   <- median(incidence_dist_df$Notifications, na.rm = TRUE)
med_inc <- median(incidence_dist_df$Incidence,     na.rm = TRUE)

incidence_dist_df %>%
  filter(!is.na(Incidence)) %>%
  mutate(quadrant = case_when(
    Notifications > med_n   & Incidence > med_inc ~ "High count, high incidence",
    Notifications <= med_n  & Incidence > med_inc ~ "Low count, high incidence",
    Notifications > med_n   & Incidence <= med_inc ~ "High count, low incidence",
    TRUE                                           ~ "Low count, low incidence"
  )) -> scatter_df

# label top 10 by incidence + top 10 by count
top_labels <- scatter_df %>%
  arrange(desc(Incidence)) %>% slice(1:10) %>%
  bind_rows(scatter_df %>% arrange(desc(Notifications)) %>% slice(1:10)) %>%
  distinct(district_standard, .keep_all = TRUE)

ggplot(scatter_df, aes(x = Notifications, y = Incidence)) +
  geom_hline(yintercept = med_inc, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_vline(xintercept = med_n,   linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = Province, size = pop_total), alpha = 0.75) +
  ggrepel::geom_label_repel(
    data = top_labels,
    aes(label = district_standard),
    size = 2.4, label.size = 0.1, max.overlaps = 25,
    fill = "white", colour = "#333333"
  ) +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  scale_size_continuous(range = c(1.5, 7), name = "Pop (thousands)") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    x = "Total notifications (n)",
    y = "Incidence per 100,000 population",
    caption = "Dashed lines = national medians. Top 10 districts by incidence and by count labelled."
  ) +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Full district incidence table

```{r tbl-district-incidence}
#| label: tbl-district-incidence
#| tbl-cap: "All districts with at least 5 notifications ranked by incidence per 100,000 population, South Africa, 2020–present."

incidence_dist_df %>%
  filter(Notifications >= 5, !is.na(Incidence)) %>%
  arrange(desc(Incidence)) %>%
  rename(District = district_standard, `Pop (k)` = pop_total,
         `Incidence per 100k` = Incidence, `CFR (%)` = CFR) %>%
  select(Province, District, Notifications, Deaths, `CFR (%)`,
         `Incidence per 100k`, `Pop (k)`) %>%
  flextable() %>%
  bold(part = "header") %>%
  color(i = ~ `Incidence per 100k` > quantile(incidence_dist_df$Incidence,
                                               0.9, na.rm = TRUE),
        j = "Incidence per 100k", color = "#C62828") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 9, part = "all") %>%
  add_footer_lines("Red = top 10% incidence districts. Pink = CFR > 5%.")
```

---

# Mortality Profile {#sec-mortality}

## Deaths by province and year

```{r deaths-prov}
#| label: fig-deaths-prov
#| fig-cap: "Annual deaths from agricultural or stock remedy poisoning by province, South Africa."
#| fig-height: 5

poison %>%
  filter(deceased) %>%
  count(year = Year_notification, prov_) %>%
  ggplot(aes(x = factor(year), y = n, fill = prov_)) +
  geom_col(position = "stack", width = 0.75) +
  scale_fill_brewer(palette = "Paired", name = "Province") +
  labs(x = "Year", y = "Deaths (n)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## CFR by province and year

```{r cfr-prov}
#| label: fig-cfr-prov
#| fig-cap: "Case fatality ratio (%) by province and year for agricultural or stock remedy poisoning, South Africa."
#| fig-height: 5

poison %>%
  group_by(year = Year_notification, prov_) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") %>%
  ggplot(aes(x = factor(year), y = cfr, fill = prov_)) +
  geom_col(position = "dodge", width = 0.85) +
  scale_fill_brewer(palette = "Paired", name = "Province") +
  labs(x = "Year", y = "CFR (%)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## Deaths in children under 12

```{r under12-deaths}
#| label: tbl-under12-deaths
#| tbl-cap: "Vital status of agricultural or stock remedy poisoning notifications in children under 12 years, by year, South Africa."

poison %>%
  filter(Year_notification %in% 2023:year(Sys.Date())) %>%
  mutate(
    under_12 = factor(if_else(Age_years <= 12, "Under 12 years", "12 years and over"),
                      levels = c("Under 12 years", "12 years and over"))
  ) %>%
  select(year = Year_notification, under_12, patient_vital_status, Age_years, sex, prov_) %>%
  tbl_strata(
    strata   = year,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by      = patient_vital_status, missing = "no",
        label   = list(under_12 ~ "Age category", Age_years ~ "Age (years)",
                       sex ~ "Sex", prov_ ~ "Province"),
        type    = list(Age_years ~ "continuous"), percent = "row"
      ) %>% bold_labels()
  ) %>%
  as_flex_table()
```

---

# Reporting & Facility Analysis {#sec-facilities}

## Unique reporting facilities per year

```{r facilities-trend}
#| label: fig-facilities
#| fig-cap: "Number of unique health facilities reporting at least one agricultural or stock remedy poisoning notification per year, South Africa."
#| fig-height: 4

poison %>%
  filter(!is.na(facility)) %>%
  group_by(Year_notification) %>%
  summarise(n_facilities = n_distinct(facility), .groups = "drop") -> fac_df

ggplot(fac_df, aes(x = factor(Year_notification))) +
  geom_col(aes(y = n_facilities, fill = "Unique facilities"), width = 0.6, alpha = 0.85) +
  geom_text(aes(y = n_facilities, label = n_facilities), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Unique facilities" = "#4CAF50"), name = NULL) +
  labs(x = "Year", y = "Unique reporting facilities (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

---

*Report generated: `r format(Sys.Date(), "%d %B %Y")` | NMC Surveillance Unit*
