---
title: "Agricultural & Stock Remedy Poisoning"
subtitle: "Epidemiological Description, South Africa · NMC Surveillance"
date: today
sidebar: special-reports
execute:
  echo: false
  warning: false
  message: false
  error: false
  cache: false
---

::: {.callout-important appearance="minimal"}
**Priority condition** — Agricultural and stock remedy poisoning is a current public health priority in South Africa. This report describes the epidemiology based on NMC notifiable conditions surveillance data, 2020–present.
:::

```{r setup}
#| label: setup
#| include: false
#| cache: false

pacman::p_load(
  ggplot2, gtsummary, tidyverse, magrittr,
  conflicted, scales, data.table, lubridate,
  flextable, gt, ggrepel, zoo,
  janitor
)

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::lag)
conflicted::conflicts_prefer(tidyr::extract)
conflicted::conflicts_prefer(dplyr::first)
conflicted::conflicts_prefer(lubridate::year)

library(NMCleaner)

# ── Load master dataset ───────────────────────────────────────────────────────
load("~/Desktop/SAFETP/CLA/NMC_database/master/new_master.rda")

# ── Filter to Agricultural poisoning only ────────────────────────────────────
CONDITION <- "Agricultural or stock remedy poisoning"

new_master %>%
  filter(
    condition %in% CONDITION,
    as.integer(Year_notification) %in% 2020:year(Sys.Date()),
    !is.na(prov_)
  ) %>%
  mutate(
    Year_notification = as.integer(Year_notification),
    patient_vital_status = if_else(
      patient_vital_status %like% "Unknown", "Alive", patient_vital_status
    ),
    Age_years = as.numeric(Age_years),
    age_group = cut(
      Age_years,
      breaks = c(-Inf, 4, 14, 29, 44, 59, Inf),
      labels = c("<5", "5\u201314", "15\u201329", "30\u201344", "45\u201359", "60+"),
      right  = TRUE
    ),
    sex = case_when(
      patient_gender %in% c("Male", "M")   ~ "Male",
      patient_gender %in% c("Female", "F") ~ "Female",
      TRUE                                  ~ NA_character_
    ),
    deceased = patient_vital_status %in% "Deceased"
  ) -> poison

# ── Shared theme ──────────────────────────────────────────────────────────────
theme_nmc <- function(...) {
  theme_classic() +
    theme(
      text              = element_text(family = "Century Gothic", size = 10, color = "#333333"),
      panel.background  = element_rect(fill = "#F8F8F8"),
      plot.background   = element_rect(fill = "#F8F8F8"),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      axis.ticks        = element_line(linewidth = 0.15, color = "grey40"),
      legend.background = element_rect(fill = "#F8F8F8", linewidth = 0.25,
                                       linetype = "solid", colour = "grey50"),
      legend.key        = element_blank(),
      strip.background  = element_rect(color = NA, fill = NA),
      ...
    )
}
```

::: {.callout-note appearance="minimal" collapse="true"}
## Report metadata
| | |
|---|---|
| **Condition** | `r CONDITION` |
| **Years** | 2020 – `r lubridate::year(Sys.Date())` |
| **Data source** | NMC Notifiable Medical Conditions surveillance database |
| **Generated** | `r format(Sys.Date(), "%d %B %Y")` |
:::

# Summary Statistics {#sec-summary}

```{r summary-stats}
#| label: tbl-summary
#| tbl-cap: "Summary of agricultural or stock remedy poisoning notifications, South Africa, 2020–present."

total_n      <- nrow(poison)
total_deaths <- sum(poison$deceased, na.rm = TRUE)
overall_cfr  <- round(total_deaths / total_n * 100, 1)

tibble(
  Metric = c(
    "Total notifications",
    "Total deaths reported",
    "Overall case fatality ratio",
    "Provinces affected",
    "Districts affected",
    "Unique reporting facilities",
    "Date range"
  ),
  Value = c(
    scales::comma(total_n),
    scales::comma(total_deaths),
    paste0(overall_cfr, "%"),
    as.character(poison$prov_ %>% na.omit() %>% n_distinct()),
    as.character(poison$district %>% na.omit() %>% n_distinct()),
    as.character(poison$facility %>% na.omit() %>% n_distinct()),
    paste(
      format(as_date(min(poison$notification_date, na.rm = TRUE)), "%d %b %Y"),
      "to",
      format(as_date(max(poison$notification_date, na.rm = TRUE)), "%d %b %Y")
    )
  )
) %>%
  flextable() %>%
  bold(part = "header") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 10, part = "all")
```

---

# Temporal Trends {#sec-trends}

## Annual notifications and CFR

```{r annual-trend}
#| label: fig-annual-trend
#| fig-cap: "Annual count of agricultural or stock remedy poisoning notifications and case fatality ratio (CFR), South Africa, 2020–present."

poison %>%
  group_by(year = Year_notification) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") -> annual_df

sf_ax <- max(annual_df$n) / max(annual_df$cfr, na.rm = TRUE) * 1.1

ggplot(annual_df, aes(x = year)) +
  geom_col(aes(y = n, fill = "Notifications"), alpha = 0.8, width = 0.6) +
  geom_line(aes(y = cfr * sf_ax, colour = "CFR (%)"), linewidth = 1.2, group = 1) +
  geom_point(aes(y = cfr * sf_ax, colour = "CFR (%)"), size = 3) +
  geom_text(aes(y = n + max(n) * 0.02, label = scales::comma(n)),
            size = 3.2, vjust = 0, color = "#333333") +
  scale_y_continuous(
    name     = "Notifications (n)",
    sec.axis = sec_axis(~ . / sf_ax, name = "CFR (%)", breaks = pretty_breaks(4))
  ) +
  scale_fill_manual(values   = c("Notifications" = "#90CAF9"), name = NULL) +
  scale_colour_manual(values = c("CFR (%)" = "#C62828"), name = NULL) +
  scale_x_continuous(breaks = annual_df$year) +
  labs(x = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Monthly epicurve

```{r epicurve-monthly}
#| label: fig-epicurve-monthly
#| fig-cap: "Monthly notification counts for agricultural or stock remedy poisoning, South Africa, 2020–present. Line = 3-month rolling average."

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(month_date = floor_date(as_date(notification_date), "month")) %>%
  count(month_date) %>%
  arrange(month_date) %>%
  mutate(rolling3 = zoo::rollmean(n, k = 3, fill = NA, align = "right")) -> epicurve_df

ggplot(epicurve_df, aes(x = month_date)) +
  geom_col(aes(y = n), fill = "#90CAF9", colour = "#64B5F6", width = 25) +
  geom_line(aes(y = rolling3), colour = "#C62828", linewidth = 1.1, na.rm = TRUE) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months",
               date_minor_breaks = "3 months", expand = expansion(add = c(0, 0))) +
  scale_y_continuous(name = "Notifications (n)", breaks = pretty_breaks(5)) +
  labs(x = "Date",
       caption = "Red line = 3-month rolling average") +
  theme_nmc() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8))
```

## Seasonal pattern — epiweek heatmap

```{r seasonal-heatmap}
#| label: fig-seasonal-heatmap
#| fig-cap: "Epiweek × year heatmap of agricultural or stock remedy poisoning notification counts, South Africa. The Sep–Nov seasonal peak is clearly visible."
#| fig-height: 5

poison %>%
  filter(!is.na(notification_date)) %>%
  mutate(
    nd      = as_date(notification_date),
    yr      = lubridate::year(nd),
    epiweek = lubridate::isoweek(nd)
  ) %>%
  count(year = yr, epiweek) %>%
  complete(year = 2020:year(Sys.Date()), epiweek = 1:52, fill = list(n = 0)) -> heat_df

ggplot(heat_df, aes(x = epiweek,
                    y = factor(year, levels = rev(sort(unique(year)))),
                    fill = n)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "white", mid = "#FFF9C4", high = "#B71C1C",
    midpoint = median(heat_df$n[heat_df$n > 0], na.rm = TRUE),
    name = "Notifications (n)"
  ) +
  scale_x_continuous(breaks = c(1, seq(5, 52, 5)), expand = c(0, 0)) +
  labs(x = "Epiweek", y = "Year") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.width = unit(1.5, "cm"),
        panel.grid.major = element_blank())
```

---

# Age & Sex Demographics {#sec-demographics}

## Population pyramid

```{r age-sex-pyramid}
#| label: fig-age-sex-pyramid
#| fig-cap: "Age–sex population pyramid of agricultural or stock remedy poisoning notifications, South Africa, 2020–present."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  count(age_group, sex) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyramid_df

max_n <- max(abs(pyramid_df$n))
step  <- ifelse(max_n > 500, 200, ifelse(max_n > 200, 50, 25))
br    <- seq(0, ceiling(max_n / step) * step, step)

ggplot(pyramid_df, aes(x = age_group, y = n_plot, fill = sex)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  geom_hline(yintercept = 0, colour = "white", linewidth = 0.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y = -max_n * 0.5, label = "Male",   color = "#1565C0", fontface = "bold", size = 3.5) +
  annotate("text", x = nlevels(pyramid_df$age_group) + 0.4,
           y =  max_n * 0.5, label = "Female", color = "#C62828", fontface = "bold", size = 3.5) +
  labs(x = "Age group (years)", y = "Number of notifications") +
  theme_nmc() +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(color = "#DDDDDD", linewidth = 0.3))
```

## Population pyramid by vital status

```{r pyramid-by-vital}
#| label: fig-pyramid-vital
#| fig-cap: "Age–sex pyramid overlaid with vital status. Solid fill = alive; faded = deceased."
#| fig-height: 5.5

poison %>%
  filter(!is.na(age_group), !is.na(sex),
         patient_vital_status %in% c("Alive", "Deceased")) %>%
  count(age_group, sex, patient_vital_status) %>%
  mutate(n_plot = if_else(sex == "Male", -n, n)) -> pyr_vital_df

ggplot(pyr_vital_df,
       aes(x = age_group, y = n_plot, fill = sex, alpha = patient_vital_status)) +
  geom_col(width = 0.8, position = "stack") +
  coord_flip() +
  scale_alpha_manual(values = c("Alive" = 0.9, "Deceased" = 0.35), name = "Vital status") +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = "Sex") +
  scale_y_continuous(breaks = c(-br, br), labels = abs(c(-br, br))) +
  geom_hline(yintercept = 0, colour = "white") +
  labs(x = "Age group (years)", y = "Notifications (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Age group distribution over time

```{r age-time}
#| label: fig-age-time
#| fig-cap: "Annual proportion of agricultural or stock remedy poisoning notifications by age group, South Africa."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group)) %>%
  count(Year_notification, age_group) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) -> age_time_df

ggplot(age_time_df, aes(x = factor(Year_notification), y = pct, fill = age_group)) +
  geom_col(position = "fill", width = 0.75) +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, name = "Age group") +
  labs(x = "Year", y = "Proportion of notifications (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Sex distribution over time

```{r sex-time}
#| label: fig-sex-time
#| fig-cap: "Annual proportion of agricultural or stock remedy poisoning notifications by sex, South Africa."
#| fig-height: 4.5

poison %>%
  filter(!is.na(sex)) %>%
  count(Year_notification, sex) %>%
  group_by(Year_notification) %>%
  mutate(pct = n / sum(n) * 100) -> sex_time_df

ggplot(sex_time_df, aes(x = factor(Year_notification), y = pct, fill = sex)) +
  geom_col(position = "fill", width = 0.75) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_fill(vjust = 0.5),
            size = 3.2, color = "white", fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  labs(x = "Year", y = "Proportion of notifications (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Case fatality ratio by age group and sex

```{r cfr-age-sex}
#| label: fig-cfr-age-sex
#| fig-cap: "Case fatality ratio (%) by age group and sex for agricultural or stock remedy poisoning, South Africa, 2020–present."
#| fig-height: 5

poison %>%
  filter(!is.na(age_group), !is.na(sex)) %>%
  group_by(age_group, sex) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") -> cfr_df

ggplot(cfr_df, aes(x = age_group, y = cfr, fill = sex)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = paste0(round(cfr, 1), "%")),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3, color = "#333333") +
  scale_fill_manual(values = c("Male" = "#1565C0", "Female" = "#C62828"), name = NULL) +
  scale_y_continuous(limits = c(0, max(cfr_df$cfr, na.rm = TRUE) * 1.25)) +
  labs(x = "Age group (years)", y = "Case fatality ratio (%)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

## Summary demographics table

```{r demo-table}
#| label: tbl-demographics
#| tbl-cap: "Summary demographics of agricultural or stock remedy poisoning notifications by year and vital status, South Africa."

poison %>%
  mutate(year_chr = factor(Year_notification)) %>%
  select(year_chr, sex, age_group, Age_years, patient_vital_status, prov_) %>%
  tbl_strata(
    strata   = year_chr,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by        = patient_vital_status,
        missing   = "no",
        label     = list(sex ~ "Sex", age_group ~ "Age group",
                         Age_years ~ "Age (years)", prov_ ~ "Province"),
        type      = list(Age_years ~ "continuous"),
        statistic = list(Age_years ~ "{median} ({p25}\u2013{p75})")
      ) %>%
      bold_labels()
  ) %>%
  as_flex_table()
```

---

# Geographical Distribution {#sec-geo}

```{r geo-setup}
#| include: false

NMCleaner::pop %>%
  mutate(Year = as.integer(Year)) %>%
  group_by(prov, Year) %>%
  summarise(population = sum(Population), .groups = "drop") -> pop_prov

poison %>%
  group_by(prov_, year = Year_notification) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(pop_prov, by = c("prov_" = "prov", "year" = "Year")) %>%
  mutate(incidence = n / population * 1e5) -> incidence_prov_df
```

## Provincial incidence over time

```{r prov-incidence-line}
#| label: fig-prov-incidence-line
#| fig-cap: "Incidence of agricultural or stock remedy poisoning notifications per 100,000 population by province and year, South Africa."
#| fig-height: 5.5

incidence_prov_df %>%
  group_by(prov_) %>%
  filter(incidence == max(incidence, na.rm = TRUE)) %>%
  ungroup() -> peak_prov_df

ggplot(incidence_prov_df, aes(x = year, y = incidence, colour = prov_, group = prov_)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  ggrepel::geom_label_repel(data = peak_prov_df, aes(label = prov_),
    size = 2.8, show.legend = FALSE, label.size = 0.15, max.overlaps = 20) +
  scale_colour_brewer(palette = "Paired", name = "Province") +
  scale_x_continuous(breaks = unique(incidence_prov_df$year)) +
  labs(x = "Year", y = "Incidence per 100,000 population") +
  theme_nmc() +
  theme(legend.position = "none")
```

## Provincial incidence table

```{r prov-table}
#| label: tbl-prov-incidence
#| tbl-cap: "Notifications and incidence per 100,000 by province and year, South Africa."

incidence_prov_df %>%
  mutate(year = as.character(year)) %>%
  pivot_wider(id_cols = prov_, names_from = year,
              values_from = c(n, incidence), names_glue = "{.value}_{year}") %>%
  rename(Province = prov_) %>%
  gt(rowname_col = "Province") %>%
  tab_spanner_delim(delim = "_") %>%
  fmt_number(starts_with("n_"),         decimals = 0, use_seps = TRUE) %>%
  fmt_number(starts_with("incidence_"), decimals = 1) %>%
  tab_header(title = md("**Provincial notifications and incidence per 100,000**")) %>%
  tab_source_note("Incidence per 100,000 using mid-year provincial population estimates.")
```

## Top 20 districts

```{r top-districts}
#| label: tbl-top-districts
#| tbl-cap: "Top 20 districts by total notification count, South Africa, 2020–present. Pink = CFR > 5%."

poison %>%
  filter(!district %in% c("Unknown", "Isazi Import", "Not Applicable"), !is.na(district)) %>%
  group_by(Province = str_to_upper(prov_), District = district) %>%
  summarise(Notifications = n(), Deaths = sum(deceased, na.rm = TRUE),
            `CFR (%)` = round(Deaths / Notifications * 100, 1), .groups = "drop") %>%
  arrange(desc(Notifications)) %>%
  slice(1:20) %>%
  flextable() %>%
  bold(part = "header") %>%
  bg(i = ~ `CFR (%)` > 5, j = "CFR (%)", bg = "#FFCDD2") %>%
  set_table_properties(layout = "autofit") %>%
  fontsize(size = 9, part = "all")
```

---

# Mortality Profile {#sec-mortality}

## Deaths by province and year

```{r deaths-prov}
#| label: fig-deaths-prov
#| fig-cap: "Annual deaths from agricultural or stock remedy poisoning by province, South Africa."
#| fig-height: 5

poison %>%
  filter(deceased) %>%
  count(year = Year_notification, prov_) %>%
  ggplot(aes(x = factor(year), y = n, fill = prov_)) +
  geom_col(position = "stack", width = 0.75) +
  scale_fill_brewer(palette = "Paired", name = "Province") +
  labs(x = "Year", y = "Deaths (n)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## CFR by province and year

```{r cfr-prov}
#| label: fig-cfr-prov
#| fig-cap: "Case fatality ratio (%) by province and year for agricultural or stock remedy poisoning, South Africa."
#| fig-height: 5

poison %>%
  group_by(year = Year_notification, prov_) %>%
  summarise(n = n(), deaths = sum(deceased, na.rm = TRUE),
            cfr = deaths / n * 100, .groups = "drop") %>%
  ggplot(aes(x = factor(year), y = cfr, fill = prov_)) +
  geom_col(position = "dodge", width = 0.85) +
  scale_fill_brewer(palette = "Paired", name = "Province") +
  labs(x = "Year", y = "CFR (%)") +
  theme_nmc() +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
```

## Deaths in children under 12

```{r under12-deaths}
#| label: tbl-under12-deaths
#| tbl-cap: "Vital status of agricultural or stock remedy poisoning notifications in children under 12 years, by year, South Africa."

poison %>%
  filter(Year_notification %in% 2023:year(Sys.Date())) %>%
  mutate(
    under_12 = factor(if_else(Age_years <= 12, "Under 12 years", "12 years and over"),
                      levels = c("Under 12 years", "12 years and over"))
  ) %>%
  select(year = Year_notification, under_12, patient_vital_status, Age_years, sex, prov_) %>%
  tbl_strata(
    strata   = year,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by      = patient_vital_status, missing = "no",
        label   = list(under_12 ~ "Age category", Age_years ~ "Age (years)",
                       sex ~ "Sex", prov_ ~ "Province"),
        type    = list(Age_years ~ "continuous"), percent = "row"
      ) %>% bold_labels()
  ) %>%
  as_flex_table()
```

---

# Reporting & Facility Analysis {#sec-facilities}

## Unique reporting facilities per year

```{r facilities-trend}
#| label: fig-facilities
#| fig-cap: "Number of unique health facilities reporting at least one agricultural or stock remedy poisoning notification per year, South Africa."
#| fig-height: 4

poison %>%
  filter(!is.na(facility)) %>%
  group_by(Year_notification) %>%
  summarise(n_facilities = n_distinct(facility), .groups = "drop") -> fac_df

ggplot(fac_df, aes(x = factor(Year_notification))) +
  geom_col(aes(y = n_facilities, fill = "Unique facilities"), width = 0.6, alpha = 0.85) +
  geom_text(aes(y = n_facilities, label = n_facilities), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Unique facilities" = "#4CAF50"), name = NULL) +
  labs(x = "Year", y = "Unique reporting facilities (n)") +
  theme_nmc() +
  theme(legend.position = "bottom")
```

---

*Report generated: `r format(Sys.Date(), "%d %B %Y")` | NMC Surveillance Unit*
