---
title: "Weekly IDSR Reports"
subtitle: "Notifiable Medical Conditions Surveillance System — Weekly Integrated Disease Surveillance & Response"
sidebar: weekly-reports
listing:
  id: reports-listing
  contents: posts
  type: grid
  sort: "date desc"
  categories: true
  fields: [title, description, categories]
  grid-columns: 3
  page-size: 12
  filter-ui: true
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
/* ── Latest Report Hero ───────────────────────────────────────────────────── */
.latest-hero {
  background: linear-gradient(135deg, #eff6ff 0%, #f0fdf4 100%);
  border: 1px solid #bfdbfe;
  border-radius: 12px;
  padding: 1.5rem 2rem;
  margin-bottom: 1.75rem;
  display: flex;
  align-items: center;
  justify-content: space-between;
  flex-wrap: wrap;
  gap: 1rem;
}
.latest-hero-label {
  display: inline-block;
  background: #2563eb;
  color: #fff;
  font-size: 0.7rem;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: 0.07em;
  padding: 0.2rem 0.65rem;
  border-radius: 99px;
  margin-bottom: 0.5rem;
}
.latest-hero-title {
  font-size: 1.3rem;
  font-weight: 700;
  color: #1e3a8a;
  margin: 0 0 0.2rem 0;
}
.latest-hero-sub {
  color: #475569;
  font-size: 0.9rem;
  margin: 0;
}
.latest-hero-actions {
  display: flex;
  gap: 0.65rem;
  flex-shrink: 0;
}

/* ── Listing card text tweaks ─────────────────────────────────────────────── */
.quarto-listing-grid .card-description {
  font-size: 0.85rem;
  color: #475569;
}

/* ── Report structure format badges ──────────────────────────────────────── */
.fmt-text, .fmt-table, .fmt-fig {
  display: inline-block;
  font-size: 0.7rem;
  font-weight: 600;
  padding: 0.15rem 0.55rem;
  border-radius: 4px;
  text-transform: uppercase;
  letter-spacing: 0.04em;
  white-space: nowrap;
}
.fmt-text  { background: #e2e8f0; color: #334155; }
.fmt-table { background: #dbeafe; color: #1d4ed8; }
.fmt-fig   { background: #dcfce7; color: #166534; }

@media (max-width: 600px) {
  .latest-hero { flex-direction: column; }
}
</style>
```

```{r setup}
#| label: setup
#| include: false

library(NMCleaner)
library(ragg)
library(DT)

conflicted::conflicts_prefer(
  magrittr::extract,
  dplyr::between,
  lubridate::isoweek,
  dplyr::filter,
  lubridate::week
)

`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a

# ── Load surveillance data ────────────────────────────────────────────────────
load("~/Desktop/SAFETP/CLA/NMC_database/master/new_master.rda")
load("~/Desktop/SAFETP/CLA/NMC_database/database_scraper_database.rda")

export_date  <- database_scraper_database$export_date[1]
export_label <- format(as.POSIXct(export_date), "%d %B %Y")
current_year <- as.integer(format(Sys.Date(), "%Y"))

# ── Locate most-recent post stub (drives the dynamic hero) ───────────────────
posts_dir  <- here::here("weekly-reports", "posts")
stub_files <- sort(
  list.files(posts_dir, pattern = "^[0-9]{4}-W[0-9]+\\.qmd$", full.names = TRUE),
  decreasing = TRUE
)

if (length(stub_files) > 0) {
  latest_meta   <- rmarkdown::yaml_front_matter(stub_files[1])
  latest_title  <- latest_meta$title  %||% "Latest Report"
  latest_date   <- as.Date(as.character(latest_meta$date %||% Sys.Date()))
  latest_period <- paste0(
    format(latest_date,     "%d %B"), " \u2013 ",
    format(latest_date + 6, "%d %B %Y")
  )
  latest_stem   <- tools::file_path_sans_ext(basename(stub_files[1]))
} else {
  latest_title  <- "No reports yet"
  latest_period <- ""
  latest_stem   <- ""
}
```

::: {.callout-note appearance="minimal"}
These reports summarise conditions notified to the **NMC Surveillance System (NMCSS)** for each CDC epiweek. They cover Category 1 & 2 notifications flagged for WHO IDSR reporting, with confirmed trends, epitable, and all-notifications overview.
:::

```{r hero}
#| output: asis
cat(sprintf(
'<div class="latest-hero">
  <div>
    <span class="latest-hero-label">Latest Report</span>
    <p class="latest-hero-title">%s</p>
    <p class="latest-hero-sub">%s</p>
  </div>
  <div class="latest-hero-actions">
    <a href="rendered/%s-weekly-report.html" class="btn btn-primary" target="_blank">View Report</a>
    <a href="rendered/%s-weekly-report.docx" class="btn btn-outline-secondary" download="%s-weekly-report.docx">&#8595;&nbsp;Download Word</a>
  </div>
</div>\n',
  latest_title, latest_period,
  latest_stem, latest_stem, latest_stem
))
```

---

## What's currently circulating

Confirmed Category 1 & 2 notifications for selected conditions — last two years. *Rolling 4-week average shown in red.*

```{r fig-overview-epicurve}
#| label: fig-overview-epicurve
#| fig-cap: !expr 'paste0("Confirmed Cat 1 & 2 notifications, key conditions — South Africa (data to ", export_label, ")")'
#| fig-height: 5
#| fig-width: 11
#| out-width: "100%"
#| dev: ragg_png
#| dpi: 180

spotlight <- c(
  "Cholera", "Malaria", "Mpox", "Measles",
  "Pertussis", "Hepatitis A"
)

df_ov <- new_master %>%
  dplyr::filter(case_definition %in% "Confirmed") %>%
  dplyr::filter(nmccategories %in% c(1, 2)) %>%
  dplyr::filter(condition %in% spotlight) %>%
  epicurve_df(., date_index = "notification_date", grouping_vars = "condition") %>%
  dplyr::filter(year %in% (current_year - 1):current_year) %>%
  dplyr::mutate(year = factor(year, levels = (current_year - 1):current_year))

p_ov <- df_ov %>%
  plot_epicurve_df(
    x_axis_option   = "epiweek",
    grouping_vars   = "condition",
    color_select    = "olivedrab",
    add_bar         = TRUE,
    add_rolling_avg = c(TRUE, 4),
    axis_text_size  = 7
  )

p_ov$plot +
  theme(strip.text.y.right = element_text(hjust = 0))
```

---

## Reports

::: {.panel-tabset}

### Browse

::: {#reports-listing}
:::

### All Reports

```{r archive-table}
#| label: archive-table

read_stub <- function(path) {
  meta <- tryCatch(rmarkdown::yaml_front_matter(path), error = function(e) NULL)
  if (is.null(meta)) return(NULL)
  stem     <- tools::file_path_sans_ext(basename(path))
  date_val <- as.Date(as.character(meta$date %||% "1970-01-01"))
  data.frame(
    Epiweek     = meta$title       %||% stem,
    `Week start` = format(date_val, "%d %b %Y"),
    Description = meta$description %||% "",
    stem        = stem,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}

all_stubs <- sort(
  list.files(posts_dir, pattern = "^[0-9]{4}-W[0-9]+\\.qmd$", full.names = TRUE),
  decreasing = TRUE
)

archive_rows <- Filter(Negate(is.null), lapply(all_stubs, read_stub))

if (length(archive_rows) > 0) {
  arch <- do.call(rbind, archive_rows)
  arch$View <- sprintf(
    '<a href="rendered/%s-weekly-report.html" target="_blank" class="btn btn-sm btn-primary">View</a>',
    arch$stem
  )
  arch$Word <- sprintf(
    '<a href="rendered/%s-weekly-report.docx" download="%s-weekly-report.docx" class="btn btn-sm btn-outline-secondary">&#8595;</a>',
    arch$stem, arch$stem
  )
  arch$stem <- NULL

  DT::datatable(
    arch,
    escape    = FALSE,
    rownames  = FALSE,
    options   = list(
      pageLength = 20,
      dom        = "ftp",
      columnDefs = list(
        list(width = "120px", targets = 0),
        list(width = "90px",  targets = 1),
        list(width = "50px",  targets = c(3, 4)),
        list(orderable = FALSE, targets = c(3, 4))
      )
    ),
    class = "display compact"
  )
}
```

:::

---

## Report structure

Each weekly report covers:

| Section | Content | Format |
|---|---|---|
| Introduction | Export date, epiweek dates | [Text]{.fmt-text} |
| Category 1 at a glance | Suspected & confirmed counts, all Category 1 conditions | [Table]{.fmt-table} |
| Confirmed Notifications — Epitable | Weekly confirmed counts per epiweek (with CI & baseline) | [Table]{.fmt-table} |
| Confirmed Notifications — Trends | Epicurve with rolling average for selected conditions | [Figure]{.fmt-fig} |
| All Notifications — Epitable | Weekly all-notification counts per epiweek | [Table]{.fmt-table} |
| All Notifications — Trends | Trends including Fever-Rash composite | [Figure]{.fmt-fig} |

: Report sections {#tbl-report-structure .striped .hover tbl-colwidths="[30,55,15]"}

---

::: {.callout-tip collapse="true" title="How to generate reports"}
**Last 4 weeks (default):**

```r
source("weekly-reports/run_all_weekly.R")
```

**Adjust number of weeks** — edit `N_WEEKS` at the top of `run_all_weekly.R` before running.

Then run `quarto render` or `source("dev/deploy.r")` to publish. A listing stub is created automatically in `weekly-reports/posts/` for each rendered week.
:::
