---
title: "NOTIFIABLE MEDICAL CONDITIONS SURVEILLANCE SYSTEM — Weekly IDSR Report"
author: "The National Institute for Communicable Diseases"
date: today
params:
  reporting_week_start: "2026-04-05"   # ← Sunday of the target CDC epiweek (YYYY-MM-DD)
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    embed-resources: true
    theme: cosmo
    css: "../styles.css"
  docx:
    reference-doc: "../NMC_reports_scripts/extras/weekly_IDSR_V2_template.docx"
    reference-location: "section"
    number-sections: true
    fig-caption: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
  error: false
---

```{r setup}
#| label: setup
#| include: false

# ── Package management ─────────────────────────────────────────────────────────
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

library(NMCleaner)
library(ragg)
library(kableExtra)

conflicted::conflicts_prefer(
  magrittr::extract,
  dplyr::between,
  lubridate::isoweek,
  dplyr::filter,
  lubridate::week
)

# ── Global flextable settings ──────────────────────────────────────────────────
flextable::set_flextable_defaults(
  font.family      = "Arial",
  font.size        = 8,
  header.font.size = 8,
  border.color     = "black",
  padding          = 0.1
)

gtsummary::theme_gtsummary_language(language = "en", big.mark = " ", decimal.mark = ".")
gtsummary::theme_gtsummary_journal("nejm")

knitr::opts_chunk$set(
  ft.align  = "left",
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  out.width = "100%",
  dev       = "ragg_png",
  dpi       = 300
)

# ── Helper: trim gtsummary header row ─────────────────────────────────────────
trim_tbl <- function(gt_summary_object) {
  n <- nrow(gt_summary_object$table_body)
  gt_summary_object$table_body <- gt_summary_object$table_body[2:n, ]
  gt_summary_object
}

# ── Helper: combine Measles + Rubella into Fever-Rash ─────────────────────────
make_fever_rash <- function(data, fever_rash_label = "Fever-Rash") {
  data %>%
    mutate(condition = if_else(
      condition %in% c("Measles", "Rubella"), fever_rash_label, condition
    ))
}

# ── ggplot theme ──────────────────────────────────────────────────────────────
# Pipe style: plot %>% as_excel()  i.e.  as_excel(plot)
as_excel <- function(.) {
  . +
    theme_minimal() +
    theme(
      text             = element_text(family = "Arial", size = 10, color = "#333333"),
      panel.background = element_rect(fill = "#F8F8F8"),
      plot.background  = element_rect(fill = "#F8F8F8"),
      panel.grid       = element_line(color = "#DDDDDD"),
      legend.background = element_rect(fill = "#F8F8F8"),
      legend.text      = element_text(family = "Arial", color = "#333333", size = 8),
      legend.title     = element_text(family = "Arial", color = "#333333", size = 10),
      plot.title       = element_text(family = "Arial", size = 12, color = "#666666"),
      plot.subtitle    = element_text(family = "Arial", size = 10, color = "#666666"),
      plot.caption     = element_text(family = "Arial", size = 8,  color = "#666666"),
      axis.text.x      = element_text(angle = 30)
    )
}

# ── Load data ──────────────────────────────────────────────────────────────────
load("~/Desktop/SAFETP/CLA/NMC_database/master/new_master.rda")
load("~/Desktop/SAFETP/CLA/NMC_database/database_scraper_database.rda")

export_date   <- database_scraper_database$export_date[1]
file_exp_date <- as.POSIXct(export_date)

# ── Reporting period from params ───────────────────────────────────────────────
reporting_week_start <- as_date(params$reporting_week_start)
epiweek_dates        <- seq(reporting_week_start, by = "day", length.out = 7)
epiweek_end          <- reporting_week_start + 6

# CDC epiweek number and year
epiweek_num  <- lubridate::epiweek(reporting_week_start)
epiweek_year <- lubridate::epiyear(reporting_week_start)

end_date_month_year <- format(export_date, "%B, %Y")

formatted_datetime <- format(file_exp_date, "%A, %B %d, %Y %I:%M %p")

reporting_period <- paste0(
  "for **Epiweek ", epiweek_num, "** of ", epiweek_year,
  ", which is from **", format(reporting_week_start, "%A, %B %d"), "** to **",
  format(epiweek_end, "%A, %B %d"), "**."
)

# ── Filter to the reporting epiweek ───────────────────────────────────────────
df <- new_master %>%
  dplyr::filter(as_date(notification_date) %in% epiweek_dates)

df1 <- df %>% dplyr::filter(back_capture_new %in% c("Current", "Delayed", "unknown"))
df2 <- df %>% dplyr::filter(!back_capture_new %in% c("Current", "Delayed", "unknown"))

new_master1 <- new_master

# ── Notifications summary table ────────────────────────────────────────────────
df1 %>%
  dplyr::select(nmccategories, time_to_notification) %>%
  mutate(time_to_notification = as.numeric(time_to_notification)) %>%
  tbl_summary(
    by    = nmccategories,
    label = list(time_to_notification ~ "Time to Notification")
  ) %>%
  modify_footnote(update = everything() ~ NA) ->
  tbl_notifications

cat1_notif_time <- tbl_notifications[["table_body"]][["stat_1"]][[1]] %>% as.character()
```

# Introduction {.unnumbered}

This document reports conditions notified to the Notifiable Medical Conditions Surveillance System (NMCSS) and flagged for reporting to WHO as part of the Integrated Disease Surveillance and Response (IDSR). Data was exported from the NMCSS on `r formatted_datetime`. It covers conditions reported `r reporting_period`

::: {.callout-warning appearance="minimal"}
**Please note:** numbers reported are subject to change as verification and confirmation procedures are carried out. Numbers presented here are the best representation at the time of release.
:::

For more notes on data interpretation please see the [NMCSS Data Interpretation Infographic](https://www.nicd.ac.za/wp-content/uploads/2022/09/NMCSS-Data-Interpretation-Infographic-final2.pdf).

**Category 1? Check Your Notification's Done!**

---

## Category 1 Conditions at a Glance {.unnumbered}

```{r cat1-glance}
#| label: tbl-cat1-glance
#| tbl-cap: !expr 'paste0("Number of suspected and confirmed category 1 notifications, Epiweek ", epiweek_num, ", ", epiweek_year)'

category_1 <- condition_df %>%
  dplyr::filter(nmccategories == 1) %>%
  pull(condition)

cat_1_glance <- df1 %>%
  dplyr::filter(nmccategories == 1) %>%
  mutate(
    condition       = factor(condition, levels = category_1),
    case_definition = factor(case_definition, levels = c("Suspected", "Confirmed"))
  ) %>%
  dplyr::select(condition, case_definition) %>%
  tbl_summary(
    by         = case_definition,
    percent    = "row",
    statistic  = list(all_categorical() ~ "{n}")
  ) %>%
  modify_header(
    label         = "**Condition**",
    all_stat_cols() ~ "**{level}**, \n N = {scales::number(n)}"
  ) %>%
  add_overall(last = FALSE, statistic = ~"{n}", col_label = "**Total** \n N = {scales::number(n)}") %>%
  trim_tbl() %>%
  modify_footnote(update = everything() ~ NA) %>%
  modify_footnote(
    all_stat_cols() ~ "Suspected and confirmed cases are independent — not totalled."
  )

ft <- cat_1_glance %>%
  as_flex_table() %>%
  flextable::set_caption(paste0(
    "Suspected and confirmed category 1 notifications, Epiweek ",
    epiweek_num, ", ", epiweek_year
  )) %>%
  fontsize(size = 9, part = "header") %>%
  flextable::set_table_properties(layout = "autofit", width = 0.99)

to_bold <- which(cat_1_glance$table_body$stat_2 != 0)

bold(ft, i = to_bold, j = 4) %>%
  bg(i = to_bold, bg = "grey90", part = "body") %>%
  bg(bg = "#A5CA64", part = "header") %>%
  flextable::color(color = "white", part = "header")
```

---

# Context {.unnumbered}

## Confirmed Notifications {.unnumbered}

### Epitable {.unnumbered}

```{r tbl-epitable-confirmed}
#| label: tbl-epitable-confirmed
#| tbl-cap: "Number of confirmed notifications on NMCSS per epiweek. Grey column = most recent complete week."

flextable::set_flextable_defaults(
  font.family = "Arial", font.size = 8, header.font.size = 8,
  border.color = "black", padding = 0.1
)

source(here::here("NMC_reports_scripts", "scripts_and_functions", "context_table.R"))

max_epitable_date <- new_master1$notification_date %>% max(na.rm = TRUE)

context_table(
  new_master1 %>%
    dplyr::filter(grepl("confirmed", ignore.case = TRUE, case_definition)),
  as_date(max_epitable_date)
) -> ft_confirmed

grey_column <- 1

ft_confirmed[[1]] %>%
  autofit() %>%
  fontsize(size = 7, part = "all") %>%
  bg(j = length(ft_confirmed[[1]]$col_keys) - grey_column, bg = "grey90", part = "body") %>%
  theme_zebra() %>%
  bg(bg = "#A5CA64", part = "header") %>%
  flextable::color(color = "white", part = "header")
```

### Trends Plot {.unnumbered}

```{r fig-epicurve-confirmed}
#| label: fig-epicurve-confirmed
#| fig-cap: !expr 'paste0("Trend of weekly confirmed notifications for selected conditions, South Africa; January 2022–", end_date_month_year)'

current_year <- as.integer(format(Sys.Date(), "%Y"))

df_epicurve_confirmed <- new_master1 %>%
  dplyr::filter(case_definition %in% "Confirmed") %>%
  dplyr::filter(nmccategories %in% c(1, 2)) %>%
  epicurve_df(., date_index = "notification_date", grouping_vars = "condition")

df_epicurve_confirmed1 <- df_epicurve_confirmed %>%
  dplyr::filter(year %in% 2022:current_year) %>%
  mutate(year = factor(year, levels = 2022:current_year))

plot_confirmed <- df_epicurve_confirmed1 %>%
  dplyr::filter(condition %in% c(
    "Cholera", "Listeriosis", "Diphtheria", "Malaria",
    "Measles", "Mpox", "Pertussis", "Rubella",
    "Hepatitis A", "Agricultural or stock remedy poisoning",
    "Food borne illness outbreak"
  )) %>%
  plot_epicurve_df(
    x_axis_option   = "epiweek",
    grouping_vars   = "condition",
    color_select    = "olivedrab",
    add_bar         = TRUE,
    add_rolling_avg = c(TRUE, 7),
    axis_text_size  = 7
  )

plot_confirmed$plot +
  theme(strip.text.y.right = element_text(hjust = 0))
```

---

## All Notifications {.unnumbered}

### Epitable {.unnumbered}

```{r tbl-epitable-all}
#| label: tbl-epitable-all
#| tbl-cap: "Number of all (suspected + confirmed) notifications on NMCSS per epiweek. Grey column = most recent complete week."

source(here::here("NMC_reports_scripts", "scripts_and_functions", "context_table.R"))

context_table(new_master1 %>% make_fever_rash(), as_date(max_epitable_date)) -> ft_all

ft_all[[1]] %>%
  autofit() %>%
  fontsize(size = 7, part = "all") %>%
  bg(j = length(ft_all[[1]]$col_keys) - grey_column, bg = "grey90", part = "body") %>%
  theme_zebra() %>%
  bg(bg = "#A5CA64", part = "header") %>%
  flextable::color(color = "white", part = "header")
```

### Trends Plot {.unnumbered}

```{r fig-epicurve-all}
#| label: fig-epicurve-all
#| fig-cap: !expr 'paste0("Trend of weekly all notifications for selected conditions, South Africa; January 2022–", end_date_month_year)'

df_epicurve_all <- new_master1 %>%
  make_fever_rash() %>%
  dplyr::filter(nmccategories %in% c(1, 2)) %>%
  epicurve_df(., date_index = "notification_date", grouping_vars = "condition")

df_epicurve_all1 <- df_epicurve_all %>%
  dplyr::filter(year %in% 2022:current_year) %>%
  mutate(year = factor(year, levels = 2022:current_year))

plot_all <- df_epicurve_all1 %>%
  dplyr::filter(condition %in% c(
    "Cholera", "Listeriosis", "Diphtheria", "Malaria",
    "Fever-Rash", "Mpox", "Pertussis",
    "Hepatitis A", "Hepatitis B", "Hepatitis C", "Hepatitis E",
    "Bilharzia (schistosomiasis)",
    "Agricultural or stock remedy poisoning",
    "Food borne illness outbreak"
  )) %>%
  plot_epicurve_df(
    x_axis_option  = "epiweek",
    grouping_vars  = "condition",
    color_select   = "olivedrab",
    axis_text_size = 7
  )

plot_all$plot +
  theme(strip.text.y.right = element_text(hjust = 0))
```

---

**END**
